---
title: Weighting Strategies in DKGE
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Weighting Strategies in DKGE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(dkge)
library(ggplot2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
options(
  tibble.print_min = 4,
  tibble.print_max = 4,
  digits = 3
)
set.seed(2024)
theme_set(theme_dkge())
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```


DKGE has separate weighting controls for locations within a subject, whole
subjects, and spatial transport. Each changes a different estimand. This
vignette shows where the controls enter the computation and what must be
reported when they are used.

Weighting is needed when “one unit, one vote” is scientifically wrong—for
example, when parcels have unequal area or some measurements are demonstrably
less reliable. It is not a generic way to make a result cleaner. Every weight
answers two questions: *what is the unit being reweighted, and at what stage?*

```text
subject beta block --spatial weights--> subject moment --subject weights--> group basis --transport masses--> common map
```

### Adaptive voxel weighting (preview)

DKGE can also rescale locations adaptively before fold bases are constructed.
`dkge_weights()` combines a prespecified prior, such as an external reliability
map, with training-fold statistics such as design-kernel energy or precision.
Training-fold scope closes one leakage path but does not establish calibration
or improved sensitivity. The [Adaptive Voxel Weighting vignette](dkge-adaptive-weighting.html)
shows the specification, structural checks, and post-hoc updates with
`dkge_update_weights()`.

## Which weighting control changes which object?

| Layer | Unit receiving weight | Argument | Enters before | Typical rationale |
|---|---|---|---|---|
| Spatial | parcel, cluster, or voxel within subject | `Omega_list` / `omega` | subject moment | area, precision, or a declared spatial metric |
| Subject | entire subject block | `w_method`, `w_tau`, or `weights` | pooled eigensolve | prevent one high-energy block from dominating, or apply external subject mass |
| Transport | source and target spatial mass | `sizes` | common-map summary | preserve parcel area or voxel density during alignment |

Effect-by-subject precision weights for unequal trial counts are a fourth,
separate layer covered in `vignette("dkge-partial-effect-spaces")`. Adaptive
training-fold weights are covered in `vignette("dkge-adaptive-weighting")`.

The example below changes one layer at a time so that a changed output can be
attributed to a named assumption.

## Example Dataset

To illustrate the behavior of different weighting strategies, we generate a synthetic dataset with three subjects that exhibit varying cluster counts, sizes, and reliability characteristics. This heterogeneity makes the effects of different weighting approaches readily apparent during inspection.

```{r example-data, message=FALSE, warning=FALSE}
library(dkge)
library(purrr)
set.seed(42)

S <- 3
q <- 4
P <- c(6, 5, 7)
T <- 40

betas <- map(P, ~ matrix(rnorm(q * .x), q, .x))
designs <- map(seq_len(S), function(s) {
  X <- matrix(rnorm(T * q), T, q)
  qr.Q(qr(X))
})

centroids <- map(P, ~ matrix(runif(.x * 3, min = -20, max = 20), .x, 3))
cluster_sizes <- map(P, ~ sample(10:40, .x, replace = TRUE))
reliability_scores <- list(
  rep(0.95, P[1]),
  rep(0.20, P[2]),  # subject 2 is the deliberately low-reliability block
  rep(0.80, P[3])
)
```

In this simulation, the `cluster_sizes` vectors serve as proxies for surface-area weights that would reflect the actual spatial extent of brain parcels, while the `reliability_scores` could represent various quality control measures such as motion artifacts or bootstrap stability assessments commonly used in neuroimaging analyses.

## Baseline Fit (No Spatial Weights)

```{r baseline-fit}
data_equal <- dkge_data(betas, designs = designs)
k_identity <- diag(data_equal$q)

fit_equal <- dkge(data_equal, K = k_identity, rank = 2, w_method = "none")
fit_equal$weights
```

In this baseline configuration where all weights are equal, every subject contributes uniformly to the construction of the shared basis. Consequently, cluster-level variability patterns are driven purely by the structure encoded in the design kernel, without any additional bias from differential weighting schemes.

## Size-Weighted Blocks

When you supply per-cluster mass values, this approach emphasizes the contribution of larger parcels while maintaining a diagonal information structure that preserves the independence assumption between clusters.

```{r size-weighting}
omega_size <- cluster_sizes

fit_size <- dkge(data_equal,
                 K = k_identity,
                 Omega_list = omega_size,
                 rank = 2,
                 w_method = "mfa_sigma1",
                 w_tau = 0.25)
fit_size$weights
```

Since the subject blocks now differ in their total weighted energy due to the size-based weighting, each block's relative influence on the underlying eigenproblem shifts accordingly. By inspecting the leading component, we can observe how the loadings are systematically rebalanced to favor larger parcels:

```{r size-loadings}
size_loadings <- lapply(fit_size$Btil, function(Bts) t(Bts) %*% fit_size$K %*% fit_size$U)
map(size_loadings, ~ round(.x[, 1, drop = FALSE], 3))
```

## Reliability Weighting

Reliability scores can be seamlessly incorporated into the `omega` weighting scheme by scaling each cluster according to its expected signal quality characteristics. When you want to combine both size and reliability considerations, this can be accomplished through a straightforward Hadamard (element-wise) product of the respective weight vectors.

```{r reliability-weighting}
omega_reliability <- map2(cluster_sizes, reliability_scores, `*`)

fit_reliability <- dkge(data_equal,
                        K = k_identity,
                        Omega_list = omega_reliability,
                        rank = 2,
                        w_method = "mfa_sigma1",
                        w_tau = 0.25)

weight_comparison <- rbind(
  size = fit_size$weights,
  size_x_reliability = fit_reliability$weights
)
round(weight_comparison, 3)

weighted_contribution_norm <- function(fit) {
  fit$weights * vapply(
    fit$contribs,
    function(x) sqrt(sum(x * x)),
    numeric(1)
  )
}
leverage_comparison <- rbind(
  size = weighted_contribution_norm(fit_size),
  size_x_reliability = weighted_contribution_norm(fit_reliability)
)
round(leverage_comparison, 1)
```

```{r reliability-leverage-plot, echo=FALSE, fig.height=3.8, fig.alt="Grouped bar chart showing how size-only and size-times-reliability weighting change each subject's contribution scale."}
leverage_long <- rbind(
  data.frame(subject = paste0("Subject ", seq_len(S)),
             scheme = "Size only", value = leverage_comparison["size", ]),
  data.frame(subject = paste0("Subject ", seq_len(S)),
             scheme = "Size × reliability",
             value = leverage_comparison["size_x_reliability", ])
)
ggplot(leverage_long, aes(subject, value, fill = scheme)) +
  geom_col(position = "dodge", width = 0.72) +
  labs(
    title = "Inspect contribution, not the scalar subject weight alone",
    x = NULL, y = "Contribution scale entering Chat", fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

The two visible rows are deliberately different. Subject 2 has the lowest
reliability scores. Its scalar MFA weight increases because inverse-leading-
singular-value scaling partly compensates for a lower-energy block; the scalar
weight alone is therefore not a measure of leverage. The second table reports
the direct contribution scale entering `Chat` (subject weight times the
Frobenius norm of that subject's unweighted contribution). On this controlled
example, adding reliability reduces subject 2's contribution relative to the
size-only fit.

```{r reliability-weighting-checks, include=FALSE}
stopifnot(!isTRUE(all.equal(fit_reliability$weights, fit_size$weights)))
stopifnot(fit_reliability$weights[2] > fit_size$weights[2])
stopifnot(leverage_comparison["size_x_reliability", 2] <
            leverage_comparison["size", 2])
```

In real data, a reliability vector might come from inverse first-level
variance, split-half or test-retest stability, a prespecified quality-control
measure, or resampling stability. The source matters: a weight estimated from
the same held-out outcome can leak information, while an external or
training-fold estimate has a clearer interpretation. When combining area and
reliability, multiply only when the resulting estimand is intended, then
rescale to a transparent reference such as mean one.

An important technical consideration is to always maintain positive weight values and add a small ridge term (such as `+ 1e-6`) when your weight construction procedure might potentially produce zero values.

## Spatial Covariance (Smoothness) Weighting

When spatial coordinate information for clusters is available, you can replace simple diagonal weights with more sophisticated smooth covariance matrices that capture spatial relationships. The example below demonstrates how to construct Gaussian affinity matrices that explicitly reward neighboring parcels for exhibiting coordinated activation patterns.

```{r smoothness-weighting}
gaussian_cov <- function(coords, bandwidth = 15) {
  D <- as.matrix(dist(coords))
  K <- exp(-(D / bandwidth)^2)
  (K + t(K)) / 2 + diag(1e-6, nrow(K))
}

omega_smooth <- lapply(centroids, gaussian_cov)

fit_smooth <- dkge(data_equal,
                   K = k_identity,
                   Omega_list = omega_smooth,
                   rank = 2,
                   w_method = "none")
fit_smooth$weights
```

These matrices define a spatial metric: DKGE right-multiplies each beta block by
the symmetric square root of `Omega`. That is weighting or smoothing, not
whitening (which would use an inverse square root). A Gaussian affinity couples
nearby parcels in the fitted energy calculation. Whether that coupling is
scientifically helpful is an assumption to check against an uncoupled fit, not
an automatic correction for sharp boundaries.

## Subject-Level Scaling

The `w_method` argument provides control over block-level reweighting that occurs after the spatial weights have already been applied to individual clusters. The default Multiple Factor Analysis scaling approach systematically shrinks the influence of dominant subjects by applying weights based on the inverse squared leading singular value of their contribution.

```{r subject-weighting}
fit_mfa <- dkge(data_equal,
                K = k_identity,
                Omega_list = omega_reliability,
                rank = 2,
                w_method = "mfa_sigma1",
                w_tau = 0.25)

fit_energy <- dkge(data_equal,
                   K = k_identity,
                   Omega_list = omega_reliability,
                   rank = 2,
                   w_method = "energy",
                   w_tau = 0)

rbind(mfa = fit_mfa$weights,
      energy = fit_energy$weights,
      size_mfa = fit_size$weights)
```

```{r subject-weighting-checks, include=FALSE}
stopifnot(!isTRUE(all.equal(fit_mfa$weights, fit_size$weights)))
stopifnot(!isTRUE(all.equal(fit_energy$weights, fit_size$weights)))
```

The `w_tau` parameter provides a mechanism to shrink extreme weight values toward more equal contributions across subjects, which becomes particularly valuable when working with imbalanced cohorts or datasets that contain potential outlier subjects.

## Weighting During Transport

The transport utilities within DKGE accept their own dedicated `sizes` argument that conceptually mirrors the role of the `omega` parameter but operates at a different stage—specifically *after* the components have been estimated. By providing consistent size information during transport, you can ensure that the reference medoid representation remains faithful to the actual surface area characteristics of the brain parcels.

```{r transport-weighting}
fit_mfa$centroids <- centroids  # attach for transport helpers; pass explicitly in production

comp_equal <- dkge_component_stats(
  fit_mfa,
  mapper = dkge_mapper_spec("sinkhorn", epsilon = 0.05, max_iter = 2000),
  centroids = centroids,
  inference = list(type = "parametric"),
  medoid = 1L
)

comp_weighted <- dkge_component_stats(
  fit_mfa,
  mapper = dkge_mapper_spec("sinkhorn", epsilon = 0.05, max_iter = 2000),
  centroids = centroids,
  sizes = cluster_sizes,
  inference = list(type = "parametric"),
  medoid = 1L
)

data.frame(equal = head(comp_equal$summary$stat),
           size_weighted = head(comp_weighted$summary$stat))
```

```{r transport-weighting-checks, include=FALSE}
n_medoid <- P[[1]]
stopifnot(nrow(comp_equal$summary) == n_medoid * fit_mfa$rank)
stopifnot(nrow(comp_weighted$summary) == n_medoid * fit_mfa$rank)
stopifnot(ncol(comp_equal$transport[[1]]) == n_medoid)
stopifnot(!isTRUE(all.equal(comp_equal$summary$stat, comp_weighted$summary$stat)))
```

Even when working with identical component scores, the application of reweighting during the transport phase systematically shifts the medoid summary statistics toward larger parcels, demonstrating the distinct influence of transport-stage weighting.

## Practical Tips

Several practical considerations can help you effectively implement weighting strategies in your DKGE analyses. First, construct your subjects using `dkge_subject()` to ensure that weight information stays properly attached to the data when passing through complex pipelines or streaming data loaders. Second, for voxelwise analyses, remember to feed cluster size information to `dkge_transport_to_voxels()` through the `sizes` parameter in order to preserve volumetric weighting throughout the mapping process.

When combining multiple weighting concepts, maintain positive vector values and consider adding a mild ridge term (such as `+ 1e-6`) to covariance matrices to prevent numerical instabilities. Additionally, make it a practice to revisit the `fit$weights`, singular values, and eigenspectra whenever you modify weighting assumptions, as these provide quick diagnostic information about the balance of contributions across subjects.

Collectively, these weighting tools allow you to tailor DKGE analyses to accommodate study-specific quality metrics or incorporate prior spatial beliefs, all without requiring modifications to the core fitting pipeline.
