---
title: "Plotting DKGE Fits"
author: "DKGE Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: false
    df_print: paged

vignette: >
  %\VignetteIndexEntry{Plotting DKGE Fits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(dkge)
library(ggplot2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
options(
  tibble.print_min = 4,
  tibble.print_max = 4,
  digits = 3
)
set.seed(2024)
theme_set(theme_dkge())
```


This vignette provides a comprehensive walkthrough of the new plotting helpers bundled with DKGE. Our approach involves simulating a small dataset with known factorial structure, fitting the DKGE model, and then showcasing what we call the "Five Fundamentals" — a core set of diagnostic visualizations that together provide essential insights into model quality and interpretability.

These five fundamental diagnostics include: first, a scree plot enhanced with one-SE overlay for rank selection guidance; second, an effect-space loadings heatmap that reveals how design factors contribute to each component; third, subject contribution plots that decompose both weights and energy across individuals; fourth, subspace stability analysis using principal angles to assess cross-validation robustness; and finally, anchor-level information maps that display both Haufe transformations and leave-one-covariate-out (LOCO) importance scores.

All plots are designed with a unified visual styling achieved through `theme_dkge()`, ensuring consistency across the entire analysis workflow. Moreover, these individual plots can be seamlessly composed into a single comprehensive dashboard using the `dkge_plot_suite()` function.

## Prerequisites

```{r libraries, message=FALSE}
library(dkge)
library(ggplot2)
library(patchwork)
```

The information-map plotting functions provide enhanced labeling capabilities by optionally utilizing the `ggrepel` package to annotate top anchors with non-overlapping text placement. When `ggrepel` is not available in your environment, the code gracefully falls back to standard base text labels to ensure functionality is maintained.

## Simulate a toy dataset

We begin by generating a small synthetic dataset manually. Each subject has a diagonal design matrix (one column per effect), and we add modest signal plus noise so the latent components are recoverable but not trivial. Using this construction keeps the vignette self-contained and avoids dependencies on specialised simulators while still providing deterministic ground truth for the plots.

```{r simulate}
q <- 4L   # number of effects
P <- 30L  # clusters (voxels) per subject
S <- 8L   # subjects

make_subject <- function(id) {
  design <- diag(q)
  colnames(design) <- paste0('eff', seq_len(q))
  signal <- matrix(rnorm(q * P, sd = 0.4), nrow = q)
  noise <- matrix(rnorm(q * P, sd = 0.2), nrow = q)
  dkge_subject(signal + noise, design = design, id = paste0('sub', id))
}

subjects <- lapply(seq_len(S), make_subject)
K <- diag(q)
fit <- dkge(subjects, kernel = K, rank = 4, w_method = 'none')
```

## Auxiliary objects

We now produce helper objects used in later sections. In a production pipeline you would typically obtain LOSO-aligned bases via `dkge_contrast(..., align = TRUE)` and real information maps from `dkge_info_map_haufe()` / `dkge_info_map_loco()`. Here we synthesise lightweight placeholders so the plots remain deterministic.

To illustrate subspace stability we create a few perturbed versions of the fitted basis and re-orthonormalise them in the \(K\)-metric.

```{r loso, results='hide'}
# In practice you can retrieve per-fold bases from
# dkge_contrast(..., align = TRUE)$metadata$bases.
# Here we synthesise a few perturbations for illustration.
fit_U <- fit[['U']]
bases <- replicate(4, {
  perturb <- matrix(rnorm(length(fit_U), sd = 0.02), nrow = nrow(fit_U))
  dkge_k_orthonormalize(fit_U + perturb, fit[['K']])
}, simplify = FALSE)
base_labels <- paste0('base', seq_along(bases))
```

The information-map visualization requires anchor-level results from either Haufe transformation analysis or leave-one-covariate-out (LOCO) importance calculations. Since we are working with a synthetic example for demonstration purposes, we create mock anchor vectors that simulate these outputs. In a real-world analysis, you would replace these placeholder vectors with actual Haufe or LOCO results computed from your fitted model.

```{r mock-info}
fake_haufe <- list(mean_anchor = rnorm(30))
fake_loco  <- list(loco_anchor = rexp(30))
```

## Individual plots

### Scree

```{r scree, fig.width=5, fig.height=4}
# For illustrative purposes we annotate the scree at component 3.
# In a real analysis you would obtain this optimal rank from dkge_cv_rank_loso()
# or dkge_cv_kernel_rank() cross-validation procedures.
one_se_pick <- 3
dkge_plot_scree(fit, one_se_pick = one_se_pick)
```

### Effect-space loadings

```{r loadings, fig.width=5, fig.height=4}
dkge_plot_effect_loadings(fit, comps = 1:3, zscore = TRUE)
```

### Subject contributions

`dkge_plot_subject_contrib()` returns two linked panels. The left panel simply
visualises the subject-level weights that were used while fitting the model
(`fit$weights`). Those weights depend on the `w_method` argument passed to
`dkge()`: in this vignette we deliberately chose `w_method = "none"`, so each
subject keeps unit weight and the bars are all equal to one. Switching to
`"mfa_sigma1"` or `"energy"` would display the corresponding MFA-style or
energy-based weighting actually used in the eigensolve. The heatmap on the
right shows how much norm ("energy") each subject contributes to the selected
components after those weights have been applied.

```{r contrib, fig.width=9, fig.height=3}
contrib <- dkge_plot_subject_contrib(fit, comps = 1:3)
contrib$weights + contrib$energy + patchwork::plot_layout(widths = c(1, 2))
```

### Subspace stability

This diagnostic compares each cross-validated basis (or synthetic perturbation
here) against a consensus basis by computing principal angles in the \(K\)
metric. Smaller angles mean that a base reproduces the consensus component more
closely, so parallel lines near zero indicate a stable subspace, whereas large
excursions highlight folds or components that deviate materially.

```{r stability, fig.width=5, fig.height=4}
dkge_plot_subspace_stability(bases, K = fit[['K']], labels = base_labels)
```

### Information maps

The information panels translate component patterns back to anchor space. The
Haufe transformation re-expresses each component in terms of observable anchor
weights, making it easier to interpret which clusters drive a latent effect once
you account for the whitening in the eigensolve. The LOCO panel measures how
the objective would drop if each anchor were removed in turn, so tall positive
bars flag anchors that materially support the fit. Together they provide a
bridge between latent structure and spatial attribution.

```{r info, fig.width=5, fig.height=4}
panels <- dkge_plot_info_anchor(info_haufe = fake_haufe,
                                info_loco = fake_loco,
                                top = 5)
panels$haufe + panels$loco
```

## The "Five Fundamentals" dashboard

```{r suite, fig.width=8, fig.height=8}
dkge_plot_suite(fit,
                one_se_pick = one_se_pick,
                comps = 1:3,
                bases = bases,
                consensus = fit[['U']],
                base_labels = base_labels,
                info_haufe = fake_haufe,
                info_loco = fake_loco,
                top = 5)
```

The `dkge_plot_suite()` function intelligently arranges all five fundamental diagnostic panels into a coherent, publication-ready dashboard with consistent spacing and visual hierarchy. When you provide real Haufe transformation and LOCO importance results, the function seamlessly incorporates these into the final row of visualizations. Conversely, if these anchor-level analyses are omitted from your call, the layout automatically adapts by placing an informative placeholder message in their position, ensuring the dashboard remains structurally sound.

## Saving the dashboard

```{r save-suite, eval=FALSE}
dkge_plot_suite(fit,
                bases = bases,
                consensus = fit[['U']],
                base_labels = base_labels,
                info_haufe = fake_haufe,
                info_loco = fake_loco,
                save_path = "dkge_dashboard.png",
                width = 10,
                height = 10)
```

## Summary

The DKGE plotting helpers are intentionally designed with computational efficiency and visual clarity as primary goals. They operate exclusively in the low-dimensional design space rather than in high-dimensional voxel space, ensuring fast rendering even for complex models. All functions inherit a cohesive visual theme that maintains consistency across different plot types, while simultaneously supporting both rapid exploratory inspection and publication-quality dashboard generation.

These plotting tools offer considerable flexibility to accommodate different analytical workflows. For interactive exploration and hypothesis generation, you can embed individual diagnostic panels directly inline within your analysis notebooks. When conducting formal model quality assurance, you can capture high-resolution snapshots using standard R graphics functions like `png()` or `ggsave()` for documentation and reporting purposes. For more comprehensive presentations, you can extend the basic five-panel suite by incorporating additional visualization rows — such as brain rendering panels showing voxel-level activations — through the powerful layout capabilities provided by the patchwork package ecosystem.

Happy plotting!
