---
title: Partial Effect Spaces
description: >
  Declare a partial global effect grid, align subject-local rows to it, and
  choose a coverage policy before any fit.
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Partial Effect Spaces}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(dkge)
set.seed(2026)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

DKGE can represent one global effect space when subjects contribute different
cells or estimate the same cells with very different trial counts. The output
is still a common q-dimensional group basis, but three distinct problems must
be handled separately:

This page is for the case where the usual rectangular data picture is false.
Perhaps controls and patients occupy different rows of a global grid, or every
subject attempted the same factorial design but some cells have many more
trials than others. If every subject has every effect with comparable
precision, you do not need these controls; use `vignette("dkge-workflow")`.

The guiding rule is simple: **absence, imprecision, and estimation noise are
not synonyms**. The rest of the vignette shows where each enters the fit.

| Problem | What changes | DKGE mechanism |
|---|---|---|
| A cell is absent for a subject | Whether an effect pair is observed | `observed_rows` and `missingness` |
| A cell has 2 trials for one subject and 20 for another | Precision of an observed effect | `dkge_effect_weights()` |
| Cell estimates contain finite-trial estimation noise | Expected diagonal noise in the second moment | `debias` |

Coverage, precision weighting, and debiasing are complementary. A count weight
does not remove estimation-noise bias; analytic subtraction does not decide
which subject should contribute more; and zero-filling alone does not record
that a cell was unobserved.

The key contract is:

- `dkge_data()` aligns subject-local effect rows to the union of all effect
  labels.
- Missing rows are rendered as zero rows in the aligned matrices.
- The original observation pattern is retained on the returned bundle as
  `observed_rows`, plus `obs_mask` and `pair_counts` under `provenance`.
- `dkge_fit(missingness = ...)` controls how the partial coverage is handled
  when the q-space covariance is accumulated.

```text
local observed rows -> global labelled grid -> coverage-aware raw moment -> precision/debiasing -> kernel transform
```

The first example isolates missing cells. The larger trialwise example then
adds unequal precision and debiasing. Keeping those examples separate prevents
one mechanism from appearing to solve all three problems.

## How do you declare a partial global grid?

Use `dkge_effect_grid()` to declare the global effect cells and which factors
are within- versus between-subject.

```{r grid}
grid <- dkge_effect_grid(
  factors = list(
    group = c("control", "patient"),
    task = c("A", "B"),
    measure = c("low", "high")
  ),
  scope = c(group = "between", task = "within", measure = "within"),
  block_factors = "group"
)

grid$cell_labels
```

`block_factors = "group"` requests independent group blocks in the design
kernel. Terms such as `task` and `measure` are replicated within each group
block rather than coupled across groups by an all-ones group factor.

```{r kernel}
kernel <- design_kernel(
  grid,
  terms = list(
    "group", "task", "measure",
    c("group", "task"),
    c("group", "measure"),
    c("task", "measure"),
    c("group", "task", "measure")
  ),
  basis = "cell",
  normalize = "none"
)

kernel$info$term_scope
kernel$info$block_factors
```

The kernel metadata classifies terms as `within`, `between`, or `mixed`.
Contrast helpers use this metadata to recommend the appropriate inference
route.

## How are subject-local rows aligned?

Each subject supplies only the four rows for their own group. `dkge_data()`
embeds those rows into the eight-row union and records which global rows were
actually observed.

```{r data}
cell4 <- c("A:low", "A:high", "B:low", "B:high")
subject_info <- data.frame(
  subject_id = paste0("s", 1:4),
  group = c("control", "control", "patient", "patient")
)

make_subject_beta <- function(group) {
  rows <- paste(group, sub(":.*$", "", cell4), sub("^.*:", "", cell4), sep = ":")
  B <- matrix(rnorm(4 * 12), nrow = 4,
              dimnames = list(rows, paste0("feature", 1:12)))
  X <- diag(4)
  colnames(X) <- rows
  list(B = B, X = X)
}

subjects <- lapply(subject_info$group, make_subject_beta)
betas <- lapply(subjects, `[[`, "B")
designs <- lapply(subjects, `[[`, "X")

bundle_partial <- dkge_data(betas, designs, subject_ids = subject_info$subject_id)
bundle_partial$effects
bundle_partial$observed_rows
bundle_partial$provenance$pair_counts
```

Controls observe rows 1-4 and patients observe rows 5-8. Cross-group row pairs
have zero pair counts.

```{r coverage-map, echo=FALSE, fig.width=7.2, fig.height=3.2, fig.alt="Binary heatmap showing that control subjects observe the first four global effect cells and patient subjects observe the last four."}
coverage_matrix <- do.call(rbind, bundle_partial$provenance$obs_mask)
colnames(coverage_matrix) <- sub(":", "\n", bundle_partial$effects)
image(
  seq_len(ncol(coverage_matrix)), seq_len(nrow(coverage_matrix)),
  t(coverage_matrix[nrow(coverage_matrix):1, , drop = FALSE]),
  col = c("#edf0f2", "#b33b2e"), zlim = c(0, 1), axes = FALSE,
  xlab = "Global effect cell", ylab = "Subject",
  main = "Observed rows before zero-filled alignment"
)
axis(1, at = seq_len(ncol(coverage_matrix)), labels = colnames(coverage_matrix),
     las = 2, cex.axis = 0.75)
axis(2, at = seq_len(nrow(coverage_matrix)), labels = rev(rownames(coverage_matrix)),
     las = 1)
box()
legend("top", inset = c(0, -0.18), xpd = TRUE, horiz = TRUE, bty = "n",
       legend = c("absent", "observed"), fill = c("#edf0f2", "#b33b2e"))
```

The empty half of each row is structural absence, not a measured beta of zero.
The observation mask preserves that distinction after the aligned matrices are
expanded to eight rows.

## How should partial coverage enter the fit?

The default `missingness = "none"` preserves the historical zero-filled
accumulation. That compatibility setting is appropriate when every subject
observes every effect. For a genuinely partial global space, choose a policy
that uses the recorded coverage:

- `"mask"` zeros entries whose applicable coverage measure is below
  `min_pairs`: observed pair mass without effect weights, and Kish pair ESS
  when effect-precision weights are active.
- `"rescale"` returns a per-pair mean: it divides by observed pair mass
  without effect weights and by total pair precision when those weights are
  active.
- `"shrink"` blends that per-pair mean toward its diagonal according to pair
  coverage (pair ESS on the precision-weighted branch).

```{r fit}
fit_partial <- dkge_fit(
  bundle_partial,
  K = kernel,
  rank = 2,
  w_method = "none",
  missingness = "mask",
  miss_args = list(min_pairs = 1)
)

fit_partial$missingness
fit_partial$pair_counts[1:4, 5:8]
```

Here too, `w_method = "none"` isolates the coverage policy. With the default
MFA subject scaling, zero filling still occurs first, but the later transformed
block energies also affect each subject's scalar contribution.

```{r partial-coverage-checks, include = FALSE}
stopifnot(identical(dim(fit_partial$pair_counts), c(8L, 8L)))
stopifnot(all(fit_partial$pair_counts[1:4, 5:8] == 0))
stopifnot(all(fit_partial$pair_counts[1:4, 1:4] == 2))
```

### What the zero-filled rows do, and do not, do

Coverage policies act in *raw* effect space, before the pooled ruler `R` and
the design kernel mix rows. A subject's raw second moment
`B_s B_s'` therefore has exact zero rows and columns wherever that subject
observed nothing: placeholders contribute no energy of their own. Zero filling
also precedes subject-weight derivation, but a non-`"none"` `w_method` scores
the transformed block after the pooled ruler `R` and `Khalf` have mixed effect
coordinates. Subject weights therefore do not evaluate each raw zero row in
isolation.

They are not, however, insulated from the metric. The fit embeds each moment as
`K^{1/2} R' (B_s B_s') R K^{1/2}`, and if `K` couples an observed cell to an
unobserved one, `K^{1/2}` will place some of the observed energy on the
unobserved coordinate. That is the K-metric doing its job — it is the same
smoothing that makes an ordinal or circular kernel useful — not a leak of
values the subject never supplied.

If you want strict separation, say so in the kernel rather than in the
accumulator. `block_factors = "group"`, as used above, makes `K` block-diagonal
across groups, so `K^{1/2}` is block-diagonal too and no control-group energy
can reach a patient-group coordinate:

```{r block-diagonal}
Khalf <- kernel_roots(kernel$K)$Khalf
# Cross-group blocks of K^{1/2} are numerically zero.
max(abs(Khalf[1:4, 5:8]))
```

```{r block-diagonal-check, include = FALSE}
stopifnot(max(abs(Khalf[1:4, 5:8])) < 1e-12)
```

## Which contrasts are estimable within subject?

Contrasts named after kernel terms are tagged with the term scope.

```{r contrasts}
task4 <- c(-0.5, -0.5, 0.5, 0.5)
contrasts <- list(
  task = rep(task4, 2),
  group = c(rep(-0.25, 4), rep(0.25, 4)),
  "group:task" = c(-task4, task4)
)

task_res <- dkge_contrast(fit_partial, contrasts["task"], method = "loso", align = FALSE)
knitr::kable(task_res$metadata$contrast_estimability)
```

```{r task-estimability-checks, include = FALSE}
stopifnot(identical(
  unname(task_res$metadata$contrast_estimability$estimability),
  "within"
))
```

`task` is a within-subject contrast, so LOSO/k-fold contrast inference is the
right family. `group` and `group:task` are between or mixed effects. DKGE will
still return descriptive cross-fitted maps, but the confirmatory route is the
between-subject RRR layer: construct a subject model with
`dkge_subject_model()`, then call `dkge_between_rrr()` and assess it with
`dkge_between_permute()`.

```{r group-contrast, warning = TRUE}
group_res <- dkge_contrast(
  fit_partial, contrasts[c("group", "group:task")],
  method = "loso", align = FALSE
)
knitr::kable(group_res$metadata$contrast_estimability)
```

```{r group-estimability-checks, include = FALSE}
stopifnot(identical(
  unname(group_res$metadata$contrast_estimability$estimability),
  c("between", "mixed")
))
stopifnot(all(grepl(
  "dkge_between",
  group_res$metadata$contrast_estimability$recommended_inference,
  fixed = TRUE
)))
```

## What is deliberately outside this workflow?

A supervised tensor or CP/Tucker model could be useful for a research extension,
especially when the goal is to discover a latent group-conditioned expression
pattern. That is not part of the DKGE core contract here. The core feature is
partial global effect spaces with explicit coverage, q-space missingness
policies, and term-scope-aware inference recommendations.

For group, trait, or mixed-effect tests on the resulting subject
representations, continue with `vignette("dkge-between-subjects")`. For a
conceptual map of component-, contrast-, and feature-level claims, see
`vignette("dkge-concepts")`.

## Where to go next

- `vignette("dkge-unbalanced-trialwise")` fits a 3 x 5 x 4 trialwise design on
  the contract this page defines.
- `vignette("dkge-between-subjects")` is the confirmatory route for the between
  and mixed contrasts named above.
- `vignette("dkge-concepts")` maps component-, contrast-, and feature-level
  claims onto what each analysis supports.
