---
title: Contrasts, Inference, and Bootstraps
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Contrasts, Inference, and Bootstraps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(dkge)
library(ggplot2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
options(
  tibble.print_min = 4,
  tibble.print_max = 4,
  digits = 3
)
set.seed(2024)
theme_set(theme_dkge())
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```


This vignette follows one estimand from construction to interpretation: the
subject-level effect-1 minus effect-2 contrast on a shared simulated spatial
support. It then distinguishes that contrast estimand from component-level
inference and shows how a subject bootstrap quantifies sampling variation
conditional on the fitted basis and transport operators.

Why start from the estimand? Because “run inference on the DKGE fit” is not a
complete instruction. A component statistic, a named effect contrast, and a
featurewise transported map have different sampling units and answer different
questions. Here the object journey is explicit:

```text
named effect contrast -> held-out subject fields -> one shared spatial support -> subject resampling
```

The component-level procedure appears afterward as a deliberate contrast: it
is a different branch, not another test of the same quantity.

## What signal should the contrast recover?

The first two effect rows contain opposite halves of a smooth spatial signal, so
their raw contrast equals `planted`. The remaining effects contain noise. Every
subject uses the same cluster ordering with small coordinate jitter, giving the
transport step a known target while retaining subject variation.

```{r data}
library(dkge)
S <- 8; q <- 4; P <- 25; T <- 60
effect_names <- paste0("effect", seq_len(q))
cluster_axis <- seq(-1, 1, length.out = P)
planted <- exp(-5 * (cluster_axis + 0.35)^2) -
  0.65 * exp(-7 * (cluster_axis - 0.45)^2)
truth <- matrix(0, q, P, dimnames = list(effect_names, NULL))
truth[1, ] <- planted / 2
truth[2, ] <- -planted / 2

betas <- lapply(seq_len(S), function(s) {
  truth + matrix(rnorm(q * P, sd = 0.30), q, P)
})
designs <- lapply(seq_len(S), function(s) {
  X <- matrix(rnorm(T * q), T, q)
  X <- qr.Q(qr(X))
  colnames(X) <- effect_names
  X
})
reference_centroids <- cbind(x = 10 * cluster_axis, y = 0, z = 0)
centroids <- lapply(seq_len(S), function(s) {
  reference_centroids + matrix(rnorm(P * 3, sd = 0.02), P, 3)
})
subjects <- lapply(seq_len(S), function(s) dkge_subject(betas[[s]], designs[[s]], id = paste0("sub", s)))
bundle <- dkge_data(subjects)
fit <- dkge(bundle, K = diag(q), rank = 3)
fit$centroids <- centroids  # attach for transport helpers; pass explicitly in production
```

## Leave-One-Subject-Out (LOSO) Contrasts

The LOSO approach refits the latent basis without each held-out subject before
computing that subject's contrast field. This limits basis-reuse optimism; it
does not by itself guarantee unbiased population inference. The public
`dkge_contrast(..., method = "loso")` path returns those subject-specific
fields, after which one fixed set of transport operators can place them on a
common reference for the inferential procedure defined below.

```{r loso}
contrast_vec <- c(1, -1, 0, 0)  # effect1 minus effect2
loso <- dkge_contrast(fit, contrast_vec, method = "loso")

transported <- dkge_transport_contrasts_to_medoid(
  fit,
  loso,
  medoid = 1L,
  centroids = centroids,
  mapper = dkge_mapper_spec("sinkhorn", epsilon = 0.05, lambda_spa = 1,
                            max_iter = 5000, tol = 1e-4)
)

str(loso$values, max.level = 1)
```

Keep the three representations distinct. `loso$values` contains raw
subject-level fields on the original subject supports.
`transported[[1]]$subj_values` is the subject-by-reference-cluster matrix after
alignment. `transported[[1]]$value` is only the coordinate-wise median of that
matrix. Inference needs the subject rows, not just the median curve.

We summarize the region where the planted raw contrast is largest. DKGE
projection and transport change the numerical scale, so agreement in direction
is meaningful here; equality to the raw planted amplitude is not expected.

```{r contrast-estimand}
active_region <- which(planted >= stats::quantile(planted, 0.80))
active_subject_values <- rowMeans(
  transported[[1]]$subj_values[, active_region, drop = FALSE]
)
contrast_summary <- data.frame(
  target = "mean transported contrast in planted-positive region",
  estimate = mean(active_subject_values),
  subject_sd = sd(active_subject_values),
  positive_subjects = sum(active_subject_values > 0),
  n_subjects = length(active_subject_values)
)
contrast_summary
```

In this seeded example, the active-region estimate is
`r round(contrast_summary$estimate, 3)`, and
`r contrast_summary$positive_subjects` of `r contrast_summary$n_subjects`
subjects have the planted positive direction. This is recovery evidence for the
simulation, not a population effect size or external validation result.

```{r transport-checks, include=FALSE}
transport_ok <- vapply(transported[[1]]$diagnostics, function(x) {
  isTRUE(x$converged)
}, logical(1))
transport_median <- apply(transported[[1]]$subj_values, 2, stats::median)
stopifnot(all(transport_ok),
          max(abs(transported[[1]]$value - transport_median)) < 1e-12)
```

Plot the subject fields as well as their summary. Agreement in shape is more
informative than a polished median with its heterogeneity hidden.

```{r loso-plot, fig.width=6.5, fig.height=4, fig.alt="Transported LOSO contrast profiles for eight subjects with their coordinate-wise median highlighted."}
loso_medoid <- transported[[1]]$value
subject_fields <- transported[[1]]$subj_values
matplot(
  t(subject_fields), type = "l", lty = 1, lwd = 0.8,
  col = grDevices::adjustcolor("grey45", alpha.f = 0.35),
  xlab = "Reference cluster", ylab = "Transported contrast",
  main = "Held-out subject fields on one support"
)
abline(h = 0, col = "grey70", lty = 2)
lines(loso_medoid, col = "#b33b2e", lwd = 2.4)
legend("topright", legend = c("subjects", "median"),
       col = c("grey55", "#b33b2e"), lwd = c(1, 2.4), bty = "n")
```

The median follows the planted positive and negative regions, while the grey
lines show the uncertainty and subject heterogeneity that the next stage must
respect. This figure is a recovery diagnostic for the simulation, not a
multiple-comparison-corrected spatial result.

## Analytic Inference on Components

`dkge_component_stats()` answers a different question from the contrast above:
it transports subject component loadings and tests their coordinate-wise group
means. The parametric option below uses a t reference distribution, then applies
the requested p-value adjustment across returned component-coordinate tests.

```{r analytic}
comp_stats <- dkge_component_stats(
  fit,
  components = 1:2,
  mapper = dkge_mapper_spec("sinkhorn", epsilon = 0.05, lambda_spa = 0.5,
                            max_iter = 2000, tol = 1e-4),
  centroids = centroids,
  inference = list(type = "parametric"),
  medoid = 1L
)
head(comp_stats$summary)
```

The table reports component, reference-cluster index, t statistic, raw p-value,
FDR-adjusted p-value, and the declared significance flag. These are
component-loading tests under the mapper and parametric assumptions above; they
are not tests of the effect-1 minus effect-2 contrast.

## Bootstrap Inference

The following nonparametric subject bootstrap resamples the already transported
contrast vectors and recomputes their mean. It estimates sampling variation
conditional on the fitted basis, chosen medoid, and fixed transport operators;
it does not propagate uncertainty from those earlier steps.

```{r bootstrap, message=FALSE}
subject_medoid <- lapply(seq_len(nrow(transported[[1]]$subj_values)), function(idx) {
  transported[[1]]$subj_values[idx, ]
})

boot <- dkge_bootstrap_projected(
  subject_medoid,
  B = 200,
  seed = 2
)

head(boot$medoid$mean)
```

For the declared active-region summary, the bootstrap distribution is:

```{r bootstrap-estimand}
active_boot <- rowMeans(boot$medoid$boot[, active_region, drop = FALSE])
bootstrap_summary <- data.frame(
  bootstrap_mean = mean(active_boot),
  lower_95 = unname(stats::quantile(active_boot, 0.025)),
  upper_95 = unname(stats::quantile(active_boot, 0.975))
)
bootstrap_summary
```

```{r bootstrap-distribution, echo=FALSE, fig.width=5.5, fig.height=3.8, fig.alt="Histogram of bootstrap means for the transported contrast in the planted positive region, with the percentile interval marked."}
hist(
  active_boot, breaks = 18, col = "#d9e1e8", border = "white",
  main = "Sampling variation of the declared summary",
  xlab = "Mean contrast in planted-positive region"
)
abline(v = bootstrap_summary$bootstrap_mean, col = "#b33b2e", lwd = 2)
abline(v = c(bootstrap_summary$lower_95, bootstrap_summary$upper_95),
       col = "#315f8c", lwd = 2, lty = 2)
legend(
  "topright", legend = c("bootstrap mean", "95% percentile limits"),
  col = c("#b33b2e", "#315f8c"), lwd = 2, lty = c(1, 2), bty = "n"
)
```

The resulting conditional 95% percentile interval is
[`r round(bootstrap_summary$lower_95, 3)`,
`r round(bootstrap_summary$upper_95, 3)`]. Because the earlier basis and
transport operators are held fixed, this interval should not be described as
unconditional uncertainty for the full DKGE pipeline.

`dkge_bootstrap_projected()` also returns coordinate-wise means, standard
deviations, z ratios, intervals, and (optionally) the draws used above. When the
estimand lives in q-space instead, `dkge_bootstrap_qspace()` performs multiplier
resampling there and can reuse a declared transport cache.

## Practical Recommendations

Choose the inferential target before choosing a procedure. The component table
tests transported component loadings; the projected bootstrap above summarizes
an explicit contrast after transport. Parametric inference relies on its
reference-distribution assumptions. Subject resampling avoids that particular
reference distribution but still requires representative, independent subjects
and enough observations for the empirical distribution to be informative.

Before reporting results, state the contrast, analysis space, aggregation rule,
resampling unit, mapper, medoid, and multiplicity correction. Inspect
subject-level values as a diagnostic, but do not remove unusual subjects without
a prespecified rule and sensitivity analysis.
