---
title: Classification with DKGE
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Classification with DKGE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

```{r library, message=FALSE}
library(dkge)
set.seed(12)
```

The DKGE classification helpers decode experimental conditions from
subject-level effect patterns. This vignette follows one complete path: define
the effect space, fit DKGE, specify the classes, and estimate held-out-subject
performance.

There are two distinct cross-validation contracts. The default `mode = "auto"`
uses subject-level folds but selects the faster `"cell"` path for
within-subject targets; its classifier is cross-validated, while the global
DKGE basis has seen every subject. Use `mode = "cell_cross"` when the basis
itself must also be re-estimated without each held-out subject. The primary
example below uses that stricter end-to-end path.

## Setup

We define two experimental factors — condition (A vs B) and time (4 time points) — and
construct a design kernel that encodes their structure:

```{r design-kernel}
factors <- list(cond = list(L = 2), time = list(L = 4))
kern    <- design_kernel(factors, basis = "effect")
q       <- nrow(kern$K)   # number of effect-coded columns
q
```

We simulate 10 subjects, each with `q` design effects and 60 voxels. A detectable signal
distinguishing the two conditions is embedded in the condition contrast:

```{r simulate}
n_subjects <- 10L
v <- 60L

make_subject <- function(id) {
  # design: identity matrix over effects (one trial per effect)
  design <- diag(q)
  colnames(design) <- rownames(kern$K)

  # signal: condition contrast (first effect column) drives voxels 1-10
  signal <- matrix(0, nrow = q, ncol = v)
  signal[1, 1:10] <- 0.5    # positive signal for condition contrast

  beta <- signal + matrix(rnorm(q * v, sd = 1.0), nrow = q)
  dkge_subject(beta, design = design, id = paste0("sub", id))
}

subjects <- lapply(seq_len(n_subjects), make_subject)
```

Fit the DKGE model using the design kernel. `design_kernel()` returns a list with `$K`
(the q×q kernel matrix) and `$info` (factor metadata used by `dkge_targets()`):

```{r fit}
fit <- dkge(subjects, K = kern, rank = 2)
fit
```

## Classification targets

`dkge_targets()` maps design formula terms to classification contrasts using the factor
metadata stored in the kernel. Each term becomes a binary (or multi-class) decoding
problem over the group embedding.

```{r targets}
targets <- dkge_targets(fit, ~ cond)
length(targets)                      # one target per formula term
targets[[1]]$name                    # "cond"
nrow(targets[[1]]$weight_matrix)     # 2 classes (A vs B)
```

The weight matrix transforms subject-level beta vectors into class-specific pattern
vectors for the downstream classifier.

## Cross-validated decoding

```{r classify}
cls <- dkge_classify(
  fit,
  targets = targets,
  method  = "lda",     # "lda" (default) or "logit"
  mode    = "cell_cross",
  n_perm  = 0,         # descriptive cross-validated decoding
  seed    = 99
)
print(cls)
```

The result is a `dkge_classification` object. Each entry of `$results` corresponds to
one target:

```{r results}
res <- cls$results[["cond"]]
res$mode       # "cell_cross": the basis is refitted inside each fold
res$metrics    # cross-validated accuracy and log-loss
```

The estimand is subject-level LOSO accuracy for classifying the condition
patterns, with both the DKGE basis and classifier refit without the held-out
subject. In this seeded simulation the planted condition signal yields:

```{r interpret-result}
condition_accuracy <- unname(res$metrics[["accuracy"]])
condition_log_loss <- unname(res$metrics[["logloss"]])
c(accuracy = condition_accuracy, logloss = condition_log_loss)
```

An accuracy of 1 means every held-out condition row was classified correctly in
this deliberately strong toy example. It is not an estimate of performance on
an external dataset.

Convert to a tidy data frame for plotting or downstream analysis:

```{r tidy}
df <- as.data.frame(cls)
head(df)
```

### Permutation p-values

The descriptive fit above deliberately sets `n_perm = 0`, so its p-values are `NULL`:

```{r pvalues}
res$p_values    # named numeric vector per metric (NULL when n_perm = 0)
```

For an inferential fit, preselect one scalar penalty independently and provide a
callback that reruns the complete data-dependent representation and classifier for
every randomized label vector. The callback must return finite named metrics:

```{r permutation-template, eval=FALSE}
full_recompute <- function(labels, row_data, target, fit, fold_assignments,
                           method, mode, lambda, metric, class_weights,
                           standardize_within_fold) {
  # Rebuild the DKGE fit, rank choice, folds, target representation, and
  # classifier from `labels`, then return all metrics requested in `metric`.
  list(metrics = c(accuracy = recomputed_accuracy))
}

cls_perm <- dkge_classify(
  fit,
  targets = targets,
  method = "lda",
  mode = "cell_cross",
  lambda = 1e-3,       # selected independently of these outcomes
  metric = "accuracy",
  n_perm = 199,
  seed = 99,
  control = list(randomization_recompute = full_recompute)
)
```

With 199 permutations and the plus-one correction, the smallest attainable
p-value is \(1/(199+1)=0.005\). A value at that boundary means none of the
sampled permutations was as extreme; it does not provide finer resolution.

### Subject-level predictions

Individual predictions and fold assignments live in `$row_data`:

```{r predictions}
head(res$row_data[, c("subject_label", "class_label", "fold")])
```

Full predicted probabilities are in `res$probabilities` (rows = subject × class
combinations, columns = class labels).

### Per-fold diagnostics

Confusion matrices and class counts per fold are in `res$diagnostics$folds`:

```{r diag}
diag_fold1 <- res$diagnostics$folds[[1]]
diag_fold1$confusion           # confusion matrix for fold 1
diag_fold1$class_counts_test   # observed class counts in test set
```

## Two modes: `cell` vs `cell_cross`

The `mode` argument controls how the group basis is applied at test time:

| Mode | Basis used | Supported claim |
|------|-----------|-----------------|
| `"cell"` | Global `fit$U` | Transductive performance within this cohort; the basis saw held-out subjects |
| `"cell_cross"` | Fold-specific LOSO `U_fold` | Prospective held-out-subject performance |
| `"delta"` | Fold-specific basis + subject labels | Prospective held-out-subject binary test; requires `y` |

`"auto"` (default) selects `"cell"` for within-subject targets and `"delta"` for
between-subject targets.

```{r cell-cross, eval=FALSE}
# Cross-fitted representation for a prospective generalisation claim
cls_strict <- dkge_classify(fit, targets = targets, mode = "cell_cross", n_perm = 0)
```

## Multiple betas per condition

When you have multiple beta estimates per condition (e.g., separate scanner runs), stack
the run-specific effects as extra rows in the design matrix and supply a weight matrix
that combines them into a single class pattern.

```{r multi-beta}
# 4 effects: A_run1, A_run2, B_run1, B_run2
make_subject_multi <- function(id) {
  design <- diag(4)
  colnames(design) <- c("A_run1", "A_run2", "B_run1", "B_run2")

  signal <- matrix(0, nrow = 4, ncol = v)
  signal[1:2, 1:10] <-  0.5   # both A runs share signal
  signal[3:4, 1:10] <- -0.5   # both B runs share the opposite signal

  beta <- signal + matrix(rnorm(4 * v, sd = 1.0), nrow = 4)
  dkge_subject(beta, design = design, id = paste0("sub", id))
}

subjects_multi <- lapply(seq_len(n_subjects), make_subject_multi)
fit_multi      <- dkge(subjects_multi, K = diag(4), rank = 2)

# Weight matrix: rows = classes, columns = effects.
# Each class averages its two runs.
W_runs <- matrix(0, nrow = 2, ncol = 4)
W_runs[1, 1:2] <- 0.5    # class A = mean(A_run1, A_run2)
W_runs[2, 3:4] <- 0.5    # class B = mean(B_run1, B_run2)
rownames(W_runs) <- c("A", "B")
```

A plain matrix is accepted directly as `targets` — DKGE wraps it automatically:

```{r multi-classify}
cls_multi <- dkge_classify(fit_multi,
                           targets = W_runs,   # plain matrix dispatch
                           mode = "cell_cross",
                           n_perm  = 0,
                           seed    = 101)

res_multi <- cls_multi$results[[1]]
res_multi$metrics
as.data.frame(cls_multi)
```

Run averaging happens inside the weight matrix, so each prospective LOSO fold observes
exactly one pattern per condition. This example is descriptive; an inferential run must
use the full-recomputation contract shown above. Cross-fitting guards against
basis-reuse leakage; it does not by itself establish transportability to a new
scanner, acquisition protocol, or population.

## Hyperdesign inputs and fold bridges

Labs using `multidesign::hyperdesign()` can supply design kernels and fold assignments
through coercion S3 generics. Any object implementing `as_dkge_kernel()` and
`as_dkge_folds()` flows through the existing pipeline without altering the core solvers.

```{r hyperdesign, eval=FALSE}
# Pseudocode — requires multidesign package and a user-defined hyperdesign object
library(multidesign)

hd     <- make_demo_hyperdesign()                # user-supplied helper
Kobj   <- as_dkge_kernel(hd, basis = "effect")   # list(K = ..., info = ...)

fit_hd <- dkge(
  betas       = dkge_data_from_hd(hd),
  K           = Kobj,
  keep_inputs = TRUE
)

folds  <- as_dkge_folds(fold_over(hd, over = "subject", k = 5, seed = 1), fit_hd)

res_hd <- dkge_contrast(fit_hd,
                        contrasts = c(1, -1, 0, 0),
                        method    = "kfold",
                        folds     = folds)
```

Matrices, lists with a `$K` element, and existing `dkge_folds` objects continue to work
as before — only packages with richer design structures need to implement these generics.

## Practical tips

**Target specification.** Define the full factorial structure with formula notation
(e.g., `~ cond + time + cond:time`) and reuse the same target definitions across both
classification and contrast analyses for methodological consistency.

**Backend choice.** LDA (`method = "lda"`) is fast and usually sufficient. Use
`method = "logit"` with `class_weights = "balanced"` when experimental conditions have
unequal numbers of trials.

**Permutation testing.** The beta inferential API requires one externally preselected
positive scalar `lambda` whenever `n_perm > 0`. Observed-data `lambda_grid` and
`lambda_fun` selection remain available only with `n_perm = 0`; freezing their selected
value across permutations would invalidate the randomization comparison. For `cell`
and `cell_cross`, `control$randomization_recompute` must also rebuild the DKGE fit,
rank choice, folds, representation, and classifier for each randomized label vector;
the callback returns the requested named metrics. Increase `n_perm` for precise p-values
near significance thresholds, and record how the fixed penalty was selected
independently.

**Spatial alignment.** When subjects have different voxel grids, apply transport to a
common medoid parcellation via `dkge_transport_contrasts_to_medoid()` before
classification.

**Input caching.** Keep `keep_inputs = TRUE` (default) to enable `dkge_update_weights()`
without re-running the full fit.

## Next steps

- **Adaptive weighting** (`vignette("dkge-adaptive-weighting")`): Emphasize spatially
  reliable voxels before computing the group embedding.
- **`dkge_pipeline()`**: Orchestrate fit → contrast → transport → inference →
  classification in a single call.
- **Diagnostics**: Inspect `fit$evals` (variance explained) and per-fold confusion
  matrices to identify underperforming subjects or poorly separated conditions.
