---
title: Classification with DKGE
description: >
  Cross-validate a latent classifier on DKGE components, choose between cell and
  cell_cross modes, and read the decoding result honestly.
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Classification with DKGE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

```{r library, message=FALSE}
library(dkge)
set.seed(12)
```

Use classification when the scientific question is predictive: can a condition
label be recovered from a subject's effect pattern when that subject was not
used to train the decoder? This is different from asking whether a component
explains much variation or whether a contrast is non-zero.

This page follows one complete path:

```text
factor definition -> DKGE representation -> class weight matrix -> held-out predictions -> accuracy and confusion
```

The weight matrix is the conceptual bridge. It says how named beta rows become
the class patterns that the classifier sees; the decoder never infers that
mapping from effect names by magic.

There are two distinct cross-validation contracts. The default `mode = "auto"`
uses subject-level folds but selects the faster `"cell"` path for
within-subject targets; its classifier is cross-validated, while the global
DKGE basis has seen every subject. Use `mode = "cell_cross"` when the basis
itself must also be re-estimated without each held-out subject. The primary
example below uses that stricter end-to-end path.

## What is being classified?

We define two experimental factors—condition (A versus B) and four time
points—and construct a design kernel that encodes their effect structure:

```{r design-kernel}
factors <- list(cond = list(L = 2), time = list(L = 4))
kern    <- design_kernel(factors, basis = "effect")
q       <- nrow(kern$K)   # number of effect-coded columns
q
```

We simulate 10 subjects, each with `q` design effects and 60 voxels. A
detectable condition signal is planted only in voxels 1–10. That artificial
localization gives the example a visible ground truth; the remaining voxels are
noise.

```{r simulate}
n_subjects <- 10L
v <- 60L

make_subject <- function(id) {
  # design: identity matrix over effects (one trial per effect)
  design <- diag(q)
  colnames(design) <- rownames(kern$K)

  # signal: condition contrast (first effect column) drives voxels 1-10
  signal <- matrix(0, nrow = q, ncol = v)
  signal[1, 1:10] <- 0.5    # positive signal for condition contrast

  beta <- signal + matrix(rnorm(q * v, sd = 1.0), nrow = q)
  dkge_subject(beta, design = design, id = paste0("sub", id))
}

subjects <- lapply(seq_len(n_subjects), make_subject)
```

```{r planted-signal, echo=FALSE, fig.height=3.2, fig.alt="Line plot showing a positive planted condition signal in voxels one through ten and zero signal elsewhere."}
planted_voxel_signal <- c(rep(0.5, 10), rep(0, v - 10))
plot(
  seq_len(v), planted_voxel_signal, type = "h", lwd = 3,
  col = ifelse(seq_len(v) <= 10, "#b33b2e", "grey75"),
  xlab = "Voxel", ylab = "Planted condition effect",
  main = "What the classifier has a chance to recover"
)
abline(h = 0, col = "grey60")
```

The signal plot describes the simulation, not the model output. The honest
performance question remains whether a decoder trained without one subject can
classify that held-out subject's A and B patterns.

Fit the DKGE model using the design kernel. `design_kernel()` returns a list with `$K`
(the q×q kernel matrix) and `$info` (factor metadata used by `dkge_targets()`):

```{r fit}
fit <- dkge(subjects, K = kern, rank = 2)
fit
```

## Classification targets

`dkge_targets()` maps design formula terms to classification contrasts using the factor
metadata stored in the kernel. Each term becomes a binary (or multi-class) decoding
problem over the group embedding.

```{r targets}
targets <- dkge_targets(fit, ~ cond)
length(targets)                      # one target per formula term
targets[[1]]$name                    # "cond"
nrow(targets[[1]]$weight_matrix)     # 2 classes (A vs B)
```

The weight matrix transforms subject-level beta vectors into class-specific pattern
vectors for the downstream classifier.

## Cross-validated decoding

```{r classify}
cls <- dkge_classify(
  fit,
  targets = targets,
  method  = "lda",     # "lda" (default) or "logit"
  mode    = "cell_cross",
  n_perm  = 99,        # coarse permutation resolution for this example
  seed    = 99
)
print(cls)
```

The result is a `dkge_classification` object. Each entry of `$results` corresponds to
one target:

```{r results}
res <- cls$results[["cond"]]
res$mode       # "cell_cross": the basis is refitted inside each fold
res$metrics    # cross-validated accuracy and log-loss
```

The estimand is subject-level LOSO accuracy for classifying the condition
patterns, with both the DKGE basis and classifier refit without the held-out
subject. In this seeded simulation the planted condition signal yields:

```{r interpret-result}
condition_accuracy <- unname(res$metrics[["accuracy"]])
condition_log_loss <- unname(res$metrics[["logloss"]])
c(accuracy = condition_accuracy, logloss = condition_log_loss)
```

An accuracy of 1 means every held-out condition row was classified correctly in
this deliberately strong toy example. It is not an estimate of performance on
an external dataset.

Accuracy is easier to interpret alongside the errors it summarizes:

```{r confusion-plot, echo=FALSE, fig.height=4, fig.width=4.8, fig.alt="Confusion-matrix heatmap for held-out condition predictions."}
conf_mat <- dkge_confusion(cls, target = "cond")
image(
  seq_len(ncol(conf_mat)), seq_len(nrow(conf_mat)),
  t(conf_mat[nrow(conf_mat):1, , drop = FALSE]),
  col = grDevices::hcl.colors(12, "Blues 3"), axes = FALSE,
  xlab = "Predicted class", ylab = "Observed class",
  main = "Held-out predictions"
)
axis(1, at = seq_len(ncol(conf_mat)), labels = colnames(conf_mat))
axis(2, at = seq_len(nrow(conf_mat)), labels = rev(rownames(conf_mat)))
for (i in seq_len(nrow(conf_mat))) {
  for (j in seq_len(ncol(conf_mat))) {
    text(j, nrow(conf_mat) - i + 1, labels = conf_mat[i, j], font = 2)
  }
}
box()
```

Rows are observed classes and columns are predictions. A strong diagonal means
the held-out labels were recovered; off-diagonal counts reveal which classes
are confused. For imbalanced classes, always pair this display with class
counts and a metric such as balanced accuracy or log-loss.

Convert to a tidy data frame for plotting or downstream analysis:

```{r tidy}
df <- as.data.frame(cls)
head(df)
```

### Permutation p-values

When `n_perm > 0`, each target gets an empirical p-value from the sign-flip max-T
permutation distribution:

```{r pvalues}
res$p_values    # named numeric vector per metric (NULL when n_perm = 0)
```

With 99 permutations and the plus-one correction, the smallest attainable
p-value is (1/(99+1)=0.01). A value at that boundary means none of the sampled
permutations was as extreme; it does not provide finer resolution than 0.01.

### Subject-level predictions

Individual predictions and fold assignments live in `$row_data`:

```{r predictions}
head(res$row_data[, c("subject_label", "class_label", "fold")])
```

Full predicted probabilities are in `res$probabilities` (rows = subject × class
combinations, columns = class labels).

### Per-fold diagnostics

Confusion matrices and class counts per fold are in `res$diagnostics$folds`:

```{r diag}
diag_fold1 <- res$diagnostics$folds[[1]]
diag_fold1$confusion           # confusion matrix for fold 1
diag_fold1$class_counts_test   # observed class counts in test set
```

## Two modes: `cell` vs `cell_cross`

The `mode` argument controls how the group basis is applied at test time:

| Mode | Basis used | Use case |
|------|-----------|----------|
| `"cell"` | Global `fit$U` | Fast; mild basis-step leakage — `fit$U` saw all subjects |
| `"cell_cross"` | Fold-specific LOSO `U_fold` | Basis and classifier refit per subject fold |
| `"delta"` | Global `fit$U` + subject labels | Subject-level binary test; requires `y` argument |

`"auto"` (default) selects `"cell"` for within-subject targets and `"delta"` for
between-subject targets.

The strict route refits the basis for every held-out subject, so it is shown
here for syntax rather than run:

```{r cell-cross, eval=FALSE}
# Refit the basis without each held-out subject.
cls_strict <- dkge_classify(fit, targets = targets, mode = "cell_cross", n_perm = 0)
```

## Multiple betas per condition

When you have multiple beta estimates per condition (e.g., separate scanner runs), stack
the run-specific effects as extra rows in the design matrix and supply a weight matrix
that combines them into a single class pattern.

```{r multi-beta}
# 4 effects: A_run1, A_run2, B_run1, B_run2
make_subject_multi <- function(id) {
  design <- diag(4)
  colnames(design) <- c("A_run1", "A_run2", "B_run1", "B_run2")

  signal <- matrix(0, nrow = 4, ncol = v)
  signal[1:2, 1:10] <-  0.5   # both A runs share signal
  signal[3:4, 1:10] <- -0.5   # both B runs share the opposite signal

  beta <- signal + matrix(rnorm(4 * v, sd = 1.0), nrow = 4)
  dkge_subject(beta, design = design, id = paste0("sub", id))
}

subjects_multi <- lapply(seq_len(n_subjects), make_subject_multi)
fit_multi      <- dkge(subjects_multi, K = diag(4), rank = 2)

# Weight matrix: rows = classes, columns = effects.
# Each class averages its two runs.
W_runs <- matrix(0, nrow = 2, ncol = 4)
W_runs[1, 1:2] <- 0.5    # class A = mean(A_run1, A_run2)
W_runs[2, 3:4] <- 0.5    # class B = mean(B_run1, B_run2)
rownames(W_runs) <- c("A", "B")
```

A plain matrix is accepted directly as `targets` — DKGE wraps it automatically:

```{r multi-classify}
cls_multi <- dkge_classify(fit_multi,
                           targets = W_runs,   # plain matrix dispatch
                           mode = "cell_cross",
                           n_perm  = 49,
                           seed    = 101)

res_multi <- cls_multi$results[[1]]
res_multi$metrics
as.data.frame(cls_multi)
```

Run averaging happens inside the weight matrix, so each fold observes one
pattern per condition. Because this call also uses `mode = "cell_cross"`, both
the basis and classifier exclude the held-out subject. This guards against the
specific basis-reuse leakage described above; it does not by itself establish
transportability to a new scanner, acquisition protocol, or population.

## Hyperdesign inputs and fold bridges

Labs using `multidesign::hyperdesign()` can supply design kernels and fold assignments
through coercion S3 generics. Any object implementing `as_dkge_kernel()` and
`as_dkge_folds()` flows through the existing pipeline without altering the core solvers.

The block below is illustrative and does not run here: it needs the
`multidesign` package and a hyperdesign object you supply yourself.

```{r hyperdesign, eval=FALSE}
library(multidesign)

hd     <- make_demo_hyperdesign()                # user-supplied helper
Kobj   <- as_dkge_kernel(hd, basis = "effect")   # list(K = ..., info = ...)

fit_hd <- dkge(
  betas       = dkge_data_from_hd(hd),
  K           = Kobj,
  keep_inputs = TRUE
)

folds  <- as_dkge_folds(fold_over(hd, over = "subject", k = 5, seed = 1), fit_hd)

res_hd <- dkge_contrast(fit_hd,
                        contrasts = c(1, -1, 0, 0),
                        method    = "kfold",
                        folds     = folds)
```

Matrices, lists with a `$K` element, and existing `dkge_folds` objects continue to work
as before — only packages with richer design structures need to implement these generics.

## Practical tips

**Target specification.** Define the full factorial structure with formula notation
(e.g., `~ cond + time + cond:time`) and reuse the same target definitions across both
classification and contrast analyses for methodological consistency.

**Backend choice.** LDA (`method = "lda"`) is fast and usually sufficient. Use
`method = "logit"` with `class_weights = "balanced"` when experimental conditions have
unequal numbers of trials.

**Permutation testing.** Increase `n_perm` for precise p-values near significance
thresholds. Set `n_perm = 0` during exploratory analysis to skip the permutation loop.

**Spatial alignment.** When subjects have different voxel grids, apply transport to a
common medoid parcellation via `dkge_transport_contrasts_to_medoid()` before
classification.

**Input caching.** Keep `keep_inputs = TRUE` (default) to enable `dkge_update_weights()`
without re-running the full fit.

## Next steps

- **Adaptive weighting** (`vignette("dkge-adaptive-weighting")`): Emphasize spatially
  reliable voxels before computing the group embedding.
- **`dkge_pipeline()`**: Orchestrate fit → contrast → transport → inference →
  classification in a single call.
- **Diagnostics**: Inspect `fit$evals` (variance explained) and per-fold confusion
  matrices to identify underperforming subjects or poorly separated conditions.
