---
title: CPCA Filtering
description: >
  Split the compressed covariance into design-aligned and residual parts
  before eigendecomposition.
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{CPCA Filtering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(dkge)
library(ggplot2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
options(
  tibble.print_min = 4,
  tibble.print_max = 4,
  digits = 3
)
set.seed(2025)
if ("theme_dkge" %in% ls("package:dkge")) {
  theme_set(dkge::theme_dkge())
}
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

Use CPCA filtering when you have already decided that one prespecified span of
effects should be studied separately from its complement. A common example is
to ask for components aligned with main effects while keeping interaction and
control directions in a residual subspace. Without that split, the leading
DKGE component may mix all of them simply because they covary strongly.

CPCA does **not** discover which scientific partition is correct. You supply
the partition; DKGE projects the compressed moment into design-aligned and
residual parts before fitting their bases. The two parts are
\(K\)-orthogonal, which is a geometric statement, not evidence that the
underlying cognitive processes are independent.

```text
mixed compressed moment -> declared effect span + K-orthogonal complement -> separate latent bases
```

This page first shows the default mixed fit, then applies one transparent split
and visualizes what moved. Custom bases, kernel sensitivity, and ridge
stabilization follow as advanced branches.

## When is the split scientifically useful?

Use CPCA filtering when a design carries several kinds of task effect that you want analyzed separately. Rather than analyzing all effects together in a single latent space, CPCA allows you to focus your analysis on specific effect types while cleanly separating others.

Common reasons to declare a split include:

**Factorial designs**: Separate main effects from interaction terms to understand how basic experimental manipulations differ from their combined effects. For example, in a 2×2 design studying attention and working memory, you might isolate the main effects of each factor from their interaction.

**Experimental versus control conditions**: Focus analysis on your experimental manipulations while factoring out baseline or control conditions. This approach can reveal cleaner patterns in your conditions of interest.

**Multi-domain studies**: When studying different cognitive processes within the same experiment, separate effects related to different domains (e.g., working memory versus attention) to understand domain-specific versus shared neural mechanisms.

**Planned versus exploratory contrasts**: Isolate your primary hypotheses from exploratory or secondary analyses, ensuring that your main effects of interest receive focused statistical attention.

**Comparative studies**: Apply the same declared effect spans across datasets,
which makes the fitted targets easier to compare. Reliability still has to be
measured rather than inferred from the decomposition.

## Where does filtering enter the fit?

The compressed covariance \(\hat C\) that DKGE analyzes contains variance from all experimental effects mixed together. CPCA filtering mathematically separates this total variance into distinct subcomponents before eigendecomposition. The process involves several key steps that preserve the mathematical relationships defined by your design kernel \(K\).

First, DKGE constructs a projector onto your chosen effect subspace using the \(K\) metric through `dkge_projector_K()`. This projector respects the similarity relationships between experimental effects that are encoded in your design kernel. Next, `dkge_cpca_split_chat()` applies this projector to split the compressed covariance into design and residual components. Finally, DKGE fits separate bases for the components you request while preserving \(K\)-orthogonality between the returned bases, ensuring they can be analyzed jointly.

You can specify the design-aligned subspace either by naming specific effects through `cpca_blocks` or by providing an explicit basis matrix via `cpca_T`. The `cpca_part` argument determines which filtered components are returned (`"design"`, `"resid"`, or `"both"`), while `cpca_ridge` optionally adds regularization to stabilize small eigenvalues during decomposition.

## Simulated Experiment: Attention-Working Memory Study

We simulate a factorial attention-working memory experiment: two main effects
(attention cue validity and working memory load), their interaction, and
additional control conditions. That structure suits CPCA, where the goal is to
isolate the main experimental effects from their interaction and the control
conditions.

The simulated dataset contains strong signals in the primary experimental manipulations (attention and working memory main effects) and weaker but meaningful variance in the interaction term and control conditions. This mirrors real neuroimaging studies where primary effects of interest typically show stronger and more consistent patterns than secondary effects.

```{r simulate-data}
S <- 8
q <- 6
P <- 16
Tlen <- 80
effects <- c("attn_valid", "attn_invalid", "wmem_high", "wmem_low", "interact", "control")

betas <- replicate(S, {
  # Strong signals for main experimental effects
  main_effects <- matrix(rnorm(2 * P, sd = 1.5), 2, P)
  # Weaker signals for interaction and control conditions
  secondary_effects <- matrix(rnorm((q - 2) * P, sd = 0.4), q - 2, P)
  mat <- rbind(main_effects, secondary_effects)
  rownames(mat) <- effects
  mat
}, simplify = FALSE)

designs <- replicate(S, {
  X <- matrix(rnorm(Tlen * q), Tlen, q)
  X <- qr.Q(qr(X))
  colnames(X) <- effects
  X
}, simplify = FALSE)

subjects <- Map(function(b, X, id) dkge_subject(b, X, id = id),
                betas, designs, paste0("sub", seq_len(S)))
bundle <- dkge_data(subjects)
```

When we fit the standard DKGE model without CPCA filtering, all task-related variance is compressed into a single latent basis. The leading eigenvalues reflect the combined influence of both the primary experimental effects and secondary conditions, making it difficult to isolate the specific patterns we want to study.

```{r fit-standard}
fit_plain <- dkge(bundle, K = diag(q), rank = 3)
round(fit_plain$evals[1:4], 3)
```

These eigenvalues represent the mixed signal from all experimental conditions.
The standard fit captures the dominant patterns but does not separate the
primary experimental manipulations from their interactions and the control
conditions.

## Isolating Primary Experimental Effects

CPCA filtering separates the primary experimental effects (attention and working memory main effects) from the secondary effects (interaction and control conditions). We achieve this by identifying the first two effects in the design as the "design-aligned" subspace through `cpca_blocks = 1:2`.

Setting `cpca_part = "both"` instructs DKGE to return both the design-aligned basis (focused on the primary effects) and the residual basis (containing the secondary effects). The two returned bases are orthogonal under the chosen \(K\) metric; this geometric property does not imply statistical independence.

```{r fit-cpca}
fit_cpca <- dkge(bundle,
                 K = diag(q),
                 cpca_blocks = 1:2,
                 cpca_part = "both",
                 rank = 3)

fit_cpca$cpca$part
round(fit_cpca$cpca$evals_design[1:3], 3)
round(fit_cpca$cpca$evals_resid[1:3], 3)
```

```{r cpca-spectrum, echo=FALSE, fig.height=3.8, fig.alt="Grouped bar chart comparing the first three eigenvalues in the design-aligned and residual CPCA subspaces."}
cpca_spectrum <- rbind(
  data.frame(component = 1:3, subspace = "Design aligned",
             eigenvalue = fit_cpca$cpca$evals_design[1:3]),
  data.frame(component = 1:3, subspace = "Residual",
             eigenvalue = fit_cpca$cpca$evals_resid[1:3])
)
ggplot2::ggplot(
  cpca_spectrum,
  ggplot2::aes(factor(component), eigenvalue, fill = subspace)
) +
  ggplot2::geom_col(position = "dodge", width = 0.7) +
  ggplot2::labs(
    title = "The declared split reorganizes the fitted covariance",
    x = "Component within subspace", y = "Eigenvalue", fill = NULL
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

The two eigenvalue sequences summarize the covariance retained by the chosen
design and residual projectors. They provide a focused decomposition under this
model, not evidence that the underlying biological processes are independent.

The following check verifies \(K\)-orthogonality between the design and residual
bases. It is a numerical statement about the fitted coordinates, not a test of
independence or absence of shared signal:

```{r orthogonality}
Ud <- fit_cpca$cpca$U_design
Ur <- fit_cpca$cpca$U_resid
round(max(abs(t(Ud) %*% fit_cpca$K %*% Ur)), 6)
```

### Influence of the Design Kernel

The split is metric-aware: changing `kernel` alters which directions count as
"design-aligned." A smooth kernel diffuses the projector across neighboring
rows, so design energy leaks into adjacent effects.

```{r kernel-effect}
K_smooth <- outer(seq_len(q), seq_len(q), function(i, j) 0.7^abs(i - j))
fit_kernel <- dkge(bundle,
                   K = K_smooth,
                   cpca_blocks = 1:2,
                   cpca_part = "both",
                   rank = 3)
round(fit_kernel$cpca$evals_design[1:3], 3)
round(fit_kernel$cpca$evals_resid[1:3], 3)
```

The printed eigenvalues show how much fitted covariance each projector retains
under `K_smooth`. They do not display the loadings themselves. The projector
honours the correlation structure encoded by `K_smooth`, so the kernel choice
shapes which latent directions count as design-aligned. When effects do not
align with coordinate axes, pass a custom `cpca_T` that expresses the intended
K-weighted span explicitly.

Behind the scenes, DKGE accomplishes this separation through mathematical projectors that operate in the design kernel metric. The function `dkge_cpca_split_chat()` applies these projectors to split the compressed covariance matrix before eigendecomposition. You can examine this split directly to understand how the total variance is partitioned:

```{r inspect-split}
T_design <- diag(1, q)[, 1:2]
split_plain <- dkge_cpca_split_chat(fit_plain$Chat, T_design, fit_plain$K)
round(diag(split_plain$Chat_design), 3)
round(diag(split_plain$Chat_resid), 3)
```

The diagonal elements show how variance is distributed between design and residual components. Notice that when `cpca_part = "design"` or `"both"`, the fitted model's compressed covariance matrix (`fit_cpca$Chat`) equals the design-filtered component, confirming that the analysis focuses specifically on your chosen effects:

```{r chat-match}
max(abs(fit_cpca$Chat - fit_cpca$cpca$Chat_design))
```

## Using Custom Effect Combinations

Sometimes the effects of interest do not correspond to simple subsets of experimental conditions. In such cases, you can provide a custom basis matrix through `cpca_T` to specify exactly which combinations of effects should be treated as "design-aligned."

The columns of your custom basis matrix span the linear combinations of effects you want to analyze together. For example, you might want to combine specific experimental conditions or weight certain effects more heavily than others. Here we demonstrate by creating a custom basis that combines multiple conditions with differential weighting:

```{r explicit-basis}
T_custom <- qr.Q(qr(cbind(c(1, 1, 0, 0, 0, 0),
                          c(0, 0, 2, 1, 0, 0))))
fit_custom <- dkge(bundle,
                   K = diag(q),
                   cpca_T = T_custom,
                   cpca_part = "design",
                   rank = 2)
round(fit_custom$cpca$evals_design[1:2], 3)
```

This custom basis creates two design components: the first combines the attention conditions equally, while the second emphasizes the working memory conditions with higher weight on the high-load condition. The fitted loadings reflect these chosen combinations, while the residual component is omitted since we requested only the design-aligned analysis.

## Numerical Stabilization with Ridge Regularization

When working with real neuroimaging data, you may encounter situations where the filtered covariance matrix becomes nearly rank-deficient, leading to numerical instability during eigendecomposition. The optional `cpca_ridge` parameter addresses this issue by adding a small diagonal ridge term before eigendecomposition, improving numerical stability without substantially altering the results.

```{r ridge}
fit_ridge <- dkge(bundle,
                  K = diag(q),
                  cpca_blocks = 1:2,
                  cpca_part = "design",
                  cpca_ridge = 1e-3,
                  rank = 3)
diag_shift <- diag(fit_ridge$cpca$Chat_design - fit_ridge$cpca$Chat_design_raw)
round(head(diag_shift), 6)
```

The ridge regularization adds the specified value to each diagonal element of the covariance matrix, as shown by the consistent shift across diagonal entries. The original unregularized matrix remains available in `Chat_design_raw` for comparison and diagnostic purposes.

## Simplified Interface for CPCA Analysis

For workflows that focus exclusively on CPCA filtering, the `dkge_cpca_fit()` function provides a streamlined interface that reduces code verbosity. This convenience wrapper handles the CPCA-specific arguments while forwarding all other parameters to the main `dkge()` function, making your analysis code more concise and readable.

```{r wrapper}
fit_wrapper <- dkge_cpca_fit(bundle,
                             K = diag(q),
                             cpca_blocks = 1:2,
                             cpca_part = "design",
                             rank = 3)
identical(round(fit_wrapper$U, 6), round(fit_cpca$U, 6))
```

## What this page showed

The split is a modelling choice, not a discovery. You declare which effects are
design-aligned — by name through `cpca_blocks` or as an explicit basis through
`cpca_T` — and `dkge_projector_K()` builds the projector in the \(K\) metric.
`dkge_cpca_split_chat()` applies it, and DKGE fits a basis for each requested
part, \(K\)-orthogonal to the other.

The returned bases are ordinary DKGE components: contrast testing, bootstrap
inference and the plotting helpers all work on them unchanged.

Whether the split improves power is design- and data-dependent, and should be
checked against an unfiltered fit rather than assumed.

## Where to go next

- `vignette("dkge-design-kernels")` — the kernel that defines which directions
  count as design-aligned, and therefore what CPCA is splitting.
- `vignette("dkge-components")` — how to read the components that come back
  from either side of the split.
- `vignette("dkge-concepts")` — why a \(K\)-orthogonal split is a statement
  about coordinates and not about independent processes.
