---
title: Transport and Dense Rendering
description: >
  Transport subject-specific cluster values onto a shared anchor space and
  render them as a dense field.
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Transport and Dense Rendering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
has_albers_v2 <- requireNamespace("albersdown", quietly = TRUE) &&
  utils::packageVersion("albersdown") >= base::package_version("2.0.0")
if (requireNamespace("systemfonts", quietly = TRUE) && has_albers_v2) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && has_albers_v2) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(dkge)
library(ggplot2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
options(
  tibble.print_min = 4,
  tibble.print_max = 4,
  digits = 3
)
set.seed(2024)
theme_set(theme_dkge())
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```


DKGE components and contrasts initially live on each subject's own clusters.
You cannot average those vectors merely because they have similar lengths:
cluster indices are not anatomical correspondences. Dense rendering makes the
spatial assumption explicit by mapping every subject through shared anchors
and, optionally, decoding the anchor field onto a voxel grid.

```text
subject cluster values -> fitted subject-to-anchor maps -> shared anchor field -> voxel interpolation
```

This page preserves the deprecated geometry-only renderer as a worked example
of rendering mechanics. It is descriptive, cannot produce valid group
inference, and must not be mistaken for functional alignment. For the typed
workflow in which independent response signatures identify correspondence,
read `vignette("dkge-functional-alignment")` instead.

## What spatial information is required?

The rendering pipeline needs coordinates, so the example below carries both
beta coefficients and spatial centroids for each subject.

```{r data}
library(dkge)
S <- 4; q <- 3; P <- 20; T <- 50
betas <- replicate(S, matrix(rnorm(q * P), q, P), simplify = FALSE)
designs <- replicate(S, {
  X <- matrix(rnorm(T * q), T, q)
  qr.Q(qr(X))
}, simplify = FALSE)
centroids <- replicate(S, matrix(runif(P * 3, -40, 40), P, 3), simplify = FALSE)
subjects <- lapply(seq_len(S), function(s) dkge_subject(betas[[s]], designs[[s]], id = paste0("sub", s)))
bundle <- dkge_data(subjects)
fit <- dkge(bundle, K = diag(q), rank = 2)
fit$centroids <- centroids  # attach for transport helpers; pass explicitly in production
```

## Legacy descriptive anchor rendering

`dkge_build_renderer()` is deprecated. It maps subject-specific cluster
coordinates onto a shared anchor space for descriptive display, defaults to
equal subject weights, and records an inferentially ineligible receipt. A bare
anchor or MNI grid supplies display coordinates; it supplies no functional
correspondence to any subject.
This legacy example uses a deterministic geometric k-nearest-neighbor mapper;
the functional Sinkhorn workflow and its numerical eligibility gates are
covered in `vignette("dkge-functional-alignment")`.

```{r renderer}
# use 5k anchors sampled from grey-matter-like cube
vox_xyz <- as.matrix(expand.grid(seq(-40, 40, by = 10), seq(-40, 40, by = 10), seq(-40, 40, by = 20)))
renderer <- suppressWarnings(dkge_build_renderer(
  fit,
  centroids = centroids,
  vox_xyz = vox_xyz,
  mapper = dkge_mapper("knn", k = 8, sigx = 3),
  graph_k = 10,
  decoder_k = 8,
  anchor_n = min(500L, nrow(vox_xyz)),
  anchor_method = "sample"
))
str(renderer, max.level = 1)
```

The renderer carries four components:

- `anchors`: the shared coordinates, derived from the `vox_xyz` grid, that
  serve as the common reference space.
- `graph`: an optional k-nearest-neighbor graph and Laplacian, which enables
  spatial smoothing across anchors.
- `mapper_fits`: Subject-specific fitted plans and target-conditional application
  operators. Reusing the renderer reuses these solved mappings.
- `mapper_stats`: transport diagnostics - costs, entropies, and support sizes -
  for assessing the mapping.

## Rendering subject values descriptively

`dkge_render_subject_values()` transports subject cluster values into the
shared anchor space, then interpolates them to the full voxel grid. It too is
deprecated and returns no inferential result.

```{r render}
values_list <- lapply(fit$Btil, function(Bts) as.numeric(Bts[1, ]))
rendered <- suppressWarnings(
  dkge_render_subject_values(renderer, values_list, lambda = 0.2, to_vox = TRUE)
)
summary(rendered$details$y)
head(rendered$details$subject_stats)
```

The render returns three things:

- `rendered$anchor`: the smoothed group field on the anchor coordinates.
  This is the primary output of the transport.
- `rendered$voxel`: the interpolated voxel map, with length `nrow(vox_xyz)`.
- `rendered$details$plan_entropy_mean`: mean diffusion of the transport plans. Higher
  values mean broader anchor coverage and more diffuse mapping patterns.

The anchor values vary along a single spatial dimension, which is the easiest
way to see the field's structure.

```{r plot-anchor, fig.height=4, fig.width=5, fig.alt="Scatter plot of anchor x-coordinate versus rendered value."}
plot(renderer$anchors[, 1], rendered$anchor, pch = 20, col = "steelblue",
     xlab = "Anchor x-coordinate", ylab = "Rendered value",
     main = "Rendered anchor field")
abline(h = 0, col = "grey70", lty = 2)
```

## Voxel Map Snapshot

Slices of the voxel grid show how the anchor-based smooth field is
interpolated onto the target space.

```{r plot-vox, fig.height=4, fig.width=5, fig.alt="Scatter plot of voxel slice values for z equals zero."}
sel <- vox_xyz[, 3] == 0  # slice at z = 0
plot(vox_xyz[sel, 1], rendered$voxel[sel], pch = 16, col = "tomato",
     xlab = "x", ylab = "value", main = "Voxel slice (z=0)")
```

## What to check before reporting

Three settings change the result:

- `lambda` in `dkge_render_subject_values()` trades fidelity to the original
  data against spatial smoothness. Watch `plan_entropy_mean` to judge whether a
  setting smooths too much or too little.
- `subject_feats` and `anchor_feats` in `dkge_build_renderer()` let the
  descriptive transport use additional latent features. They do not establish
  fold-safe or independent functional correspondence.
- Reuse a renderer for repeated maps: its fitted plans avoid solving the same
  subject-to-anchor mappings again. Separately, DKGE keeps a process-local cache
  of converged Sinkhorn duals for byte-identical fit problems; inspect
  `mapper_fits[[s]]$stats$diagnostics$cache_hit` when diagnosing a repeated fit.

## Where to go next

- `vignette("dkge-anchors")` builds the shared anchor representation this page
  renders onto, when items rather than coordinates define the alignment.
- `vignette("dkge-performance")` compares the mappers and shows when warm starts
  and caching pay for themselves.
- `vignette("dkge-weighting")` covers the transport weighting layer, which
  changes what a rendered field means.
- `vignette("dkge-functional-alignment")` is the supported typed workflow for
  functional correspondence, aligned subject rows, group inference, and
  rendering on one identified reference support.
