---
title: Weighting Strategies in DKGE
description: >
  Separate the four weighting layers - location, subject, spatial, and transport -
  and see which object each one changes.
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Weighting Strategies in DKGE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(dkge)
library(ggplot2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
options(
  tibble.print_min = 4,
  tibble.print_max = 4,
  digits = 3
)
set.seed(2024)
theme_set(theme_dkge())
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```


DKGE has separate weighting controls for locations within a subject, whole
subjects, and spatial transport. Each changes a different estimand. This
vignette shows where the controls enter the computation and what must be
reported when they are used.

Weighting is needed when “one unit, one vote” is scientifically wrong—for
example, when parcels have unequal area or some measurements are demonstrably
less reliable. It is not a generic way to make a result cleaner. Every weight
answers two questions: *what is the unit being reweighted, and at what stage?*

```text
subject beta block --spatial weights--> subject moment --subject weights--> group basis --transport masses--> common map
```

### Adaptive voxel weighting (preview)

DKGE can also rescale locations adaptively before fold bases are constructed.
`dkge_weights()` combines a prespecified prior, such as an external reliability
map, with training-fold statistics such as design-kernel energy or precision.
Training-fold scope closes one leakage path but does not establish calibration
or improved sensitivity. `vignette("dkge-adaptive-weighting")`
shows the specification, structural checks, and post-hoc updates with
`dkge_update_weights()`.

## Which weighting control changes which object?

| Layer | Unit receiving weight | Argument | Enters before | Typical rationale |
|---|---|---|---|---|
| Spatial | parcel, cluster, or voxel within subject | `Omega_list` / `omega` | subject moment | area, precision, or a declared spatial metric |
| Subject | entire subject block | `w_method`, `w_tau`, or `weights` | pooled eigensolve | prevent one high-energy block from dominating, or apply external subject mass |
| Transport | source and target spatial mass | `sizes` | common-map summary | preserve parcel area or voxel density during alignment |

Effect-by-subject precision weights for unequal trial counts are a fourth,
separate layer covered in `vignette("dkge-partial-effect-spaces")`, with the
precision-weighted mean derived in `vignette("dkge-unbalanced-trialwise")`.
Adaptive
training-fold weights are covered in `vignette("dkge-adaptive-weighting")`.

The example below changes one layer at a time so that a changed output can be
attributed to a named assumption.

## Example Dataset

The dataset below has three subjects with different cluster counts, sizes, and
reliability scores. That heterogeneity is what the weighting controls act on.

```{r example-data, message=FALSE, warning=FALSE}
library(dkge)
library(purrr)
set.seed(42)

S <- 3
q <- 4
P <- c(6, 5, 7)
T <- 40

betas <- map(P, ~ matrix(rnorm(q * .x), q, .x))
designs <- map(seq_len(S), function(s) {
  X <- matrix(rnorm(T * q), T, q)
  qr.Q(qr(X))
})

centroids <- map(P, ~ matrix(runif(.x * 3, min = -20, max = 20), .x, 3))
cluster_sizes <- map(P, ~ sample(10:40, .x, replace = TRUE))
reliability_scores <- list(
  rep(0.95, P[1]),
  rep(0.20, P[2]),  # subject 2 is the deliberately low-reliability block
  rep(0.80, P[3])
)
```

`cluster_sizes` stands in for surface area, and `reliability_scores` for a
per-cluster quality metric such as split-half correlation. Both are supplied
by you; DKGE does not compute either.

## Baseline Fit (No Spatial Weights)

```{r baseline-fit}
bundle <- dkge_data(betas, designs = designs)
k_identity <- diag(bundle$q)

fit_equal <- dkge(bundle, K = k_identity, rank = 2, w_method = "none")
fit_equal$weights
```

With equal weights every subject contributes uniformly to the shared basis, so
cluster-level variation reflects the planted signal rather than any weighting
choice.

## Size-Weighted Blocks

Per-cluster mass values emphasise larger parcels while keeping a diagonal
information structure, so clusters stay independent of one another.

```{r size-weighting}
omega_size <- cluster_sizes

fit_size <- dkge(bundle,
                 K = k_identity,
                 Omega_list = omega_size,
                 rank = 2,
                 w_method = "mfa_sigma1",
                 w_tau = 0.25)
fit_size$weights
```

The subject blocks now differ in total weighted energy, because size weighting
scales each block by its own mass. `w_method` decides whether that difference
survives into the basis.

```{r size-loadings}
size_loadings <- lapply(fit_size$Btil, function(Bts) t(Bts) %*% fit_size$K %*% fit_size$U)
map(size_loadings, ~ round(.x[, 1, drop = FALSE], 3))
```

## Reliability Weighting

Reliability scores enter the same `omega` slot as size weights: supply a per-
cluster vector and the fit downweights clusters you have flagged as noisy.

```{r reliability-weighting}
omega_reliability <- map2(cluster_sizes, reliability_scores, `*`)

fit_reliability <- dkge(bundle,
                        K = k_identity,
                        Omega_list = omega_reliability,
                        rank = 2,
                        w_method = "mfa_sigma1",
                        w_tau = 0.25)

weight_comparison <- rbind(
  size = fit_size$weights,
  size_x_reliability = fit_reliability$weights
)
round(weight_comparison, 3)

weighted_contribution_norm <- function(fit) {
  fit$weights * vapply(
    fit$contribs,
    function(x) sqrt(sum(x * x)),
    numeric(1)
  )
}
leverage_comparison <- rbind(
  size = weighted_contribution_norm(fit_size),
  size_x_reliability = weighted_contribution_norm(fit_reliability)
)
round(leverage_comparison, 1)
```

```{r reliability-leverage-plot, echo=FALSE, fig.height=3.8, fig.alt="Grouped bar chart showing how size-only and size-times-reliability weighting change each subject's contribution scale."}
leverage_long <- rbind(
  data.frame(subject = paste0("Subject ", seq_len(S)),
             scheme = "Size only", value = leverage_comparison["size", ]),
  data.frame(subject = paste0("Subject ", seq_len(S)),
             scheme = "Size × reliability",
             value = leverage_comparison["size_x_reliability", ])
)
ggplot(leverage_long, aes(subject, value, fill = scheme)) +
  geom_col(position = "dodge", width = 0.72) +
  labs(
    title = "Inspect contribution, not the scalar subject weight alone",
    x = NULL, y = "Contribution scale entering Chat", fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

The two visible rows are deliberately different. Subject 2 has the lowest
reliability scores. Its scalar MFA weight increases because inverse-leading-
singular-value scaling partly compensates for a lower-energy block; the scalar
weight alone is therefore not a measure of leverage. The second table reports
the direct contribution scale entering `Chat` (subject weight times the
Frobenius norm of that subject's unweighted contribution). On this controlled
example, adding reliability reduces subject 2's contribution relative to the
size-only fit.

```{r reliability-weighting-checks, include=FALSE}
stopifnot(!isTRUE(all.equal(fit_reliability$weights, fit_size$weights)))
stopifnot(fit_reliability$weights[2] > fit_size$weights[2])
stopifnot(leverage_comparison["size_x_reliability", 2] <
            leverage_comparison["size", 2])
```

In real data, a reliability vector might come from inverse first-level
variance, split-half or test-retest stability, a prespecified quality-control
measure, or resampling stability. The source matters: a weight estimated from
the same held-out outcome can leak information, while an external or
training-fold estimate has a clearer interpretation. When combining area and
reliability, multiply only when the resulting estimand is intended, then
rescale to a transparent reference such as mean one.

Keep weight values positive, and add a small ridge such as `+ 1e-6` when a
weight construction can produce zeros.

## Spatial Covariance (Smoothness) Weighting

With cluster coordinates you can replace diagonal weights with a smooth
covariance matrix. The example below builds a Gaussian affinity matrix, which
rewards neighboring parcels for activating together.

```{r smoothness-weighting}
gaussian_cov <- function(coords, bandwidth = 15) {
  D <- as.matrix(dist(coords))
  K <- exp(-(D / bandwidth)^2)
  (K + t(K)) / 2 + diag(1e-6, nrow(K))
}

omega_smooth <- lapply(centroids, gaussian_cov)

fit_smooth <- dkge(bundle,
                   K = k_identity,
                   Omega_list = omega_smooth,
                   rank = 2,
                   w_method = "none")
fit_smooth$weights
```

These matrices define a spatial metric: DKGE right-multiplies each beta block by
the symmetric square root of `Omega`. That is weighting or smoothing, not
whitening (which would use an inverse square root). A Gaussian affinity couples
nearby parcels in the fitted energy calculation. Whether that coupling is
scientifically helpful is an assumption to check against an uncoupled fit, not
an automatic correction for sharp boundaries.

## Subject-Level Scaling

`w_method` controls block-level reweighting, applied after the spatial weights
have acted on individual clusters. The default Multiple Factor Analysis
scaling shrinks dominant subjects by the inverse squared leading singular
value of their contribution.

```{r subject-weighting}
fit_mfa <- dkge(bundle,
                K = k_identity,
                Omega_list = omega_reliability,
                rank = 2,
                w_method = "mfa_sigma1",
                w_tau = 0.25)

fit_energy <- dkge(bundle,
                   K = k_identity,
                   Omega_list = omega_reliability,
                   rank = 2,
                   w_method = "energy",
                   w_tau = 0)

rbind(mfa = fit_mfa$weights,
      energy = fit_energy$weights,
      size_mfa = fit_size$weights)
```

```{r subject-weighting-checks, include=FALSE}
stopifnot(!isTRUE(all.equal(fit_mfa$weights, fit_size$weights)))
stopifnot(!isTRUE(all.equal(fit_energy$weights, fit_size$weights)))
```

`w_tau` shrinks extreme weights toward equal subject contributions, which
matters most when one subject would otherwise dominate.

## Weighting During Transport

Transport takes its own `sizes` argument. It mirrors `omega` but acts at a
different stage, *after* the components are estimated. Supplying consistent
sizes there keeps the medoid representation faithful to parcel surface area.

```{r transport-weighting}
fit_mfa$centroids <- centroids  # attach for transport helpers; pass explicitly in production

comp_equal <- dkge_component_stats(
  fit_mfa,
  mapper = dkge_mapper_spec("sinkhorn", epsilon = 0.05, max_iter = 2000),
  centroids = centroids,
  inference = list(type = "parametric"),
  medoid = 1L
)

comp_weighted <- dkge_component_stats(
  fit_mfa,
  mapper = dkge_mapper_spec("sinkhorn", epsilon = 0.05, max_iter = 2000),
  centroids = centroids,
  sizes = cluster_sizes,
  inference = list(type = "parametric"),
  medoid = 1L
)

data.frame(equal = head(comp_equal$summary$stat),
           size_weighted = head(comp_weighted$summary$stat))
```

```{r transport-weighting-checks, include=FALSE}
n_medoid <- P[[1]]
stopifnot(nrow(comp_equal$summary) == n_medoid * fit_mfa$rank)
stopifnot(nrow(comp_weighted$summary) == n_medoid * fit_mfa$rank)
stopifnot(ncol(comp_equal$transport[[1]]) == n_medoid)
stopifnot(!isTRUE(all.equal(comp_equal$summary$stat, comp_weighted$summary$stat)))
```

Reweighting during transport shifts the medoid summary even when the component
scores are identical, because it changes how cluster values are pooled onto
the reference parcellation.

## What to check before reporting

Construct subjects with `dkge_subject()`: weights stay attached to the data
through pipelines and streaming loaders. For voxelwise analyses, pass cluster
sizes to `dkge_transport_to_voxels()` via `sizes`, or volumetric weighting is
lost in the mapping.

Keep vectors positive and add a mild ridge such as `+ 1e-6` to covariance
matrices. Revisit each fit's `weights`, singular values, and eigenspectra
whenever you change a weighting assumption; they show how contributions are
balanced across subjects.

These controls tailor an analysis to study-specific quality metrics or prior
spatial beliefs without changing the core fitting pipeline.

## Where to go next

- `vignette("dkge-adaptive-weighting")` derives weights from
  the data instead of from a prior.
- `vignette("dkge-partial-effect-spaces")` covers effect-by-subject
  precision, which is a separate layer from these four.
- `vignette("dkge-dense-rendering")` shows where
  transport weighting acts.
