---
title: "Adaptive Voxel Weighting in DKGE"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: false
    df_print: paged

vignette: >
  %\VignetteIndexEntry{Adaptive Voxel Weighting in DKGE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(dkge)
library(ggplot2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
options(
  tibble.print_min = 4,
  tibble.print_max = 4,
  digits = 3
)
set.seed(2024)
theme_set(theme_dkge())
```


**See also:** The [Weighting Strategies](dkge-weighting.html) vignette covers the
three complementary weighting *layers* — spatial (`Omega_list`), subject-level
(`w_method`/`w_tau`), and transport (`sizes`). This vignette focuses on the
separate `dkge_weights()` API for training-fold *voxel-level* adaptive
weighting. The mechanisms enter different stages of the computation and can be
combined when their joint estimand is explicit.

---

DKGE fits are driven by cross-subject second moments. Equal voxel weights can
let large noisy regions dominate those moments, but adaptive weights can also
create leakage if they use held-out data or change under a permutation. This
vignette shows how to define, inspect, and structurally check a weighting rule.
Those checks are necessary safeguards; they are not a substitute for
design-specific null calibration.

We will proceed through four main steps to build understanding:

1. First, we clarify the terminology and concepts used by `dkge_weights()` to establish a foundation;
2. Next, we build several weight specifications and demonstrate how to fit DKGE models with them;
3. Then, we inspect the effects of mixing prior knowledge with adaptive statistics; and
4. Finally, we run quick safety checks and demonstrate post-hoc weight updates.

The examples throughout this vignette use small synthetic datasets so that everything runs quickly and the concepts can be clearly illustrated.

## 1. Understanding the ingredients

The weight specification is created with `dkge_weights()`, which packages four
essential pieces of information that work together to define the weighting strategy:

- **Prior weights** describe locations you already trust based on external information (for example, a
  reliability map from previous studies). These weights must be *sign invariant*, meaning that flipping the
  sign of a beta map does not change the weight assigned to that voxel. Squares, absolute values, and
  variances all meet this important requirement and can be used safely as priors.
- **Adaptive statistics** are computed dynamically from the training subjects inside each
  cross-validation split. The built-in options examine *energy* (squared
  amplitude in the design metric), *precision* (inverse variance), or their
  product. Energy is simply $\|K^{1/2} B_{:\alpha}\|_2^2$, representing the amount of signal a
  voxel carries in effect space; precision is defined as $1/(\sigma^2 + \varepsilon)$ and captures measurement reliability.
- **Combination rules** specify how prior knowledge and adaptive components are blended together
  (through product, sum, or override operations). The `mix` parameter provides fine-grained control over the relative balance between these two sources of information.
- **Shrinkage** keeps the final weights well-behaved by applying several regularization steps: we winsorize extreme values and pull
  everything back towards 1 to ensure that no single voxel dominates the fit.

With `scope = "fold"`, adaptive statistics are computed from the training
subjects in each fold. This prevents the held-out subject from entering the
weight estimate. It addresses that specific leakage path, but does not by
itself guarantee calibrated p-values or external generalisation.

## 2. Building weight specifications and fitting DKGE

We begin our exploration with the default configuration, which relies purely on adaptive
statistics without any prior knowledge.

```{r make-weight}
library(dkge)

wts_adapt <- dkge_weights(
  prior   = NULL,            # no prior information
  adapt   = "kenergy_prec",  # energy × precision, a simple SNR proxy
  combine = "product",
  mix     = 0.6,             # adaptive term has more influence than the prior
  shrink  = list(alpha = 0.5, winsor = 0.99, normalize = "mean"),
  scope   = "fold"           # recomputed on the training subjects of each fold
)

print(wts_adapt)
```

Next, we create a small synthetic dataset to work with: ten subjects, four effects, and sixty voxels
per subject. While the exact numbers are unimportant for our purposes, this scale allows us to clearly demonstrate the
API and observe the effects of different weighting strategies.

```{r toy-data}
make_subject <- function(id, q = 4L, v = 60L) {
  design <- diag(q)
  colnames(design) <- paste0("eff", seq_len(q))
  beta <- matrix(rnorm(q * v), nrow = q)
  dkge_subject(beta, design = design, id = paste0("sub", id))
}

subjects <- lapply(seq_len(10L), make_subject)
K <- diag(4)

fit_uniform  <- dkge(subjects, K = K, rank = 2, w_method = "none")
fit_adaptive <- dkge(subjects, K = K, rank = 2, weights = wts_adapt)
```

The second fit internally computes voxel weights for every training fold, applying the adaptive weighting strategy we specified. We
can inspect both the global weights and examine the average weight values across subjects to understand how the algorithm has adjusted the influence of different voxels.

```{r inspect-weights}
head(fit_adaptive$voxel_weights)
dkge_diagnostics(fit_adaptive)$voxel_weights
```

You should observe modest deviations from 1.0 rather than extreme spikes in the weight values. This well-behaved pattern is
exactly what the shrinkage parameters are designed to enforce, preventing any single voxel from dominating the analysis.

## 3. Mixing priors with adaptive statistics

Now let us consider a scenario where we have an external reliability image from previous studies or analyses. We need to convert this reliability information into a sign-invariant quantity (here, squared reliability) and incorporate it as a prior. The adaptive
component remains the same as before, but now we blend it with this external knowledge.

```{r prior}
reliability_prior <- runif(60, min = 0.1, max = 1)
reliability_weight <- reliability_prior^2 + 1e-6

wts_mixed <- dkge_weights(
  prior   = reliability_weight,
  adapt   = "kenergy_prec",
  combine = "product",
  mix     = 0.4,   # stronger prior, weaker adaptive term
  shrink  = list(alpha = 0.6, winsor = 0.995)
)

fit_mixed <- dkge(subjects, K = K, rank = 2, weights = wts_mixed)
```

The squared transform is monotone on this non-negative reliability scale, so a
larger reliability value implies a larger prior weight. The adaptive term and
shrinkage can still modify the final ordering. Inspect that relationship rather
than assuming the prior had its intended effect:

```{r prior-diagnostics}
prior_check <- data.frame(
  reliability = reliability_prior,
  final_weight = fit_mixed$voxel_weights
)
stats::cor(prior_check$reliability, prior_check$final_weight,
           method = "spearman")
```

```{r prior-diagnostics-checks, include=FALSE}
stopifnot(all(reliability_weight > 0))
stopifnot(stats::cor(
  prior_check$reliability, prior_check$final_weight,
  method = "spearman"
) > 0)
```

## 4. Safety checks and post-hoc updates

### Sign-invariance check

For sign-flip inference, a fixed weighting rule should not change when every
beta is multiplied by -1. The following executable check verifies that
structural property for this example.

```{r sign-invariance}
subjects_flipped <- lapply(subjects, function(subject) {
  subject$beta <- -subject$beta
  subject
})
fit_flipped <- dkge(subjects_flipped, K = K, rank = 2, weights = wts_mixed)
max(abs(fit_flipped$voxel_weights - fit_mixed$voxel_weights))
```

```{r sign-invariance-checks, include=FALSE}
stopifnot(max(abs(
  fit_flipped$voxel_weights - fit_mixed$voxel_weights
)) < 1e-12)
```

This equality is a necessary check for the implemented sign-flip route, not a
claim of exact finite-sample calibration. A new weighting rule still needs a
simulation study matched to its folds, permutation scheme, sample size, and
null data-generating process, with Monte Carlo uncertainty reported.

If you plan to implement a custom adaptive rule, there are two critical requirements to verify:

1. The rule must depend only on even statistics (such as squares or absolute values). If your custom rule depends
   on signed means or other odd functions, you must set `perm_recompute = "always"` so that weights are recomputed
   inside each permutation flip to maintain validity.
2. The rule must be fold-specific (enforced by `scope = "fold"`), otherwise the held-out subject
   would contaminate the training statistics and introduce bias into the cross-validation procedure.

### Updating an existing fit

Sometimes you may fit a DKGE model once, then later decide to revisit the weighting strategy without
having to rerun the entire preprocessing pipeline. In such cases, you can use `dkge_update_weights()` to efficiently rebuild the fold bases
in place with your new weighting scheme.

```{r update}
fit_refit <- dkge_update_weights(fit_uniform, wts_mixed)
```

This update operation only modifies the weighting machinery itself; all subject-level projections and stored
projections remain fully compatible with the updated model.

## Summary

Adaptive weighting in DKGE has four building blocks: a prior, an adaptive
statistic, a combination rule, and shrinkage parameters. Training-fold scope
and sign-invariance close two important leakage paths. They do not prove that a
particular analysis has calibrated error rates.

The code examples throughout this vignette have demonstrated how to inspect the resulting weights, how to effectively mix prior knowledge with
adaptive rules, and how to retrofit new weighting choices onto an existing fit without recomputing everything from scratch. You should employ these
weighting controls when external reliability information or a prespecified
adaptive rule justifies unequal voxel influence. Compare the weighted fit with
an equal-weight baseline, report the rule and shrinkage settings, and calibrate
the full inferential procedure under a defensible null before making
sensitivity or error-control claims.
