---
title: Transport and Dense Rendering
description: >
  Transport subject-specific cluster values onto a shared anchor space and
  render them as a dense field.
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Transport and Dense Rendering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(dkge)
library(ggplot2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
options(
  tibble.print_min = 4,
  tibble.print_max = 4,
  digits = 3
)
set.seed(2024)
theme_set(theme_dkge())
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```


DKGE components and contrasts initially live on each subject's own clusters.
You cannot average those vectors merely because they have similar lengths:
cluster indices are not anatomical correspondences. Dense rendering makes the
spatial assumption explicit by mapping every subject through shared anchors
and, optionally, decoding the anchor field onto a voxel grid.

```text
subject cluster values -> fitted subject-to-anchor maps -> shared anchor field -> voxel interpolation
```

This page shows that object flow and the diagnostics attached to it. The toy
coordinates are deliberately geometry-only, so the final picture demonstrates
rendering mechanics rather than an anatomical result. Read
`vignette("dkge-workflow")` first if you do not yet have a component or
contrast field to render.

## What spatial information is required?

The rendering pipeline needs coordinates, so the example below carries both
beta coefficients and spatial centroids for each subject.

```{r data}
library(dkge)
S <- 4; q <- 3; P <- 20; T <- 50
betas <- replicate(S, matrix(rnorm(q * P), q, P), simplify = FALSE)
designs <- replicate(S, {
  X <- matrix(rnorm(T * q), T, q)
  qr.Q(qr(X))
}, simplify = FALSE)
centroids <- replicate(S, matrix(runif(P * 3, -40, 40), P, 3), simplify = FALSE)
subjects <- lapply(seq_len(S), function(s) dkge_subject(betas[[s]], designs[[s]], id = paste0("sub", s)))
bundle <- dkge_data(subjects)
fit <- dkge(bundle, K = diag(q), rank = 2)
fit$centroids <- centroids  # attach for transport helpers; pass explicitly in production
```

## How is the shared anchor representation built?

`dkge_build_renderer()` maps subject-specific cluster coordinates onto a
shared anchor space. It needs a target voxel grid and the transport
parameters.

```{r renderer}
# use 5k anchors sampled from grey-matter-like cube
vox_xyz <- as.matrix(expand.grid(seq(-40, 40, by = 10), seq(-40, 40, by = 10), seq(-40, 40, by = 20)))
renderer <- dkge_build_renderer(
  fit,
  centroids = centroids,
  vox_xyz = vox_xyz,
  mapper = dkge_mapper("sinkhorn", epsilon = 0.05, lambda_xyz = 1, lambda_feat = 0),
  graph_k = 10,
  decoder_k = 8,
  anchor_n = min(500L, nrow(vox_xyz)),
  anchor_method = "sample"
)
str(renderer, max.level = 1)
```

The renderer carries four components:

- `anchors`: the shared coordinates, derived from the `vox_xyz` grid, that
  serve as the common reference space.
- `graph`: an optional k-nearest-neighbor graph and Laplacian, which enables
  spatial smoothing across anchors.
- `mapper_fits`: Subject-specific fitted plans and target-conditional application
  operators. Reusing the renderer reuses these solved mappings.
- `mapper_stats`: transport diagnostics - costs, entropies, and support sizes -
  for assessing the mapping.

## Rendering Subject Values

`dkge_render_subject_values()` transports subject cluster values into the
shared anchor space, then interpolates them to the full voxel grid.

```{r render}
values_list <- lapply(fit$Btil, function(Bts) as.numeric(Bts[1, ]))
rendered <- dkge_render_subject_values(renderer, values_list, lambda = 0.2, to_vox = TRUE)
summary(rendered$details$y)
head(rendered$details$subject_stats)
```

The render returns three things:

- `rendered$anchor`: the smoothed group field on the anchor coordinates.
  This is the primary output of the transport.
- `rendered$voxel`: the interpolated voxel map, with length `nrow(vox_xyz)`.
- `rendered$details$plan_entropy_mean`: mean diffusion of the transport plans. Higher
  values mean broader anchor coverage and more diffuse mapping patterns.

The anchor values vary along a single spatial dimension, which is the easiest
way to see the field's structure.

```{r plot-anchor, fig.height=4, fig.width=5, fig.alt="Scatter plot of anchor x-coordinate versus rendered value."}
plot(renderer$anchors[, 1], rendered$anchor, pch = 20, col = "steelblue",
     xlab = "Anchor x-coordinate", ylab = "Rendered value",
     main = "Rendered anchor field")
abline(h = 0, col = "grey70", lty = 2)
```

## Voxel Map Snapshot

Slices of the voxel grid show how the anchor-based smooth field is
interpolated onto the target space.

```{r plot-vox, fig.height=4, fig.width=5, fig.alt="Scatter plot of voxel slice values for z equals zero."}
sel <- vox_xyz[, 3] == 0  # slice at z = 0
plot(vox_xyz[sel, 1], rendered$voxel[sel], pch = 16, col = "tomato",
     xlab = "x", ylab = "value", main = "Voxel slice (z=0)")
```

## What to check before reporting

Three settings change the result:

- `lambda` in `dkge_render_subject_values()` trades fidelity to the original
  data against spatial smoothness. Watch `plan_entropy_mean` to judge whether a
  setting smooths too much or too little.
- `subject_feats` and `anchor_feats` in `dkge_build_renderer()` let the
  transport use additional latent features.
- Reuse a renderer for repeated maps: its fitted plans avoid solving the same
  subject-to-anchor mappings again. Separately, DKGE keeps a process-local cache
  of converged Sinkhorn duals for byte-identical fit problems; inspect
  `mapper_fits[[s]]$stats$diagnostics$cache_hit` when diagnosing a repeated fit.

## Where to go next

- `vignette("dkge-anchors")` builds the shared anchor representation this page
  renders onto, when items rather than coordinates define the alignment.
- `vignette("dkge-performance")` compares the mappers and shows when warm starts
  and caching pay for themselves.
- `vignette("dkge-weighting")` covers the transport weighting layer, which
  changes what a rendered field means.
