---
title: 'DKGE Workflow: From Subject Betas to a Shared Map'
description: >
  Take subject beta matrices to a shared map: fit, read components, obtain
  held-out contrast fields, and transport them to one parcellation.
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{DKGE Workflow: From Subject Betas to a Shared Map}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(dkge)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
set.seed(2024)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

Suppose each subject has a GLM beta matrix in a common effect space but a
different spatial parcellation. Your practical goal is to learn a shared
effect-space representation, evaluate a planned contrast on held-out subjects,
and place those subject-specific contrast fields on one reference
parcellation.

This page follows that one path. It uses synthetic data so the code is
self-contained; replace the beta matrices, design matrices, and centroids with
your own. If `q by P_s` and “design kernel” are not yet familiar, begin with
`vignette("dkge")`.

The workflow has two coordinate systems. The fit and contrast are defined in
the shared **effect space**; the final map is defined in a shared **spatial
space**. Keeping those stages separate is the central organizing idea:

```text
validate inputs -> fit in effect space -> name a contrast -> cross-fit subject fields -> transport to one map
```

Each section below ends by naming the object you now have and the operation
that becomes valid next.

## What are the inputs?

We use a 2 by 3 factorial design with a planted condition and load signal. Six
subjects have different numbers of clusters, which is the reason transport is
needed later.

```{r make-workflow-data}
toy <- dkge_sim_toy(
  factors = list(condition = list(L = 2), load = list(L = 3)),
  active_terms = c("condition", "load"),
  S = 6, P = c(16, 18, 15, 17, 16, 19), snr = 4, seed = 2024
)
vapply(toy$B_list, dim, integer(2))
```

The output confirms a common number of effect rows and subject-specific
cluster counts. In real data, beta-row names must match design-column names.
`dkge_data()` validates that contract and preserves subject IDs.

```{r input-shapes, echo=FALSE, fig.height=3.4, fig.alt="Bar chart showing that all six subjects share five effects but have different numbers of clusters."}
input_shapes <- data.frame(
  subject = toy$subject_ids,
  clusters = vapply(toy$B_list, ncol, integer(1))
)
ggplot2::ggplot(input_shapes, ggplot2::aes(subject, clusters)) +
  ggplot2::geom_col(fill = "#b33b2e", width = 0.72) +
  ggplot2::geom_hline(yintercept = nrow(toy$K), linetype = 3, colour = "grey45") +
  ggplot2::annotate("text", x = 5.7, y = nrow(toy$K) + 0.7,
                    label = "5 shared effects", hjust = 1, size = 3.4) +
  ggplot2::labs(
    title = "The common dimension is effects, not clusters",
    x = NULL, y = "Subject-specific clusters"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

The bars differ because each subject has a different spatial partition. The
dashed line marks the five-row effect dimension that is common to every beta
matrix. DKGE can pool information along that common dimension without
pretending that cluster 7 in one subject is cluster 7 in another.

```{r bundle-workflow-data}
bundle <- dkge_data(
  toy$B_list, designs = toy$X_list, subject_ids = toy$subject_ids
)
c(subjects = length(bundle$subject_ids), effects = bundle$q)
```

```{r bundle-workflow-checks, include=FALSE}
stopifnot(length(bundle$subject_ids) == 6L)
stopifnot(bundle$q == nrow(toy$K))
stopifnot(length(unique(vapply(bundle$betas, nrow, integer(1)))) == 1L)
```

**Next operation:** fit a low-rank group basis. If your subjects lack effects
or have very unequal cell precision, stop here and read
`vignette("dkge-partial-effect-spaces")` first.

## What is the baseline fit?

Start with an identity kernel. It retains the GLM/design scaling but adds no
kernel-imposed similarity between effects.

```{r fit-identity, fig.alt="Scree plot for the two-component identity-kernel baseline fit."}
K_identity <- diag(bundle$q)
fit_identity <- dkge(bundle, K = K_identity, rank = 2, w_method = "none")
dkge_plot_scree(fit_identity)
```

This fit returns a two-column group basis `fit_identity$U`. The scree plot
describes variation represented by those components; it does not establish
their reliability or scientific importance.

```{r fit-identity-checks, include=FALSE}
identity_var <- dkge_variance_explained(fit_identity)
stopifnot(dim(fit_identity$U)[1] == bundle$q)
stopifnot(dim(fit_identity$U)[2] == 2L)
stopifnot(all(is.finite(identity_var$prop_var)))
stopifnot(sum(identity_var$prop_var) > 0)
```

**Output:** an unconstrained effect-space baseline.

**Next operation:** fit the scientifically motivated kernel and compare it
with this baseline.

## What changes when the design kernel is used?

The simulator supplies the structured kernel used to plant the signal. A real
analysis would construct this object explicitly with `design_kernel()` and
prespecify the included terms. Both fits disable subject block weighting so
the comparison isolates the kernel change.

```{r fit-structured}
fit_structured <- dkge(bundle, K = toy$K, rank = 2, w_method = "none")

comparison <- data.frame(
  component = 1:2,
  identity = dkge_variance_explained(fit_identity)$prop_var,
  structured = dkge_variance_explained(fit_structured)$prop_var
)
round(comparison, 3)
```

```{r compare-kernel-variance, echo=FALSE, fig.height=3.5, fig.alt="Grouped bar chart comparing variance proportions for identity and structured design-kernel fits."}
comparison_long <- rbind(
  data.frame(component = comparison$component, kernel = "Identity", proportion = comparison$identity),
  data.frame(component = comparison$component, kernel = "Structured", proportion = comparison$structured)
)
ggplot2::ggplot(
  comparison_long,
  ggplot2::aes(factor(component), proportion, fill = kernel)
) +
  ggplot2::geom_col(position = "dodge", width = 0.7) +
  ggplot2::scale_y_continuous(labels = function(x) paste0(round(100 * x), "%")) +
  ggplot2::labs(
    title = "Where the two fits allocate retained variation",
    x = "Component", y = "Proportion", fill = "Kernel"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

Compare the retained subspaces as well as the scree values. Because the bases
use a kernel metric, a raw `crossprod()` is not a valid rotation comparison.

```{r compare-subspaces}
angles <- dkge_principal_angles_K(
  fit_identity$U, fit_structured$U, fit_structured$K
)
round(angles * 180 / pi, 1)
```

Here component 1 accounts for `r round(100 * comparison$identity[1], 1)`% of
the retained variation under the identity kernel and
`r round(100 * comparison$structured[1], 1)`% under the structured kernel. Both
principal angles round to `r paste(round(angles * 180 / pi, 1), collapse = " and ")`
degrees: the kernel changes how variation is allocated within this seeded
two-dimensional subspace, but not the subspace itself at displayed precision.
This is a sensitivity comparison, not a test of which kernel is true.

```{r compare-subspaces-checks, include=FALSE}
stopifnot(length(angles) == 2L, all(is.finite(angles)))
stopifnot(all(angles >= 0), all(angles <= pi / 2 + 1e-8))
stopifnot(all(is.finite(unlist(comparison[-1]))))
```

**Output:** a structured fit plus an explicit record of how it differs from
the identity baseline.

**Next operation:** inspect effect saliences, then specify the contrast you
actually want to estimate.

## Which effect directions define the components?

```{r show-effect-loadings, fig.alt="Heatmap of effect-space saliences for the first two components under the structured design kernel."}
dkge_plot_effect_loadings(fit_structured, comps = 1:2)
```

The heatmap displays \(K U\): effect-space saliences, not voxel coefficients.
Look for coherent relative patterns within a component. Do not compare a
component's arbitrary sign across independent fits without alignment.

## How do you obtain held-out contrast fields?

Build the contrast in fitted effect order. Here the simulator identifies the
coordinate for the planted condition term.

```{r fit-loso-contrast}
condition_contrast <- numeric(bundle$q)
condition_contrast[toy$active_cols$condition] <- 1

condition_loso <- dkge_contrast(
  fit_structured,
  list(condition = condition_contrast),
  method = "loso"
)
condition_loso
```

The output is a named list of six subject-specific cluster vectors. LOSO
cross-fitting keeps each subject out of the basis used to score that subject.
It does not make a randomized causal contrast out of an observational design,
nor does it supply group uncertainty by itself.

```{r fit-loso-checks, include=FALSE}
loso_values <- condition_loso$values$condition
stopifnot(length(loso_values) == length(bundle$subject_ids))
stopifnot(identical(unname(vapply(loso_values, length, integer(1))),
                    unname(vapply(toy$B_list, ncol, integer(1)))))
stopifnot(all(vapply(loso_values, function(x) all(is.finite(x)), logical(1))))
```

**Output:** one cross-fitted contrast field per subject, still indexed by that
subject's clusters.

**Next operation:** transport those fields before stacking or performing
feature-level group inference.

## How do subject fields reach one parcellation?

Transport needs real spatial features, normally cluster centroids in a common
coordinate system. The synthetic coordinates below exist only to demonstrate
the interface; their resulting map has no anatomical interpretation.

```{r make-centroids}
centroids <- lapply(seq_along(toy$B_list), function(s) {
  p <- ncol(toy$B_list[[s]])
  cbind(x = seq_len(p), y = rep(s / 100, p), z = rep(0, p))
})
```

Use the ridge mapper for a dependency-light example and select subject 1 as the
reference parcellation:

```{r transport-contrast}
transported <- dkge_transport_contrasts_to_medoid(
  fit_structured,
  condition_loso,
  medoid = 1,
  centroids = centroids,
  method = "ridge"
)
dim(transported$condition$subj_values)
```

The returned matrix has subjects in rows and reference clusters in columns;
`transported$condition$value` is its across-subject median on the medoid
parcellation. The attached `attr(transported, "cache")` can be reused for
other contrasts or resampling.

```{r transported-field, echo=FALSE, fig.height=4.2, fig.alt="Six faint subject-level transported contrast profiles with their median highlighted across the reference clusters."}
transported_matrix <- transported$condition$subj_values
transport_long <- data.frame(
  subject = rep(toy$subject_ids, each = ncol(transported_matrix)),
  cluster = rep(seq_len(ncol(transported_matrix)), times = nrow(transported_matrix)),
  value = as.vector(t(transported_matrix))
)
transport_median <- data.frame(
  cluster = seq_len(ncol(transported_matrix)),
  value = apply(transported_matrix, 2, stats::median)
)
ggplot2::ggplot(transport_long, ggplot2::aes(cluster, value, group = subject)) +
  ggplot2::geom_hline(yintercept = 0, colour = "grey75") +
  ggplot2::geom_line(colour = "#7f8c8d", alpha = 0.42, linewidth = 0.55) +
  ggplot2::geom_line(
    data = transport_median,
    ggplot2::aes(cluster, value),
    inherit.aes = FALSE, colour = "#b33b2e", linewidth = 1.2
  ) +
  ggplot2::labs(
    title = "Cross-fitted subject fields after transport",
    subtitle = "Grey: subjects; red: coordinate-wise median on the reference parcellation",
    x = "Reference cluster", y = "Condition contrast"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

This is the first point at which a feature-by-feature group summary is
well-defined. The grey profiles preserve the between-subject variation needed
for inference; the red line is a descriptive center, not a substitute for
subject-level uncertainty.

```{r transport-checks, include=FALSE}
stopifnot(nrow(transported_matrix) == length(toy$B_list))
stopifnot(ncol(transported_matrix) == ncol(toy$B_list[[1]]))
stopifnot(all(is.finite(transported_matrix)))
stopifnot(stats::sd(as.vector(transported_matrix)) > 0)
```

Transport quality depends on the features, masses, mapper, and reference
choice. Inspect mapper diagnostics and repeat defensible sensitivity analyses;
a successful matrix multiplication is not evidence that parcels are
biologically homologous.

## What should you save and report?

For a reproducible analysis, retain:

- effect names and the design matrices used to produce each beta matrix;
- the identity and structured kernels, their construction parameters, and the
  chosen rank;
- the identity-versus-structured comparison;
- cross-fitting folds or LOSO specification;
- cluster coordinates, masses, mapper settings, medoid choice, and transport
  diagnostics; and
- the inference unit and multiplicity procedure for every reported claim.

Continue with `vignette("dkge-concepts")` for the fitted moment and inference
levels, `vignette("dkge-contrasts-inference")` for uncertainty, and
`vignette("dkge-weighting")` for the distinct roles of spatial, effect,
subject, and transport weights.
