---
title: "Classification with DKGE"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: false
    df_print: paged

vignette: >
  %\VignetteIndexEntry{Classification with DKGE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r library, message=FALSE}
library(dkge)
set.seed(12)
```

The DKGE classification helpers allow you to reuse the cross-fitted group basis to decode
subject-level contrasts with statistical rigor. This vignette demonstrates a complete
workflow: data preparation → model fit → target specification → cross-validated decoding
→ statistical inference.

The approach is fully cross-validated: by default `dkge_classify()` uses
leave-one-subject-out (LOSO) cross-fitting, re-estimating the group basis without the
held-out subject at each fold. This ensures that the reported accuracies are unbiased
estimates of generalisation performance.

## Setup

We define two experimental factors — condition (A vs B) and time (4 time points) — and
construct a design kernel that encodes their structure:

```{r design-kernel}
factors <- list(cond = list(L = 2), time = list(L = 4))
kern    <- design_kernel(factors, basis = "effect")
q       <- nrow(kern$K)   # number of effect-coded columns
q
```

We simulate 10 subjects, each with `q` design effects and 60 voxels. A detectable signal
distinguishing the two conditions is embedded in the condition contrast:

```{r simulate}
n_subjects <- 10L
v <- 60L

make_subject <- function(id) {
  # design: identity matrix over effects (one trial per effect)
  design <- diag(q)
  colnames(design) <- rownames(kern$K)

  # signal: condition contrast (first effect column) drives voxels 1-10
  signal <- matrix(0, nrow = q, ncol = v)
  signal[1, 1:10] <- 0.5    # positive signal for condition contrast

  beta <- signal + matrix(rnorm(q * v, sd = 1.0), nrow = q)
  dkge_subject(beta, design = design, id = paste0("sub", id))
}

subjects <- lapply(seq_len(n_subjects), make_subject)
```

Fit the DKGE model using the design kernel. `design_kernel()` returns a list with `$K`
(the q×q kernel matrix) and `$info` (factor metadata used by `dkge_targets()`):

```{r fit}
fit <- dkge(subjects, K = kern, rank = 2)
fit
```

## Classification targets

`dkge_targets()` maps design formula terms to classification contrasts using the factor
metadata stored in the kernel. Each term becomes a binary (or multi-class) decoding
problem over the group embedding.

```{r targets}
targets <- dkge_targets(fit, ~ cond + time)
length(targets)                      # one target per formula term
targets[[1]]$name                    # "cond"
nrow(targets[[1]]$weight_matrix)     # 2 classes (A vs B)
```

The weight matrix transforms subject-level beta vectors into class-specific pattern
vectors for the downstream classifier.

## Cross-validated decoding

```{r classify}
cls <- dkge_classify(
  fit,
  targets = targets,
  method  = "lda",     # "lda" (default) or "logit"
  n_perm  = 199,       # permutation test for significance
  seed    = 99
)
print(cls)
```

The result is a `dkge_classification` object. Each entry of `$results` corresponds to
one target:

```{r results}
res <- cls$results[["cond"]]
res$mode       # decoding mode: "cell" for within-subject targets
res$metrics    # cross-validated accuracy and log-loss
```

Convert to a tidy data frame for plotting or downstream analysis:

```{r tidy}
df <- as.data.frame(cls)
head(df)
```

### Permutation p-values

When `n_perm > 0`, each target gets an empirical p-value from the sign-flip max-T
permutation distribution:

```{r pvalues}
res$p_values    # named numeric vector per metric (NULL when n_perm = 0)
```

### Subject-level predictions

Individual predictions and fold assignments live in `$row_data`:

```{r predictions}
head(res$row_data[, c("subject_label", "class_label", "fold")])
```

Full predicted probabilities are in `res$probabilities` (rows = subject × class
combinations, columns = class labels).

### Per-fold diagnostics

Confusion matrices and class counts per fold are in `res$diagnostics$folds`:

```{r diag}
diag_fold1 <- res$diagnostics$folds[[1]]
diag_fold1$confusion           # confusion matrix for fold 1
diag_fold1$class_counts_test   # observed class counts in test set
```

## Two modes: `cell` vs `cell_cross`

The `mode` argument controls how the group basis is applied at test time:

| Mode | Basis used | Use case |
|------|-----------|----------|
| `"cell"` | Global `fit$U` | Fast; mild basis-step leakage — `fit$U` saw all subjects |
| `"cell_cross"` | Fold-specific LOSO `U_fold` | Fully cross-validated, more conservative |
| `"delta"` | Global `fit$U` + subject labels | Subject-level binary test; requires `y` argument |

`"auto"` (default) selects `"cell"` for within-subject targets and `"delta"` for
between-subject targets.

```{r cell-cross, eval=FALSE}
# Fully cross-validated basis — recommended for publication-quality results
cls_strict <- dkge_classify(fit, targets = targets, mode = "cell_cross", n_perm = 0)
```

## Multiple betas per condition

When you have multiple beta estimates per condition (e.g., separate scanner runs), stack
the run-specific effects as extra rows in the design matrix and supply a weight matrix
that combines them into a single class pattern.

```{r multi-beta}
# 4 effects: A_run1, A_run2, B_run1, B_run2
make_subject_multi <- function(id) {
  design <- diag(4)
  colnames(design) <- c("A_run1", "A_run2", "B_run1", "B_run2")

  signal <- matrix(0, nrow = 4, ncol = v)
  signal[1:2, 1:10] <-  0.5   # both A runs share signal
  signal[3:4, 1:10] <- -0.5   # both B runs share the opposite signal

  beta <- signal + matrix(rnorm(4 * v, sd = 1.0), nrow = 4)
  dkge_subject(beta, design = design, id = paste0("sub", id))
}

subjects_multi <- lapply(seq_len(n_subjects), make_subject_multi)
fit_multi      <- dkge(subjects_multi, K = diag(4), rank = 2)

# Weight matrix: rows = classes, columns = effects.
# Each class averages its two runs.
W_runs <- matrix(0, nrow = 2, ncol = 4)
W_runs[1, 1:2] <- 0.5    # class A = mean(A_run1, A_run2)
W_runs[2, 3:4] <- 0.5    # class B = mean(B_run1, B_run2)
rownames(W_runs) <- c("A", "B")
```

A plain matrix is accepted directly as `targets` — DKGE wraps it automatically:

```{r multi-classify}
cls_multi <- dkge_classify(fit_multi,
                           targets = W_runs,   # plain matrix dispatch
                           n_perm  = 99,
                           seed    = 101)

res_multi <- cls_multi$results[[1]]
res_multi$metrics
as.data.frame(cls_multi)
```

The cross-validation remains unbiased: run averaging happens inside the weight matrix,
so each LOSO fold observes exactly one pattern per condition.

## Hyperdesign inputs and fold bridges

Labs using `multidesign::hyperdesign()` can supply design kernels and fold assignments
through coercion S3 generics. Any object implementing `as_dkge_kernel()` and
`as_dkge_folds()` flows through the existing pipeline without altering the core solvers.

```{r hyperdesign, eval=FALSE}
# Pseudocode — requires multidesign package and a user-defined hyperdesign object
library(multidesign)

hd     <- make_demo_hyperdesign()                # user-supplied helper
Kobj   <- as_dkge_kernel(hd, basis = "effect")   # list(K = ..., info = ...)

fit_hd <- dkge(
  betas       = dkge_data_from_hd(hd),
  K           = Kobj,
  keep_inputs = TRUE
)

folds  <- as_dkge_folds(fold_over(hd, over = "subject", k = 5, seed = 1), fit_hd)

res_hd <- dkge_contrast(fit_hd,
                        contrasts = c(1, -1, 0, 0),
                        method    = "kfold",
                        folds     = folds)
```

Matrices, lists with a `$K` element, and existing `dkge_folds` objects continue to work
as before — only packages with richer design structures need to implement these generics.

## Practical tips

**Target specification.** Define the full factorial structure with formula notation
(e.g., `~ cond + time + cond:time`) and reuse the same target definitions across both
classification and contrast analyses for methodological consistency.

**Backend choice.** LDA (`method = "lda"`) is fast and usually sufficient. Use
`method = "logit"` with `class_weights = "balanced"` when experimental conditions have
unequal numbers of trials.

**Permutation testing.** Increase `n_perm` for precise p-values near significance
thresholds. Set `n_perm = 0` during exploratory analysis to skip the permutation loop.

**Spatial alignment.** When subjects have different voxel grids, apply transport to a
common medoid parcellation via `dkge_transport_contrasts_to_medoid()` before
classification.

**Input caching.** Keep `keep_inputs = TRUE` (default) to enable `dkge_update_weights()`
without re-running the full fit.

## Next steps

- **Adaptive weighting** (`vignette("dkge-adaptive-weighting")`): Emphasize spatially
  reliable voxels before computing the group embedding.
- **`dkge_pipeline()`**: Orchestrate fit → contrast → transport → inference →
  classification in a single call.
- **Diagnostics**: Inspect `fit$evals` (variance explained) and per-fold confusion
  matrices to identify underperforming subjects or poorly separated conditions.
