---
title: "Weighting Strategies in DKGE"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: false
    df_print: paged

vignette: >
  %\VignetteIndexEntry{Weighting Strategies in DKGE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(dkge)
library(ggplot2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
options(
  tibble.print_min = 4,
  tibble.print_max = 4,
  digits = 3
)
set.seed(2024)
theme_set(theme_dkge())
```


Design-Kernel Group Embedding (DKGE) provides several sophisticated weighting mechanisms that allow you to emphasize reliable clusters, incorporate spatial smoothness constraints, or balance contributions from subjects in heterogeneous cohorts. This vignette systematically catalogues the available weighting options and demonstrates how they interact throughout the various stages of fitting, projection, and transport operations.

### Adaptive voxel weighting (preview)

In addition to the subjects × clusters weighting controls showcased throughout this vignette, DKGE can also rescale individual voxels adaptively before fold bases are constructed. This adaptive weighting approach enhances informative brain regions by intelligently combining sign-invariant priors (such as reliability maps) with training-only statistics including design-kernel energy or precision measures. This provides a fold-safe method to sharpen statistical contrasts without discarding potentially valuable data. The companion [Adaptive Voxel Weighting vignette](dkge-adaptive-weighting.html) provides detailed coverage of the `dkge_weights()` specification, null-safety validation checks, and post-hoc weight updates using `dkge_update_weights()`.

## Weighting Landscape

DKGE operates with three distinct yet complementary weighting layers that work together to provide fine-grained control over the analysis:

The first layer consists of **spatial weights (`omega`)**, which operate within each individual subject block before the compressed covariance is accumulated across subjects. You can supply either vectors (which implement diagonal weights) or positive semi-definite matrices (which capture full covariances) when constructing subjects or calling the main `dkge()` function.

The second layer involves **subject weights (`w_method`, `w_tau`)**, which rescale entire subject blocks based on their energy measured in the K-metric. This layer offers Multiple Factor Analysis style scaling (`"mfa_sigma1"`) and Frobenius-based energy scaling (`"energy"`) schemes, alongside options for manual weight overrides.

The third layer comprises **transport weights (`sizes`)**, which shape the optimal-transport alignment process when producing consensus maps. This allows you to respect important spatial characteristics such as cluster surface areas or voxel densities during the mapping procedure.

The following sections systematically walk through each weighting layer using a carefully constructed simulated dataset.

## Example Dataset

To illustrate the behavior of different weighting strategies, we generate a synthetic dataset with three subjects that exhibit varying cluster counts, sizes, and reliability characteristics. This heterogeneity makes the effects of different weighting approaches readily apparent during inspection.

```{r example-data, message=FALSE}
library(dkge)
library(purrr)
set.seed(42)

S <- 3
q <- 4
P <- c(6, 5, 7)
T <- 40

betas <- map(P, ~ matrix(rnorm(q * .x), q, .x))
designs <- map(seq_len(S), function(s) {
  X <- matrix(rnorm(T * q), T, q)
  qr.Q(qr(X))
})

centroids <- map(P, ~ matrix(runif(.x * 3, min = -20, max = 20), .x, 3))
cluster_sizes <- map(P, ~ sample(10:40, .x, replace = TRUE))
reliability_scores <- map(P, ~ runif(.x, min = 0.4, max = 1))
```

In this simulation, the `cluster_sizes` vectors serve as proxies for surface-area weights that would reflect the actual spatial extent of brain parcels, while the `reliability_scores` could represent various quality control measures such as motion artifacts or bootstrap stability assessments commonly used in neuroimaging analyses.

## Baseline Fit (No Spatial Weights)

```{r baseline-fit}
data_equal <- dkge_data(betas, designs = designs)
k_identity <- diag(data_equal$q)

fit_equal <- dkge(data_equal, kernel = k_identity, rank = 2, w_method = "none")
fit_equal$weights
```

In this baseline configuration where all weights are equal, every subject contributes uniformly to the construction of the shared basis. Consequently, cluster-level variability patterns are driven purely by the structure encoded in the design kernel, without any additional bias from differential weighting schemes.

## Size-Weighted Blocks

When you supply per-cluster mass values, this approach emphasizes the contribution of larger parcels while maintaining a diagonal information structure that preserves the independence assumption between clusters.

```{r size-weighting}
omega_size <- cluster_sizes

fit_size <- dkge(data_equal,
                 kernel = k_identity,
                 omega = omega_size,
                 rank = 2,
                 w_method = "none")
fit_size$weights
```

Since the subject blocks now differ in their total weighted energy due to the size-based weighting, each block's relative influence on the underlying eigenproblem shifts accordingly. By inspecting the leading component, we can observe how the loadings are systematically rebalanced to favor larger parcels:

```{r size-loadings}
size_loadings <- lapply(fit_size$Btil, function(Bts) t(Bts) %*% fit_size$K %*% fit_size$U)
map(size_loadings, ~ round(.x[, 1, drop = FALSE], 3))
```

## Reliability Weighting

Reliability scores can be seamlessly incorporated into the `omega` weighting scheme by scaling each cluster according to its expected signal quality characteristics. When you want to combine both size and reliability considerations, this can be accomplished through a straightforward Hadamard (element-wise) product of the respective weight vectors.

```{r reliability-weighting}
omega_reliability <- map2(cluster_sizes, reliability_scores, `*`)

fit_reliability <- dkge(data_equal,
                        kernel = k_identity,
                        omega = omega_reliability,
                        rank = 2,
                        w_method = "none")
fit_reliability$weights
```

By comparing `fit_reliability$weights` with `fit_size$weights`, we can observe that subjects containing systematically noisier clusters receive reduced leverage in the analysis, even before any additional subject-level scaling is applied.

In real-world applications, reliability vectors can be derived from several methodologically sound sources. First, you can use per-cluster GLM uncertainty measures, such as inverse residual variance or `1 / se^2` values obtained from first-level statistical fits. Second, split-half or test-retest variance estimates provide another approach by down-weighting parcels that show poor temporal stability. Third, motion and quality control summaries can identify and flag clusters that were acquired under conditions of excessive artifact contamination. Fourth, bootstrap or jackknife stability scores can be converted into positive weight values that reflect resampling-based reliability. Finally, you can create combinations of the above approaches, often multiplying them by parcel sizes and rescaling the results so that the subject-level mean weight equals one.

An important technical consideration is to always maintain positive weight values and add a small ridge term (such as `+ 1e-6`) when your weight construction procedure might potentially produce zero values.

## Spatial Covariance (Smoothness) Weighting

When spatial coordinate information for clusters is available, you can replace simple diagonal weights with more sophisticated smooth covariance matrices that capture spatial relationships. The example below demonstrates how to construct Gaussian affinity matrices that explicitly reward neighboring parcels for exhibiting coordinated activation patterns.

```{r smoothness-weighting}
gaussian_cov <- function(coords, bandwidth = 15) {
  D <- as.matrix(dist(coords))
  K <- exp(-(D / bandwidth)^2)
  (K + t(K)) / 2 + diag(1e-6, nrow(K))
}

omega_smooth <- lapply(centroids, gaussian_cov)

fit_smooth <- dkge(data_equal,
                   kernel = k_identity,
                   omega = omega_smooth,
                   rank = 2,
                   w_method = "none")
fit_smooth$weights
```

These matrix-based weights function conceptually like whitening transforms in that they first apply spatial smoothing to the beta coefficients within cluster neighborhoods, and then propagate this blended energy information into the shared covariance structure. This approach proves particularly useful when working with coarse parcellations where you want to down-weight artificial sharp transitions between adjacent cluster boundaries.

## Subject-Level Scaling

The `w_method` argument provides control over block-level reweighting that occurs after the spatial weights have already been applied to individual clusters. The default Multiple Factor Analysis scaling approach systematically shrinks the influence of dominant subjects by applying weights based on the inverse squared leading singular value of their contribution.

```{r subject-weighting}
fit_mfa <- dkge(data_equal,
                kernel = k_identity,
                omega = omega_reliability,
                rank = 2,
                w_method = "mfa_sigma1",
                w_tau = 0.25)

fit_energy <- dkge(data_equal,
                   kernel = k_identity,
                   omega = omega_reliability,
                   rank = 2,
                   w_method = "energy",
                   w_tau = 0)

rbind(mfa = fit_mfa$weights,
      energy = fit_energy$weights,
      none = fit_size$weights)
```

The `w_tau` parameter provides a mechanism to shrink extreme weight values toward more equal contributions across subjects, which becomes particularly valuable when working with imbalanced cohorts or datasets that contain potential outlier subjects.

## Weighting During Transport

The transport utilities within DKGE accept their own dedicated `sizes` argument that conceptually mirrors the role of the `omega` parameter but operates at a different stage—specifically *after* the components have been estimated. By providing consistent size information during transport, you can ensure that the reference medoid representation remains faithful to the actual surface area characteristics of the brain parcels.

```{r transport-weighting}
fit_mfa$centroids <- centroids

comp_equal <- dkge_component_stats(
  fit_mfa,
  mapper = list(strategy = "sinkhorn", epsilon = 0.05),
  centroids = centroids,
  inference = list(type = "parametric"),
  medoid = 1L
)

comp_weighted <- dkge_component_stats(
  fit_mfa,
  mapper = list(strategy = "sinkhorn", epsilon = 0.05),
  centroids = centroids,
  sizes = cluster_sizes,
  inference = list(type = "parametric"),
  medoid = 1L
)

data.frame(equal = head(comp_equal$summary$stat),
           size_weighted = head(comp_weighted$summary$stat))
```

Even when working with identical component scores, the application of reweighting during the transport phase systematically shifts the medoid summary statistics toward larger parcels, demonstrating the distinct influence of transport-stage weighting.

## Practical Tips

Several practical considerations can help you effectively implement weighting strategies in your DKGE analyses. First, construct your subjects using `dkge_subject()` to ensure that weight information stays properly attached to the data when passing through complex pipelines or streaming data loaders. Second, for voxelwise analyses, remember to feed cluster size information to `dkge_transport_to_voxels()` through the `sizes` parameter in order to preserve volumetric weighting throughout the mapping process.

When combining multiple weighting concepts, maintain positive vector values and consider adding a mild ridge term (such as `+ 1e-6`) to covariance matrices to prevent numerical instabilities. Additionally, make it a practice to revisit the `fit$weights`, singular values, and eigenspectra whenever you modify weighting assumptions, as these provide quick diagnostic information about the balance of contributions across subjects.

Collectively, these weighting tools allow you to tailor DKGE analyses to accommodate study-specific quality metrics or incorporate prior spatial beliefs, all without requiring modifications to the core fitting pipeline.
