---
title: "DKGE Architecture Guide"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: false
    df_print: paged

vignette: >
  %\VignetteIndexEntry{DKGE Architecture Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.show = "hide",
  message = FALSE,
  warning = FALSE
)
library(dkge)
set.seed(2026)
```

This is a contributor-oriented description of the current implementation. The
central design constraint is simple: subject data may be wide in features, but
group estimation must stay in q-dimensional effect space. DKGE forms q-by-q
moments and never a feature-by-feature covariance matrix.

## Where does each responsibility live?

| Domain | Key files | Current responsibility |
|---|---|---|
| Subjects and alignment | `dkge-data.R`, `dkge-align-data.R`, `dkge-effect-grid.R` | Validate local effects, align them to a named union, and retain observation masks. |
| Trialwise reduction | `dkge-trial-subject.R` | Reduce trial-by-feature maps to effect estimates and uncertainty sufficient statistics. |
| Moment construction | `dkge-moments.R`, `dkge-effect-weights.R` | Apply coverage, reliability, analytic/split-half debiasing, and pair normalization in raw effect space. |
| Fit stages | `dkge-fit.R`, `dkge-fit-core.R` | Prepare inputs, accumulate q-space moments, solve the eigensystem, and assemble the fit. |
| Cross-fitting | `dkge-folds.R`, `dkge-loso.R`, `dkge-kfold.R`, `dkge-bootstrap.R` | Re-pool stored subject moments without leaking held-out subjects. |
| Between-subject layer | `dkge-subject-model.R`, `dkge-between-rrr.R`, `dkge-between-infer.R` | Fit reduced-rank subject models and evaluate Freedman-Lane or residual-rotation nulls. |
| Transport and rendering | `dkge-transport*.R`, `dkge-render*.R`, C++ Sinkhorn/distance helpers | Move subject fields to common anchors or voxels without changing the q-space estimator. |
| Classification | `dkge-classify*.R`, `dkge-latent*.R`, `dkge-info-maps.R` | Cross-fit latent classifiers and construct decoder, Haufe, and LOCO maps. |

Files under `future/` are design drafts and are not loaded or exported. A
contributor must move and reconcile a draft before treating it as package code.

## What is the fit data flow?

The public `dkge_fit()` wrapper orchestrates four implemented internal stages:

```
dkge_subject() / dkge_trial_subject()
                  |
                  v
             dkge_data()
        align effects and masks
                  |
                  v
       .dkge_fit_prepare()
    kernel roots, ruler, weights
                  |
                  v
      .dkge_fit_accumulate()
  raw q-space moments and coverage
                  |
                  v
        .dkge_fit_solve()
  symmetric q-space eigenproblem / JD
                  |
                  v
       .dkge_fit_assemble()
 basis, projections, diagnostics, cache
```

The stages are internal so their signatures may evolve, but their mathematical
contracts are regression-tested and reflected in the public fit object.

## What enters the pooled moment?

For subject \(s\), the ordinary raw effect moment is

\[
M_s = B_s \Omega_s B_s^\top,
\]

where \(B_s\) is q-by-\(P_s\) and \(\Omega_s\) is an optional spatial weight.
Partial observation masks zero unobserved rows and columns before any design or
kernel transform.

Trialwise inputs add two alternatives:

- **Analytic debiasing:**
  \(M_s = B_s\Omega_sB_s^\top - t_s\Lambda_s\), where
  \(\Lambda_s=(X_s^\top X_s)^{-1}\) and \(t_s\) is the spatially weighted
  residual-variance trace.
- **Split-half debiasing:**
  \(M_s = (B_s^{(1)}\Omega_sB_s^{(2)\top} +
  B_s^{(2)}\Omega_sB_s^{(1)\top})/2\).

Effect precision \(p_{sc}\) contributes pair reliability
\(h_{scc'}=\sqrt{p_{sc}p_{sc'}}\). The pool computes a precision-weighted mean
per observed pair and, except under `missingness = "rescale"`, restores the
cohort mass so unit precision exactly recovers the legacy sum. It records:

- `pair_counts`: observed subject mass for each effect pair;
- `pair_weight`: summed reliability mass;
- `pair_ess`: Kish effective sample size under reliability weights.

Only after this raw-effect pool is resolved does DKGE transform it:

\[
\widehat C = K^{1/2}R^\top M R K^{1/2}.
\]

This order is important. Coverage and precision describe the data; \(R\) and
\(K\) describe the metric. Moving masks after the congruence would change the
estimand because a coupling kernel has already mixed effect coordinates.

## Which eigensolver is used?

The pooled solver symmetrizes the q-by-q matrix and uses a dense symmetric
eigendecomposition. That is intentional at the package's target q (usually
10-100): it is deterministic, exposes negative eigenvalues introduced by
debiasing, and avoids a feature-space solver. Positive leading eigenpairs are
mapped back to a K-orthonormal basis.

`solver = "jd"` is a separate joint-diagonalization path. It currently rejects
effect weighting, missingness normalization, and debiasing instead of silently
solving a different objective. Adding those features to JD requires a derived
objective and dedicated validation, not argument forwarding.

## How can you inspect the public contract?

The following small fit shows that all group-level quadratic objects remain
q-by-q.

```{r architecture-fit}
toy <- dkge_sim_toy(
  factors = list(A = list(L = 2), B = list(L = 3)),
  active_terms = c("A", "B"), S = 4, P = 12, snr = 4, seed = 21
)
data <- dkge_data(toy$B_list, toy$X_list,
                  subject_ids = toy$subject_ids)
fit <- dkge_fit(data, K = toy$K, rank = 2, w_method = "none")

data.frame(
  object = c("pooled raw-effect moment", "transformed covariance",
             "pair effective sample size", "group basis"),
  rows = c(nrow(fit$effect_moment), nrow(fit$Chat),
           nrow(fit$pair_ess), nrow(fit$U)),
  columns = c(ncol(fit$effect_moment), ncol(fit$Chat),
              ncol(fit$pair_ess), ncol(fit$U))
)
```

```{r architecture-checks, include = FALSE}
stopifnot(identical(dim(fit$effect_moment), c(data$q, data$q)))
stopifnot(identical(dim(fit$Chat), c(data$q, data$q)))
stopifnot(identical(dim(fit$pair_ess), c(data$q, data$q)))
stopifnot(nrow(fit$U) == data$q, ncol(fit$U) == fit$rank)
stopifnot(all(is.finite(fit$Chat)))
```

The fit also stores per-subject effect moments for fold refits. It intentionally
does not retain redundant full beta/design/weight copies in `fit$subjects`.
`effect_moments_raw` differs from `effect_moments` only for analytic debiasing;
`noise_moments` is `NULL` otherwise.

## How do resampling paths reuse the fit?

LOSO, k-fold, bootstrap, and multiplier paths re-pool the stored subject
q-by-q moments. Each fold recomputes data-derived pair weights, coverage, and
missingness decisions from its training subjects. This avoids both feature-
space work and leakage from a full-sample reliability pool.

The between-subject layer has a different shape: its input is an n-subject by
P-feature target. Freedman-Lane and residual rotation both reduce each
replicate to design-by-feature cross-products before solving the reduced-rank
problem. Residual rotation fixes the reduced-design coordinates and rotates
only their complement; it forms a residual-dimension-square orthogonal matrix,
not a P-by-P covariance.

## What invariants should contributors preserve?

1. **Never form \(P_s\times P_s\).** New quadratic work belongs in q-space or
   in a documented subject residual space.
2. **Symmetrize every quadratic form.** Debiased moments may be indefinite;
   retain and report negative spectral mass rather than hiding it with an SVD.
3. **Apply coverage before \(R/K\).** Observation masks and effect precision
   have meaning only in the named raw effect coordinate.
4. **Keep label alignment explicit.** Effect matrices, kernels, uncertainty
   matrices, and counts must share a checked ordering.
5. **Refit data-derived quantities inside resampling folds.** Cached sufficient
   statistics are welcome; cached full-sample decisions are leakage.
6. **Fail closed on unsupported combinations.** The JD and residual-rotation
   guards are deliberate scientific boundaries.
7. **Protect the compatibility path.** Full-coverage, unit-precision,
   `debias = "none"` fits must reproduce the historical pooled estimator to
   numerical tolerance.

For the user workflow, see `vignette("dkge-workflow")`. For partial coverage and
trialwise sufficient statistics, see `vignette("dkge-partial-effect-spaces")`.
For between-subject rotation and its frozen calibration, see
`vignette("dkge-between-subjects")`.
