---
title: Plotting DKGE Fits
author: DKGE Team
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Plotting DKGE Fits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(dkge)
library(ggplot2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
options(
  tibble.print_min = 4,
  tibble.print_max = 4,
  digits = 3
)
set.seed(2024)
theme_set(theme_dkge())
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```


This vignette shows how to inspect one DKGE fit without mistaking a polished
figure for statistical evidence. We will plot its eigenspectrum, effect
saliences, subject contributions, and subspace sensitivity. Each panel answers
a descriptive question; none supplies a p-value or validates the scientific
model on its own.

## Prerequisites

```{r libraries, message=FALSE}
library(dkge)
library(ggplot2)
library(patchwork)
```

The information-map plotting functions provide enhanced labeling capabilities by optionally utilizing the `ggrepel` package to annotate top anchors with non-overlapping text placement. When `ggrepel` is not available in your environment, the code gracefully falls back to standard base text labels to ensure functionality is maintained.

## Simulate a toy dataset

We begin by generating a small synthetic dataset manually. Each subject has a diagonal design matrix (one column per effect), and we add modest signal plus noise so the latent components are recoverable but not trivial. Using this construction keeps the vignette self-contained and avoids dependencies on specialised simulators while still providing deterministic ground truth for the plots.

```{r simulate}
q <- 4L   # number of effects
P <- 30L  # clusters (voxels) per subject
S <- 8L   # subjects

make_subject <- function(id) {
  design <- diag(q)
  colnames(design) <- paste0('eff', seq_len(q))
  signal <- matrix(rnorm(q * P, sd = 0.4), nrow = q)
  noise <- matrix(rnorm(q * P, sd = 0.2), nrow = q)
  dkge_subject(signal + noise, design = design, id = paste0('sub', id))
}

subjects <- lapply(seq_len(S), make_subject)
K <- diag(q)
fit <- dkge(subjects, K = K, rank = 3, w_method = 'mfa_sigma1')
```

## Auxiliary objects

To explain the stability plot's input contract, we create perturbed versions
of the fitted rank-three basis and re-orthonormalise them in the \(K\)-metric.
These are controlled sensitivity perturbations, not cross-validation folds. In
an analysis, supply bases actually refitted inside the relevant folds.

```{r loso, results='hide'}
fit_U <- fit[['U']]
bases <- replicate(4, {
  perturb <- matrix(rnorm(length(fit_U), sd = 0.02), nrow = nrow(fit_U))
  dkge_k_orthonormalize(fit_U + perturb, fit[['K']])
}, simplify = FALSE)
base_labels <- paste0('base', seq_along(bases))
```

The optional information-map panels require real outputs from
`dkge_info_map_haufe()` or `dkge_info_map_loco()`. We omit them here because
random vectors would demonstrate only the plotting signature while looking
like scientific evidence. See `vignette("dkge-classification")` for the
classifier workflow that precedes those attribution methods.

## Individual plots

### Scree

```{r scree, fig.width=5, fig.height=4}
# For illustrative purposes we annotate the scree at component 3.
# In a real analysis you would obtain this optimal rank from dkge_cv_rank_loso()
# or dkge_cv_kernel_rank() cross-validation procedures.
one_se_pick <- 3
dkge_plot_scree(fit, one_se_pick = one_se_pick)
```

### Effect-space loadings

```{r loadings, fig.width=5, fig.height=4}
dkge_plot_effect_loadings(fit, comps = 1:3, zscore = TRUE)
```

### Subject contributions

`dkge_plot_subject_contrib()` returns two linked panels. The left panel simply
visualises the subject-level weights that were used while fitting the model
(`fit$weights`). Those weights depend on the `w_method` argument passed to
`dkge()`: in this vignette we deliberately chose `w_method = "none"`, so each
subject keeps unit weight and the bars are all equal to one. Switching to
`"mfa_sigma1"` or `"energy"` would display the corresponding MFA-style or
energy-based weighting actually used in the eigensolve. The heatmap on the
right shows how much norm ("energy") each subject contributes to the selected
components after those weights have been applied.

```{r contrib, fig.width=9, fig.height=3}
contrib <- dkge_plot_subject_contrib(fit, comps = 1:3)
contrib$weights + contrib$energy + patchwork::plot_layout(widths = c(1, 2))
```

### Subspace stability

This diagnostic compares each supplied basis (controlled perturbations here)
against a consensus basis by computing principal angles in the \(K\)
metric. Smaller angles mean that a base reproduces the consensus component more
closely, so parallel lines near zero indicate a stable subspace, whereas large
excursions highlight folds or components that deviate materially.

```{r stability, fig.width=5, fig.height=4}
dkge_plot_subspace_stability(bases, K = fit[['K']], labels = base_labels)
```

## Combine the fit diagnostics

```{r suite, fig.width=8, fig.height=8}
dkge_plot_suite(fit,
                one_se_pick = one_se_pick,
                comps = 1:3,
                bases = bases,
                consensus = fit[['U']],
                base_labels = base_labels,
                top = 5)
```

`dkge_plot_suite()` arranges the available panels and leaves the attribution row
empty when Haufe/LOCO results are absent. Before reporting the dashboard,
replace the sensitivity perturbations with actual refitted bases and interpret
each panel next to the numerical diagnostic that produced it.

## Saving the dashboard

```{r save-suite, eval=FALSE}
dkge_plot_suite(fit,
                bases = bases,
                consensus = fit[['U']],
                base_labels = base_labels,
                save_path = "dkge_dashboard.png",
                width = 10,
                height = 10)
```

## Summary

Use the scree plot to describe retained variation, the salience heatmap to name
effect directions, the contribution panels to detect subject dominance, and
principal angles to summarize refit sensitivity. Save the figure only after
the objects behind those panels come from the analysis being reported. Visual
coherence improves communication; it does not upgrade descriptive diagnostics
into inferential evidence.
