---
title: DKGE Architecture Guide
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{DKGE Architecture Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(dkge)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.show = "hide"
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

This guide describes the current DKGE implementation for contributors. The
fit lifecycle is already split into internal stages, and public constructors
already exist for pipeline specifications and services. The important boundary
is therefore not “old monolith versus future refactor,” but which interfaces
are public, which stages are internal, and which service objects the pipeline
actually consumes today.

## Where are the public entry points?

| Task | Current public API | Main implementation |
|---|---|---|
| Harmonize subject data | `dkge_subject()`, `dkge_data()` | `R/dkge-data.R`, `R/dkge-align-data.R` |
| Define effects and kernels | `dkge_effect_grid()`, `design_kernel()` | `R/dkge-effect-grid.R`, `R/design-kernel.R` |
| Fit a group embedding | `dkge()`, `dkge_fit()` | `R/dkge-data.R`, `R/dkge-fit.R`, `R/dkge-fit-core.R` |
| Compute held-out contrasts | `dkge_contrast()` | `R/dkge-contrast.R`, `R/dkge-loso.R`, `R/dkge-kfold.R`, `R/dkge-analytic.R` |
| Align subject maps | `dkge_prepare_transport()`, `dkge_transport_contrasts_to_medoid()` | `R/dkge-transport.R`, `R/dkge-mapper.R` |
| Test aligned maps | `dkge_signflip_maxT()`, `dkge_infer()` | `R/dkge-inference.R` |
| Run the common workflow | `dkge_pipeline()` | `R/dkge-pipeline.R`, `R/dkge-services.R` |
| Score new subjects | `dkge_predict_subjects()`, `dkge_predict()` | `R/dkge-predict.R` |
| Classify latent outcomes | `dkge_classify()`, `dkge_classification_spec()` | classification and latent-classifier modules |

These functions are exported in `NAMESPACE`. Functions beginning with
`.dkge_` are implementation details even when they provide useful seams for
unit tests. Vignettes and downstream packages should not call them with
`dkge:::`.

## How does `dkge_fit()` work now?

`dkge_fit()` validates its user-facing options and orchestrates four internal
stages in order:

```text
.dkge_fit_prepare()
        |
        v
.dkge_fit_accumulate()
        |
        v
.dkge_fit_solve()
        |
        v
.dkge_fit_assemble()
```

All four stages live in `R/dkge-fit-core.R`; `dkge_fit()` itself lives in
`R/dkge-fit.R`.

### 1. Prepare

`.dkge_fit_prepare()` accepts a `dkge_data` bundle or raw beta/design lists. It
validates finite, common-q inputs; aligns kernel labels with the global effect
order; builds the pooled design ruler `R`; computes row-standardized `Btil`;
derives `Khalf` and `Kihalf`; and resolves spatial and effect-weight
specifications.

The output is an internal payload containing harmonized data and everything
needed by later stages. It is not a partially fitted public object.

### 2. Accumulate

`.dkge_fit_accumulate()` resolves subject weights and effect precision, builds
per-subject raw effect moments, applies analytic or split-half debiasing when
requested, pools observed effect pairs under the selected missingness policy,
and transforms the pooled moment into `Chat`.

The order is a scientific contract: observation masks, precision weighting,
and debiasing operate in raw effect space before the pooled ruler and kernel
mix effect rows. The fitted object retains `effect_moment`, `pair_counts`,
`pair_weight`, `pair_ess`, and `moment_diagnostics` so this step remains
auditable.

```{r missingness-contract-checks, include=FALSE}
set.seed(104)
S_check <- 4L
q_check <- 2L
B_check <- replicate(
  S_check, matrix(rnorm(q_check * 3L), q_check, 3L), simplify = FALSE
)
X_check <- replicate(S_check, diag(q_check), simplify = FALSE)
fit_none_check <- dkge_fit(
  B_check, X_check, K = diag(q_check), rank = 1L,
  w_method = "none", missingness = "none"
)
fit_rescale_check <- dkge_fit(
  B_check, X_check, K = diag(q_check), rank = 1L,
  w_method = "none", missingness = "rescale"
)
fit_shrink_check <- dkge_fit(
  B_check, X_check, K = diag(q_check), rank = 1L,
  w_method = "none", missingness = "shrink",
  miss_args = list(min_pairs = 1L, gamma = 1)
)
stopifnot(isTRUE(all.equal(
  fit_none_check$effect_moment,
  S_check * fit_rescale_check$effect_moment,
  tolerance = 1e-12
)))
stopifnot(isTRUE(all.equal(
  fit_shrink_check$effect_moment,
  fit_rescale_check$effect_moment,
  tolerance = 1e-12
)))
```

The hidden check records the complete-coverage boundary: `none` pools a sum,
whereas `rescale` returns the observed-pair mean; with full coverage and the
shown settings, `shrink` reduces to that same mean.

### 3. Solve

`.dkge_fit_solve()` applies optional CPCA partitioning and ridge regularization,
then dispatches to the pooled eigensolver or joint-diagonalization solver. It
reduces a requested rank when the transformed moment has fewer positive
directions and maps retained eigenvectors back through `Kihalf` to form `U`.

The JD branch deliberately fails closed when effect weighting, missingness
normalization, or debiasing is requested; those combinations are not currently
implemented by that solver.

### 4. Assemble

`.dkge_fit_assemble()` attaches diagnostics, provenance, kernel metadata, and
the stable fit fields. It also assigns one of two representation contracts:

- `block_biprojector` for an exact unregularized block factorization. These
  fits can expose physical block loadings and, with `keep_X = TRUE`, the
  concatenated training matrix.
- `qspace_moment` when pair normalization, debiasing, ridge, CPCA, JD, or other
  moment transformations remove an exact subject-by-feature factorization.
  These fits set `v` and `X_concat` to `NULL` and fail if `keep_X = TRUE` is
  requested.

Contributors must preserve this distinction. A valid q-space eigensystem is
not evidence that physical block loadings exist.

## What does `dkge_pipeline()` integrate?

`dkge_pipeline()` currently performs five steps: obtain a fit, compute
contrasts, optionally transport them, optionally classify, and run sign-flip
inference. It returns all five results plus fit diagnostics.

The configuration surface is uneven but explicit:

| Concern | Objects that exist | Pipeline behavior today |
|---|---|---|
| Contrast | `dkge_contrast_service()` | The pipeline constructs this service internally from `method` and `ridge`; callers cannot pass a contrast-service object as a pipeline argument. |
| Transport | `dkge_transport_spec()`, `dkge_transport_service()` | The pipeline accepts a list, spec, or transport service and runs the typed service. |
| Inference | `dkge_inference_spec()`, `dkge_inference_service()` | The pipeline accepts a list, spec, or inference service and runs the typed service. |
| Classification | `dkge_classification_spec()` | The pipeline accepts a classification result, spec/list, or direct target and calls `dkge_classify()` directly. There is no classification-service class. |

`dkge_run_service()` is the exported generic for manually executing the three
service classes that do exist: contrast, transport, and inference. This makes
them reusable outside `dkge_pipeline()`, but it does not imply that the
pipeline has a general service graph or accepts an arbitrary service sequence.

Likewise, `dkge_predict_subjects()` is already public. It harmonizes matrices,
`dkge_subject` objects, lists, or a `dkge_data` bundle and forwards the named
beta list to `dkge_predict()`. It is current API, not a proposed wrapper.

## What is genuinely unfinished?

Future work should be described narrowly from the current seams:

- allow callers to inject a contrast service instead of having the pipeline
  always construct one;
- decide whether classification needs a service abstraction, then implement
  and test it before documenting one; and
- if a general service graph is introduced, define typed inputs and outputs,
  ordering, cache ownership, and backward compatibility with the present
  pipeline arguments.

Streaming accumulators, alternative solvers, or new backends may be reasonable
experiments, but they are not implied public contracts. Add them only behind
tests that preserve the raw-moment ordering, K-metric algebra, representation
classification, and held-out inference boundaries.

## How should contributors change the architecture?

1. Put input reconciliation in the prepare stage, moment-estimand changes in
   accumulation, eigensystem changes in solve, and output-contract changes in
   assembly.
2. Add focused stage tests and at least one public integration test. The fit
   algebra tests should continue to verify `Chat`, K-orthonormality, and the
   fail-closed `qspace_moment` contract.
3. Keep services and specs honest about current pipeline support. Do not
   document an object merely because a constructor name has been proposed.
4. Preserve subject as the inference unit and reuse transport operators across
   resamples only when the intended inferential procedure conditions on that
   mapping.
5. Update user workflow vignettes only for exported behavior; keep internal
   stage details here and in source comments.

For the user-facing analysis path, see `vignette("dkge-workflow")`. For the
statistical meaning of the fitted moment, see `vignette("dkge-concepts")` and
`vignette("dkge-partial-effect-spaces")`.
