---
title: DKGE versus Partial Least Squares
description: >
  Compare DKGE with PLS on estimand, input contract, and inference route
  rather than on accuracy.
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{DKGE versus Partial Least Squares}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(dkge)
library(ggplot2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
options(
  tibble.print_min = 4,
  tibble.print_max = 4,
  digits = 3
)
set.seed(2024)
theme_set(theme_dkge())
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```


You are choosing between DKGE and PLS for a study, and the two answer different
questions from different inputs. “PLS” names several related methods, and
implementation choices matter, so the summary here follows the neuroimaging
accounts of
[McIntosh and Lobaugh (2004)](https://doi.org/10.1016/j.neuroimage.2004.07.020)
and [Krishnan et al. (2011)](https://doi.org/10.1016/j.neuroimage.2010.07.034).

This page compares what each method estimates and what each requires as input.
It does not benchmark accuracy or power.



---
## What classical PLS does

Classical PLS builds latent variables that maximize the covariance between an
exogenous block, such as task design or behavior, and a brain-data block of
elements by time.

Four **variants** answer different questions: task PLS analyzes condition
differences, behavior PLS examines brain-behavior coupling, seed PLS
investigates connectivity, and spatiotemporal PLS treats space and time
jointly.

The **computational workflow** is:
1. Data are arranged as one matrix \(M \in \mathbb{R}^{(n k) \times (m t)}\),
observations by measurements.
2. Cross-block covariance is formed against an orthonormal design matrix, then
decomposed by SVD.
3. Element and time saliences, design saliences, and singular values are
extracted.
4. Brain scores \(B = M U\) and design scores \(D = C V\) characterize the
patterns.
5. Behavior PLS replaces the design block with behavior matrices and
correlates them per group.

**Inference** is by resampling: permutation for LV significance, bootstrap for
salience reliability, and Procrustes alignment to stabilize resampled LVs.

The method targets whole-pattern effects and time-resolved couplings, using
resampling rather than a parametric null.

## How DKGE relates

Both methods construct compact latent representations, but they begin from
different objects. Classical task or behavior PLS constructs a cross-block
matrix linking a design/behavior block to brain measurements. DKGE pools
subject-level GLM effect moments and applies an explicit effect-space kernel.
The table records implemented differences; it does not establish that one
method is generally more accurate or more stable.

The word *salience* also refers to different objects here. In PLS, brain and
design saliences are singular vectors of the cross-block matrix. In DKGE, the
effect-space salience of a component is \(K U\): the fitted component direction
\(U\) expressed through the design kernel \(K\).

## When the objectives give different answers

The following construction isolates the difference before any implementation
detail can obscure it. Six subjects share a coherent effect of magnitude 2.
A second effect is twice as large within every subject but alternates sign
across subjects. A task-mean PLS SVD sees the cross-subject mean, so the second
effect cancels. DKGE pools each subject's second moment, so its squared energy
does not cancel and the second effect leads.

```{r differing-inputs}
subject_sign <- rep(c(-1, 1), 3)
effect_names <- c("coherent", "sign_varying")

beta_list <- lapply(subject_sign, function(sign) {
  beta <- diag(c(2, 4 * sign))
  dimnames(beta) <- list(effect_names, c("feature1", "feature2"))
  beta
})

subjects <- lapply(seq_along(beta_list), function(i) {
  design <- diag(2)
  colnames(design) <- effect_names
  dkge_subject(beta_list[[i]], design = design, id = paste0("sub", i))
})
```

The PLS calculation below is the SVD of the task-mean beta matrix, written
directly so the example needs no PLS package. The DKGE fit uses an identity
kernel and disables adaptive and subject weighting, leaving the pooled
second-moment objective as the only source of the difference.

```{r differing-answers}
mean_beta <- Reduce(`+`, beta_list) / length(beta_list)
pls_direction <- svd(mean_beta, nu = 2, nv = 0)$u[, 1]

fit_difference <- dkge(
  subjects,
  K = diag(2),
  rank = 2,
  w_method = "none",
  effect_scaling = "none",
  weights = dkge_weights(adapt = "none")
)
dkge_direction <- fit_difference$U[, 1]

data.frame(
  method = c("task-mean PLS", "DKGE"),
  coherent = abs(c(pls_direction[1], dkge_direction[1])),
  sign_varying = abs(c(pls_direction[2], dkge_direction[2]))
)
```

```{r differing-answers-checks, include=FALSE}
pls_moment <- tcrossprod(mean_beta)
dkge_moment <- Reduce(`+`, lapply(beta_list, tcrossprod)) / length(beta_list)
pls_oracle <- eigen(pls_moment, symmetric = TRUE)$vectors[, 1]
dkge_oracle <- eigen(dkge_moment, symmetric = TRUE)$vectors[, 1]

stopifnot(all.equal(unname(pls_moment), diag(c(4, 0)), tolerance = 1e-12))
stopifnot(all.equal(unname(dkge_moment), diag(c(4, 16)), tolerance = 1e-12))
stopifnot(abs(drop(crossprod(pls_direction, pls_oracle))) > 1 - 1e-12)
stopifnot(abs(drop(crossprod(dkge_direction, dkge_oracle))) > 1 - 1e-12)
stopifnot(abs(drop(crossprod(pls_direction, dkge_direction))) < 1e-12)
```

The methods return orthogonal leading directions because they answer different
questions: coherent mean signal versus average within-subject second-moment
energy. This is an estimand example, not a power comparison; neither answer is
better without first deciding which quantity the study is meant to recover.

<style>
table.dkge-vs-pls {
  table-layout: fixed;
  width: 100%;
}
table.dkge-vs-pls th,
table.dkge-vs-pls td {
  vertical-align: top;
}
table.dkge-vs-pls td {
  overflow-wrap: anywhere;
}
</style>

<table class="dkge-vs-pls">
  <colgroup>
    <col style="width: 20%;">
    <col style="width: 40%;">
    <col style="width: 40%;">
  </colgroup>
  <thead>
    <tr>
      <th scope="col">Aspect</th>
      <th scope="col">Partial Least Squares</th>
      <th scope="col">DKGE</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <th scope="row">Latent-space construction</th>
      <td>SVD on cross-block covariance between design/behavior and brain data; columns of \(U\) and \(V\) are saliences.</td>
      <td>Eigen-decomposition of a design-kernel-weighted covariance, producing orthonormal components \(U\).</td>
    </tr>
    <tr>
      <th scope="row">Design information</th>
      <td>Specified through the exogenous/design block; the exact coding and available structure depend on the PLS variant.</td>
      <td>Explicit design kernel \(K\) encodes factorial structure, smoothness, or prior correlations among effects.</td>
    </tr>
    <tr>
      <th scope="row">Data normalisation</th>
      <td>Conditions averaged or mean-centred before SVD; each voxel treated equally.</td>
      <td>Row standardisation of subject betas using the pooled design Cholesky factor; optional spatial/reliability weights \(\Omega_s\).</td>
    </tr>
    <tr>
      <th scope="row">Cross-validation</th>
      <td>Classical neuroimaging workflows commonly use permutation for LV significance and bootstrap for salience stability; predictive cross-validation can be added but is not the same estimand.</td>
      <td>LOSO / K-fold cross-fitting (`dkge_contrast`), analytic approximations, parametric or bootstrap inference with cached transports.</td>
    </tr>
    <tr>
      <th scope="row">Transport / alignment</th>
      <td>Outputs latent scores; spatial interpretation relies on the original voxel grid.</td>
      <td>Provides barycentric kNN and Sinkhorn transports, anchor graphs, and voxel decoders for consistent spatial maps across parcellations.</td>
    </tr>
    <tr>
      <th scope="row">Reliability weighting</th>
      <td>Baseline formulations often use uniform observation weights; weighted, sparse, and regularized PLS variants also exist.</td>
      <td>Subject- and cluster-level reliabilities enter directly (e.g. sizes, inverse variances), influencing fits and transport.</td>
    </tr>
    <tr>
      <th scope="row">Spatiotemporal support</th>
      <td>ST-PLS handles time by stacking features.</td>
      <td>DKGE works on any GLM-derived beta blocks; temporal modelling is delegated to the design matrix and optional kernels.</td>
    </tr>
    <tr>
      <th scope="row">Implementation focus</th>
      <td>Exploratory LVs; complementary to other analyses.</td>
      <td>Integrated workflow for group GLM analysis, transport, and inference tailored to fMRI/ERP pipelines.</td>
    </tr>
  </tbody>
</table>

## Similarities worth noting

The two share a mathematical core. Both use SVD or eigendecomposition to
obtain orthogonal latent patterns and scores.

Both rest on resampling, implemented differently: PLS permutes and bootstraps
directly; DKGE cross-fits and then resamples the held-out fields.

Both require reading latent loadings or saliences alongside subject scores to
see what a component means.

## Which method matches the question?

Choose a PLS formulation when the target quantity is cross-block covariance
between a brain block and an experimental or behavioral block and the chosen
PLS implementation supplies the resampling procedure you need. Choose DKGE
when the inputs are already subject-level GLM beta matrices, the scientific
model requires an explicit effect-space metric, or subject-specific
parcellations must be transported before feature-level summaries.

These are workflow distinctions, not demonstrated performance advantages.
DKGE's cross-fitting addresses basis reuse only when the selected mode actually
refits the basis; reliability weights help only when their construction is
defensible; and spatial transport is useful only when its features and mapper
support plausible correspondence. Conversely, the table does not cover the
many sparse, regularized, predictive, or multilevel PLS variants.

For an empirical comparison, prespecify one estimand, fit both methods to the
same analysis units, use identical outer resampling splits, and compare a
held-out metric with uncertainty. Do not infer superiority from the fact that
one package exposes a helper that the other workflow leaves to the analyst.

## Where to go next

- `vignette("dkge-concepts")` sets out what each DKGE analysis licenses you to
  claim, which is the question this comparison keeps returning to.
- `vignette("dkge-design-kernels")` covers the kernel that gives DKGE its
  metric, and which classical PLS has no analogue for.
- `vignette("dkge-contrasts-inference")` is the inference route for a
  prespecified effect, where PLS would resample the latent variable instead.
