---
title: "Components and Interpretability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Components and Interpretability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(4)
```

This vignette shows how to interpret DKGE components, rotate them if needed, and
project new data.

## Example Fit

```{r fit}
library(dkge)
S <- 4; q <- 5; P <- 18; T <- 70
betas <- replicate(S, matrix(rnorm(q * P), q, P), simplify = FALSE)
designs <- replicate(S, {
  X <- matrix(rnorm(T * q), T, q)
  qr.Q(qr(X))
}, simplify = FALSE)
subjects <- lapply(seq_len(S), function(s) dkge_subject(betas[[s]], designs[[s]], id = paste0("sub", s)))
 bundle <- dkge_data(subjects)
fit <- dkge(bundle, kernel = diag(q), rank = 3)
```

## Inspect Loadings and Scores

```{r loadings}
round(fit$U, 3)         # effect-space loadings
lapply(fit$Btil, function(Bts) round(Bts[, 1:3, drop = FALSE], 2))[1]
```

- Columns of `fit$U` correspond to components in effect space.
- `Btil` stores subject betas after row-standardisation; multiply with `fit$U`
  to obtain cluster scores per component.

## Projecting Subjects into Component Space

```{r project}
subject_scores <- dkge_project_btil(fit, fit$Btil)
str(subject_scores, max.level = 1)
```

Each subject score matrix has dimensions clusters × components. Average scores
for a quick component summary.

```{r score-plot, fig.height=4, fig.width=5, fig.alt="Line plot showing average DKGE component scores across clusters."}
avg_scores <- Reduce("+", subject_scores) / length(subject_scores)
matplot(avg_scores, type = "l", lwd = 2, xlab = "Cluster", ylab = "Score",
        main = "Average component scores")
legend("topright", legend = paste0("Component ", 1:ncol(avg_scores)), col = 1:ncol(avg_scores), lty = 1:ncol(avg_scores))
```

## Rotating Components

Use `dkge_procrustes_K()` to rotate components toward a target basis (e.g.
canonical contrasts or prior loadings) while staying in the design-kernel
metric.

```{r rotate}
# target basis: identity for first two effects
B_target <- diag(1, q)[, 1:2]
rot <- dkge_procrustes_K(fit$U[, 1:2], B_target, fit$K)
round(rot$R, 3)
```

Multiply the original loadings by `rot$R` to obtain aligned components.

## Projecting New Data

```{r new-data}
new_beta <- matrix(rnorm(q * P), q, P)
projected <- dkge_project_clusters(fit, new_beta)
head(projected)
```

`dkge_project_clusters()` returns component scores for each cluster in the new
data, ready for contrast analysis or transport.

## Interpreting Components

- Inspect `fit$weights` to see which subjects dominate each component.
- Plot rows of `fit$v` (cluster-space loadings) to identify spatial patterns.
- Use `dkge_component_stats()` (from the inference vignette) to obtain
  statistical evidence per component.
