---
title: "Functional alignment across subjects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Functional alignment across subjects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(dkge)
```

Suppose each participant has a beta map on a different subject-specific
parcellation. Averaging after resampling parcels to MNI coordinates assumes
that anatomical proximity supplies correspondence. Functional alignment asks a
stronger question: which parcels have comparable response signatures, and how
should each participant's contrast field be moved onto one identified display
support? The result is a matrix with one aligned row per participant, followed
by group inference and rendering on that same support.

This workflow keeps five objects separate: source features, reference support,
functional template, fitted correspondence, and aligned maps. That separation
is what prevents a bare coordinate grid or a plotting helper from quietly
becoming a functional model.

```{r simulate-inputs, include=FALSE}
set.seed(2711)
S <- 6L
q <- 5L
P <- 8L
effects <- paste0("effect", seq_len(q))
parcels <- paste0("parcel", seq_len(P))
subject_ids <- paste0("subject", seq_len(S))

latent_signal <- lapply(seq_len(S), function(s) {
  matrix(rnorm(q * P), q, P,
         dimnames = list(effects, parcels))
})
draw_channel <- function(noise) {
  out <- lapply(latent_signal, function(signal) {
    signal + matrix(rnorm(q * P, sd = noise), q, P,
                    dimnames = dimnames(signal))
  })
  names(out) <- subject_ids
  out
}
primary_betas <- draw_channel(0.5)
alignment_betas <- draw_channel(0.3)
validation_betas <- draw_channel(0.35)
designs <- lapply(seq_len(S), function(s) {
  X <- qr.Q(qr(matrix(rnorm(15 * q), 15, q)))
  colnames(X) <- effects
  X
})
subjects <- lapply(seq_len(S), function(s) {
  dkge_subject(primary_betas[[s]], designs[[s]], id = subject_ids[[s]])
})
K <- diag(q)
dimnames(K) <- list(effects, effects)
centroids <- lapply(seq_len(S), function(s) {
  cbind(x = seq_len(P), y = sin(seq_len(P) / 2) + s / 20, z = 0)
})
names(centroids) <- subject_ids
sizes <- stats::setNames(lapply(seq_len(S), function(s) rep(1, P)), subject_ids)
```

## What are the inputs?

For subject (s), the primary beta map is (q \times P_s): rows share effect
names, while columns are that subject's parcels. `K` is the (q \times q)
effect-space covariance or kernel. Functional alignment needs at least one
separate beta-map channel with the same effects and parcel ordering; a second
independent channel lets reference selection be evaluated out of sample.

Fit the shared effect-space model and estimate the contrast without using each
held-out subject to define its own basis:

```{r fit-and-contrast}
fit <- dkge_fit(
  dkge_data(subjects), K = K, rank = 2,
  w_method = "mfa_sigma1", effect_scaling = "none"
)
contrast <- dkge_contrast(
  fit, c(1, -1, rep(0, q - 2)),
  method = "loso", align = FALSE
)
```

The fit-level MFA weights affect the pooled moment. They are not imported as
second-level inverse-variance weights:

```{r fit-weights}
range(fit$weights)
mean(fit$weights)
```

## How is functional correspondence identified?

Build over-ranked kernel-image features from two independent channels. These
features retain `rank(K)` coordinates rather than the smaller estimation rank.

```{r independent-features}
features <- dkge_alignment_features(
  fit, contrast, independent_betas = alignment_betas,
  independent_data_hash = "alignment-acquisition-A"
)
validation <- dkge_alignment_features(
  fit, contrast, independent_betas = validation_betas,
  independent_data_hash = "alignment-acquisition-B"
)
```

Use one channel to fit each candidate correspondence and the other to score
symmetric reconstruction. Only a subject selected by this criterion is a
medoid; a caller-fixed subject is simply a reference subject.

```{r select-reference}
mapper <- dkge_mapper_spec(
  "sinkhorn", epsilon = 0.1, lambda_emb = 0.8,
  lambda_spa = 0.2, warm_start = FALSE
)
selection <- dkge_select_reference_subject(
  centroids, features, validation, sizes = sizes,
  mapper = mapper, method = "functional_heldout"
)
selection
```

```{r check-selection, include=FALSE}
stopifnot(
  inherits(selection, "dkge_reference_selection"),
  selection$eligibility$eligible,
  identical(selection$method, "functional_heldout")
)
```

## Why learn a group template on the selected support?

The selected subject supplies real coordinates and labels, but its functional
features are still one noisy measurement. Initialize those features on the
same support, then iteratively refit every subject and take mass-aware template
updates. Unit row normalization keeps the functional term from disappearing
through repeated barycentric contraction.

```{r make-support}
ref <- selection$reference_subject
support <- dkge_reference_support(
  centroids[[ref]], labels = parcels,
  provenance = list(
    kind = "selected_subject_support",
    subject_id = selection$reference_subject_id,
    reference_selection_hash = selection$structural_hash
  )
)
initializer <- dkge_functional_template(
  support, features$features[[ref]], feature_source = "independent",
  provenance = list(independent_data_hash = "alignment-acquisition-A"),
  reference_selection = selection
)
```

```{r fit-template}
template <- dkge_fit_functional_template(
  support, features, centroids, reference_selection = selection,
  sizes = sizes, mapper = mapper,
  initialization = "supplied_independent",
  initial_template = initializer,
  max_iter = 25, tolerance = 0.01, objective_tolerance = 0.01
)
template
```

```{r template-checks, include=FALSE}
stopifnot(
  template$fitting$converged,
  all(is.finite(template$fitting$trajectory$objective)),
  all(abs(sqrt(rowSums(template$features^2)) - 1) < 1e-8),
  template$fitting$final_feature_stats$effective_rank > 1
)
```

The trajectory is an audit record, not a decorative optimizer trace:

```{r template-summary}
data.frame(
  iterations = template$fitting$iterations,
  final_objective = template$fitting$final_objective,
  effective_rank = template$fitting$final_feature_stats$effective_rank,
  eligibility = template$eligibility$status
)
```

## How do you obtain aligned rows and inference?

Apply the final operators to the cross-fitted subject contrasts. Equal subject
weighting is explicit in the resulting object.

```{r align-subject-rows}
aligned <- dkge_align_to_template(template, support, contrast)
aligned
```

```{r aligned-checks, include=FALSE}
stopifnot(
  identical(aligned$subject_weighting, "equal_subject"),
  all(vapply(aligned$values, function(x) all(is.finite(x)), logical(1))),
  all(vapply(aligned$values, nrow, integer(1)) == S),
  all(vapply(aligned$values, ncol, integer(1)) == support$n_locations)
)
```

Independent correspondence is the recommended default, but it does not erase
the current contrast estimator's dependence on an estimated, rank-truncated
same-data latent span. The frozen calibration court therefore labels this
combination **approximate**. The explicit override below records that choice;
it does not turn the test into an exact one.

The optional same-data residualized feature mode has an additional assumption:
its covariance cancellation is exact only under the declared separable
effect-by-parcel covariance model, conditional on fixed generating directions.
Its preprocessing receipt verifies the algebraic transforms, not that
separability holds in an observed cohort.

```{r aligned-inference}
set.seed(2712)
inference <- dkge_infer_aligned(
  aligned, inference = "signflip", correction = "maxT", n_perm = 199,
  allow_approximate_alignment = TRUE
)
head(inference$p_adjusted[[1]])
```

```{r inference-checks, include=FALSE}
stopifnot(
  identical(inference$metadata$alignment$status, "approximate"),
  isTRUE(inference$metadata$alignment$approximate_override),
  all(is.finite(inference$p_adjusted[[1]])),
  all(inference$p_adjusted[[1]] >= 0 & inference$p_adjusted[[1]] <= 1)
)
```

## What does rendering do?

Rendering consumes the identified support and already aligned values. It never
learns correspondence. A support can instead carry a voxel decoder, or its
labels can be passed to `dkge_paint_reference_map()` for a volume.

```{r render-aligned}
renderer <- dkge_renderer(support)
rendered <- dkge_render_aligned(renderer, aligned)
head(rendered$support_values[[1]])
```

```{r render-plot, echo=FALSE, fig.height=3.5, fig.alt="Weighted group contrast across the selected reference support."}
plot(
  rendered$support_values[[1]], type = "b",
  xlab = "Reference-support parcel", ylab = "Aligned group contrast"
)
abline(h = 0, lty = 2, col = "grey60")
```

The horizontal order comes from the selected support. The ordinate is a group
summary of aligned subject rows, not a value learned from the renderer.

## A compact glossary

| Object | What it owns | What it does not imply |
|---|---|---|
| Shared effect space | Named design effects and the metric `K` | Spatial correspondence |
| Reference support | Coordinates, labels, topology, optional decoder | Functional data or a mapping |
| Correspondence | Fitted subject-to-support operators | A group statistic |
| Functional template | Target response signatures and masses | Privileged truth from one subject |
| Aligned subject rows | One contrast row per subject on one support | Independent parcel-level samples |
| Group inference | Subject-level uncertainty and multiplicity control | New correspondence fitting |
| Rendering | Display or decoding of an existing support statistic | Functional alignment |

A fixed MNI lattice can be useful display support, but coordinates alone have
no functional mapping to any participant. It becomes a functional target only
after template features have been learned on it from an eligible channel.

For the effect-space model, see `vignette("dkge-concepts")`; for dense spatial
decoding, see `vignette("dkge-dense-rendering")`; and for the frozen calibration
limits behind the approximate label, consult `?dkge_infer_aligned`.
