---
title: "Design Kernels and Model Tuning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Design Kernels and Model Tuning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

This note focuses on how the design kernel steers DKGE fits and how to tune rank
or penalty parameters before running large analyses.

## Simulated Working Example

```{r data}
library(dkge)
S <- 4; q <- 4; P <- 30; T <- 80

betas <- replicate(S, matrix(rnorm(q * P, sd = 0.5), q, P), simplify = FALSE)
true_kernel <- 0.9 * (diag(q) + 0.3)
# inject structure: effects 1&2 correlate, 3&4 share mild coupling
cholK <- chol(true_kernel)
betas <- lapply(betas, function(B) cholK %*% B + 0.1 * matrix(rnorm(q * P), q, P))

designs <- replicate(S, {
  X <- matrix(rnorm(T * q), T, q)
  qr.Q(qr(X))
}, simplify = FALSE)

subjects <- lapply(seq_len(S), function(s) dkge_subject(betas[[s]], designs[[s]], id = paste0("sub", s)))
data_bundle <- dkge_data(subjects)
```

`dkge_data()` aligns effect ordering and caches common metadata. The kernel we
supply next controls how strongly effects are smoothed or grouped.

## Building Kernels with `design_kernel()`

```{r kernels}
# tell dkge that effects 1&2 belong to the same factor, 3&4 to another
spec <- list(
  list(name = "factor12", type = "nominal", levels = c("e1", "e2"), rho = 0.8),
  list(name = "factor34", type = "nominal", levels = c("e3", "e4"), rho = 0.6)
)
K_struct <- design_kernel(spec, effects = colnames(data_bundle$designs[[1]]))

round(K_struct$K, 2)
```

- Each term introduces a block-similarity component.
- `rho` weights the term; defaults to 1. Combine multiple terms to encode
  interactions or ordered trends.
- `design_kernel()` returns both the kernel matrix and metadata used by CV
  helpers below.

## Rank and Kernel Tuning via Cross-Validation

`dkge_cv_kernel_rank()` evaluates candidate ranks and kernel weights using
leave-one-subject-out style objectives. Here we scan ranks 1–3 and shrink a
single `rho` parameter for the two factors.

```{r kernel-cv, message=FALSE}
cv_grid <- expand.grid(rank = 1:3, rho = seq(0.4, 1.0, by = 0.3))
cv_results <- dkge_cv_kernel_rank(
  data_bundle,
  kernel_spec = spec,
  rank_candidates = cv_grid$rank,
  rho_grid = list(factor12 = cv_grid$rho, factor34 = cv_grid$rho),
  folds = NULL,            # default LOSO
  summary = median         # robust aggregation of fold errors
)

head(cv_results)
```

The score reported by `dkge_cv_kernel_rank()` is the average reconstruction
error of left-out subjects in the K-metric. Lower is better. Pick the
rank/weights combination with the minimum score and refit.

```{r refit}
best <- cv_results[which.min(cv_results$cv_score), ]
best

K_best <- design_kernel(spec, rho = list(factor12 = best$factor12, factor34 = best$factor34))$K
fit <- dkge(data_bundle, kernel = K_best, rank = best$rank)
round(fit$sdev, 3)
```

## Penalising Noisy Effects

When certain effects are known to be high variance (e.g. motion regressors), add
diagonal ridge terms via `dkge(..., ridge = value)` or pre-scale the kernel.
Higher ridge values shrink small singular values and stabilise fits when subject
counts are limited.

```{r ridge}
fit_ridge <- dkge(data_bundle, kernel = K_best, rank = best$rank, ridge = 0.2)
round(fit_ridge$sdev, 3)
```

## Summary

- Use `design_kernel()` to encode your scientific priors; start with `diag(q)`
  if in doubt, then add structured terms gradually.
- Run `dkge_cv_kernel_rank()` (loso by default) to select rank and kernel
  weights before expensive inference.
- Add ridge regularisation when singular values drop sharply or subjects are
  few; check `fit$sdev` to verify stability.
