---
title: Mapper Customization and Performance
description: >
  Choose and cache a spatial mapper, or add your own, when transport
  dominates runtime.
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Mapper Customization and Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(dkge)
library(ggplot2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
options(
  tibble.print_min = 4,
  tibble.print_max = 4,
  digits = 3
)
set.seed(2024)
theme_set(theme_dkge())
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```


Spatial mapping can dominate runtime when it is repeatedly refit for many
contrasts or resamples. The first performance decision is therefore not simply
“which mapper is fastest?” but “which stage can be fitted once and safely
reused?”

This page separates mapper **fit** cost from **apply** cost, then explains when
kNN, Sinkhorn, warm starts, or a custom mapper are appropriate. It assumes you
already understand why transport is needed; see
`vignette("dkge-dense-rendering")` for that workflow.

| Need | Start with | Tradeoff |
|---|---|---|
| Fast local interpolation | kNN | no global mass-matching constraint |
| Mass-preserving global coupling | Sinkhorn | more expensive fit and tuning |
| Many fields on unchanged geometry | either fitted mapper | reuse the operator; do not resolve |

## Mapper Factory Recap

```{r mapper-factory}
library(dkge)
dkge_mapper("knn", k = 8, sigx = 3)
dkge_mapper("sinkhorn", epsilon = 0.05, lambda_xyz = 1, lambda_feat = 0)
```

The `dkge_mapper()` function configures the dense spatial pipeline. A fitted
Sinkhorn mapper keeps the joint plan, the target-conditional application
operator, and convergence diagnostics. Reusing that fitted object makes later
`apply_mapper()` calls matrix multiplications rather than new OT solves.

## Speed Comparison

```{r performance-compute, include=FALSE}
P <- 80
centroids <- matrix(rnorm(P * 3), P, 3)
anchors <- matrix(rnorm(200 * 3), 200, 3)
values <- rnorm(P)

knn_mapper <- dkge_mapper("knn", k = 12, sigx = 5)
sink_mapper <- dkge_mapper("sinkhorn", epsilon = 0.05, lambda_xyz = 1,
                           warm_start = FALSE)

time_fit <- function(mapper, reps = 10L) {
  system.time(replicate(
    reps, fit_mapper(mapper, centroids, anchors), simplify = FALSE
  ))[["elapsed"]] / reps
}
knn_fit_seconds <- time_fit(knn_mapper)
sink_fit_seconds <- time_fit(sink_mapper)
knn_fit <- fit_mapper(knn_mapper, centroids, anchors)
sink_fit <- fit_mapper(sink_mapper, centroids, anchors)

time_apply <- function(fitted, reps = 2000L) {
  system.time(replicate(reps, apply_mapper(fitted, values)))[["elapsed"]] / reps
}
performance <- data.frame(
  stage = rep(c("fit", "apply"), each = 2),
  mapper = rep(c("kNN", "Sinkhorn"), 2),
  seconds_per_call = c(knn_fit_seconds, sink_fit_seconds,
                       time_apply(knn_fit), time_apply(sink_fit))
)
stopifnot(all(is.finite(performance$seconds_per_call)),
          all(performance$seconds_per_call >= 0))
```

```{r performance}
knitr::kable(performance, digits = 5)
```

```{r performance-plot, echo=FALSE, fig.height=3.8, fig.alt="Log-scale bar chart comparing mapper fit and apply time for k-nearest-neighbour and Sinkhorn mappings."}
performance$seconds_for_plot <- pmax(performance$seconds_per_call, .Machine$double.eps)
ggplot(performance, aes(mapper, seconds_for_plot, fill = mapper)) +
  geom_col(width = 0.65) +
  facet_wrap(~ stage, scales = "free_y") +
  scale_y_log10() +
  labs(
    title = "Fit and apply are different performance questions",
    x = NULL, y = "Seconds per call (log scale)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

The table separates the solve from application. kNN fitting constructs local
neighbourhoods; Sinkhorn fitting constructs a dense cost matrix and iterates to
match both marginals, so its fit is usually the expensive stage. Once fitted,
both backends reuse their mappings. Measure both stages at your own parcel and
anchor counts rather than extrapolating from apply-only timings.

## Warm Starts and Dual Caching

DKGE caches duals only after a solve satisfies its requested marginal
tolerance. The key includes every cost and mass entry, so warm starts cannot
cross between merely similar problems. The cache is process-local; fitted
renderer objects store plans and operators, not dual vectors.

```{r warm-start}
dkge_clear_sinkhorn_cache()
cached_mapper <- dkge_mapper("sinkhorn", epsilon = 0.05, lambda_xyz = 1)
cold_fit <- fit_mapper(cached_mapper, centroids, anchors)
warm_fit <- fit_mapper(cached_mapper, centroids, anchors)

rbind(
  cold = unlist(cold_fit$stats$diagnostics[c("iterations", "cache_hit")]),
  repeated = unlist(warm_fit$stats$diagnostics[c("iterations", "cache_hit")])
)
```

```{r warm-start-checks, include=FALSE}
stopifnot(cold_fit$stats$diagnostics$converged,
          warm_fit$stats$diagnostics$converged,
          isTRUE(warm_fit$stats$diagnostics$cache_hit),
          identical(as.matrix(cold_fit$plan), as.matrix(warm_fit$plan)))
```

For this exact repeat, the cached converged plan is returned with zero new
iterations. If you request a tighter tolerance, DKGE instead uses the cached
duals as an initialization and continues solving.

## Adding Custom Mappers

To integrate a custom mapping approach into the DKGE framework, you need to implement two key S3 methods that follow the established naming convention recognized by `dkge_mapper()`.

```{r custom-mapper}
fit_mapper.dkge_mapper_mytransport <- function(mapper, subj_points,
                                                anchor_points, ...) {
  structure(list(Q = nrow(anchor_points)),
            class = "dkge_mapper_fit_mytransport")
}

apply_mapper.dkge_mapper_fit_mytransport <- function(fitted_mapper, values, ...) {
  rep(mean(values), fitted_mapper$Q)
}

custom <- dkge_mapper("mytransport")
custom_fit <- fit_mapper(custom, centroids, anchors)
head(apply_mapper(custom_fit, values))
```

The constructor accepts the custom identifier and S3 dispatch finds the two
methods. A renderer-compatible extension must also validate the standard point
inputs and return one value per anchor, as this example does.

## Practical Guidance

The choice between mapping methods should be guided by your specific analysis requirements and data characteristics. For quick exploratory analyses or situations where subject clusters already show good anatomical alignment, kNN mapping provides an efficient and straightforward solution.

Prefer Sinkhorn mapping when anatomical shifts between subjects are substantial, or when latent features should guide the matching. In these cases, you can tune the `epsilon` parameter upward to achieve faster convergence and smoother transport plans, though this comes with a trade-off in matching precision.

For repeated maps, reuse renderer or fitted-mapper objects so the already-solved
plans and application operators are retained. Warm-start caching is most useful
when an identical fit is requested again; check `warm_fit$stats$diagnostics` on the fitted mapper instead of
assuming that a cache was used.


## Where to go next

- `vignette("dkge-dense-rendering")` — what the mappers benchmarked here are
  actually for, if you arrived at this page from a search.
- `vignette("dkge-anchors")` — anchor construction, which sets the problem size
  the mappers are solving.
