---
title: "Getting Started with DKGE"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: false
    df_print: paged

vignette: >
  %\VignetteIndexEntry{Getting Started with DKGE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r library, message=FALSE}
library(dkge)
set.seed(42)
```

Design-Kernel Group Embedding (DKGE) is an R package for multi-subject fMRI
analysis. It works in the compressed **q-space** of design effects (typically
10–100 dimensions) rather than voxel-space, enabling memory-efficient group
fits, leave-one-subject-out contrasts, and optimal-transport alignment to a
common parcellation — all without ever forming large subject-by-voxel matrices.

## Minimal example

Five subjects, three effects, twenty voxels each:

```{r quick-start}
S <- 5; q <- 3; P <- 20
subjects <- lapply(seq_len(S), function(s) {
  beta   <- matrix(rnorm(q * P), q, P)
  design <- diag(q); colnames(design) <- paste0("eff", seq_len(q))
  dkge_subject(beta, design = design, id = paste0("sub", s))
})

fit <- dkge(subjects, K = diag(q), rank = 2)
fit          # print method shows eigenspectrum and subject count
```

The returned object is a `multiblock_biprojector` with q-space loadings in
`fit$U`, compressed projections in `fit$Btil`, and diagnostics in
`fit$weights` and `fit$evals`.

## Reading order

| Vignette | Purpose |
|----------|---------|
| **You are here** | Overview and entry point |
| [Workflow](dkge-workflow.html) | Complete fit → contrast → transport → visualise pipeline |
| [Design Kernels](dkge-design-kernels.html) | Encoding factor structure with `design_kernel()` |
| [Contrasts & Inference](dkge-contrasts-inference.html) | LOSO contrasts, analytic and bootstrap inference |
| [Classification](dkge-classification.html) | Cross-validated decoding with `dkge_classify()` |
| [Weighting Strategies](dkge-weighting.html) | Spatial (`Omega_list`), subject, and transport weights |
| [Adaptive Voxel Weighting](dkge-adaptive-weighting.html) | Fold-safe `dkge_weights()` API |
| [Components](dkge-components.html) | Inspecting group-level components and loadings |
| [Plotting](dkge-plotting.html) | `theme_dkge()`, scree plots, contribution maps |
| [Dense Rendering](dkge-dense-rendering.html) | Mapping cluster values to voxel space |
| [Anchors](dkge-anchors.html) | `dkge_anchor_fit()` and anchor-based prediction |
| [CPCA](dkge-cpca.html) | Contrastive PCA inside the DKGE framework |
| [DKGE vs PLS](dkge-vs-pls.html) | Comparison with partial least squares approaches |
| [Performance](dkge-performance.html) | Mapper selection, warm starts, custom backends |
| [Architecture](dkge-architecture.html) | Design document: planned modular refactor (not current API) |

## Key concepts at a glance

**`dkge_subject(beta, design)`** — wraps a single subject's `q × P` beta
matrix and `T × q` design matrix. Supply optional `omega` (cluster weights)
and `id`.

**`dkge_data(betas, designs)`** — bundles a list of subjects, aligns
partially-overlapping effect sets, and records provenance.

**`dkge(subjects, K, rank)`** — main entry point. `K` is the `q × q` design
kernel; pass `design_kernel(factors, basis = "effect")` for structured factor
designs.

**`dkge_contrast(fit, contrast_vec, method = "loso")`** — leave-one-subject-out
contrasts in q-space; pair with `dkge_transport_contrasts_to_medoid()` to map
to a common parcellation.

**`dkge_classify(fit, targets, method = "lda")`** — fully cross-validated
decoding with optional permutation tests. Supply targets via
`dkge_targets(fit, ~ factor1 + factor2)` or a plain contrast weight matrix.

## Getting help

- `?dkge` — main fitting function
- `?dkge_contrast` — LOSO and k-fold contrasts
- `?dkge_classify` — classification and decoding
- `?design_kernel` — structured factor kernels
- Issues: <https://github.com/bbuchsbaum/dkge/issues>
