---
title: Components and Interpretability
description: >
  Read a fitted component in effect space and on shared clusters, and rotate
  or project it without treating description as inference.
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Components and Interpretability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(dkge)
library(ggplot2)
library(tidyr)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
options(
  tibble.print_min = 4,
  tibble.print_max = 4,
  digits = 3
)
set.seed(2024)
theme_set(theme_dkge())
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```


A component number has no intrinsic scientific meaning. Once DKGE returns a
fitted basis you still have to say what each direction represents, and that
requires reading it in more than one place.

You need to connect three views of the same fitted direction: the effects that
define it, the spatial units on which subjects express it, and any target used
to rotate or label it.

This vignette asks how a fitted component is expressed in effect space and on
a shared toy cluster grid. It then shows rotation and projection without
treating those descriptive operations as inferential results.

```text
effect saliences (what) -> subject cluster scores (where, within support) -> rotation or new-data projection (reuse)
```

## Example Fit

```{r fit}
library(dkge)
S <- 4; q <- 5; P <- 18; T <- 70
effect_names <- paste0("effect", seq_len(q))
cluster_names <- paste0("cluster", seq_len(P))
cluster_axis <- seq(-1, 1, length.out = P)
shared_signal <- rbind(
  exp(-4 * (cluster_axis + 0.45)^2),
  -exp(-4 * (cluster_axis - 0.45)^2),
  cluster_axis,
  rep(0, P),
  rep(0, P)
)
betas <- lapply(seq_len(S), function(s) {
  B <- shared_signal + matrix(rnorm(q * P, sd = 0.35), q, P)
  dimnames(B) <- list(effect_names, cluster_names)
  B
})
designs <- replicate(S, {
  X <- matrix(rnorm(T * q), T, q)
  X <- qr.Q(qr(X))
  colnames(X) <- effect_names
  X
}, simplify = FALSE)
subjects <- lapply(seq_len(S), function(s) dkge_subject(betas[[s]], designs[[s]], id = paste0("sub", s)))
bundle <- dkge_data(subjects)
fit <- dkge(bundle, K = diag(q), rank = 3)
```

By default `dkge()` accumulates subject contributions in the shared effect
space and diagonalizes the pooled moment. To request joint diagonalization of
the subject matrices instead, set `solver = "jd"` (and optionally pass
`jd_control = dkge_jd_control(...)` to configure the optimizer). Treat agreement
or disagreement with the pooled solution as a model comparison; the solver
choice does not guarantee greater stability.

```{r fit-jd, eval=FALSE}
fit_jd <- dkge(bundle,
               K = diag(q),
               rank = 3,
               solver = "jd",
               jd_control = dkge_jd_control(maxit = 200, tol = 1e-8))
```

Both solvers return the same object structure (`fit$U`, `fit$Chat`,
projections), so downstream code does not change.
subsequent workflow in this vignette applies unchanged regardless of the solver choice.

## Which effects define each component?

```{r loadings}
c(effects = nrow(fit$U), components = ncol(fit$U), subjects = length(fit$Btil))
```

The compact summary establishes the object shapes without printing internal
matrices. The interpretable effect-space quantity is the design-weighted
salience (K U):

```{r loading-heatmap, fig.height=4, fig.alt="Heatmap showing how five design effects contribute to the first three DKGE components."}
dkge_plot_effect_loadings(fit, comps = 1:3)
```

Read each column as a relative pattern. Effects with opposite colors pull the
component in opposite directions; a column dominated by one row is closer to a
single-effect axis. Component signs are arbitrary, so the within-column pattern
matters more than whether red or blue is positive.

`fit$Btil` stores the row-standardized subject beta blocks. Multiplying those
blocks by the component directions produces spatial-unit scores, which answer
the next question.

## Projecting Subjects into Component Space

```{r project}
subject_scores <- dkge_project_btil(fit, fit$Btil)
str(subject_scores, max.level = 1)
```

Each subject score matrix has dimensions clusters × components. Cluster rows
are only comparable across subjects when they already refer to the same spatial
support. This simulation deliberately uses a shared 18-cluster indexing scheme,
so a row-wise mean is meaningful. With subject-specific parcellations, first
transport the scores to a common medoid, anchor, or voxel space; equal matrix
dimensions alone do not establish anatomical correspondence.

```{r score-plot, fig.height=4, fig.width=5, fig.alt="Line plot showing average DKGE component scores across clusters."}
avg_scores <- Reduce("+", subject_scores) / length(subject_scores)
avg_df <- as.data.frame(avg_scores)
component_cols <- paste0("Component ", seq_len(ncol(avg_df)))
names(avg_df) <- component_cols
avg_df$Cluster <- seq_len(nrow(avg_df))
avg_long <- tidyr::pivot_longer(avg_df,
                                cols = seq_along(component_cols),
                                names_to = "Component",
                                values_to = "Score")

ggplot(avg_long, aes(x = Cluster, y = Score, colour = Component)) +
  geom_line(linewidth = 1.1) +
  labs(title = "Average component scores", y = "Score") +
  scale_x_continuous(breaks = seq_len(nrow(avg_scores))) +
  theme(legend.position = "top")
```

The plot summarizes expression on this shared toy grid. It is descriptive: a
large value identifies a cluster-component pairing to inspect, not a
significant spatial effect.

## Rotating Components

`dkge_procrustes_K()` rotates components toward a target basis in the \(K\)
metric: canonical contrasts, or loadings from an earlier study. The rotation
stays inside the design-kernel metric, so the solution's properties are
preserved.

```{r rotate}
# target basis: identity for first two effects
B_target <- dkge_k_orthonormalize(diag(1, q)[, 1:2], fit$K)
rot <- dkge_procrustes_K(B_target, fit$U[, 1:2], fit$K)
round(rot$U_aligned, 3)
```

The returned `rot$U_aligned` is the fitted basis rotated toward `B_target`.
`rot$d` reports the objective actually achieved by `rot$R`; when reflections
are forbidden, it can be smaller than `rot$unconstrained_d`.

```{r rotate-checks, include=FALSE}
stopifnot(
  max(abs(crossprod(rot$U_aligned, fit$K %*% rot$U_aligned) - diag(2))) < 1e-7,
  abs(rot$d - sum(diag(crossprod(B_target, fit$K %*% fit$U[, 1:2]) %*% rot$R))) < 1e-8
)
```

## Projecting New Data

```{r new-data}
new_beta <- matrix(rnorm(q * P), q, P)
projected <- dkge_project_clusters(fit, new_beta)
head(projected)
```

`dkge_project_clusters()` returns component scores for each cluster in the new
data, ready for contrast testing or transport to a reference parcellation.

## Interpreting Components

Several outputs answer different interpretive questions. `fit$weights` contains
one fitting weight per subject, not a component-specific contribution. For
component-specific participation, inspect `fit$contribs`,
`dkge_plot_subject_contrib()`, or `dkge_subject_component_projections()`.
`fit$v` concatenates subject-block cluster loadings; interpret its rows spatially
only when the blocks share a declared support or after transport to a common
space. Finally, `dkge_component_stats()` supplies component summaries under its
chosen mapper and inference assumptions. Those assumptions and the target of
inference must be reported; the function does not make every extracted
component generically “significant.”


## Where to go next

- `vignette("dkge-plotting")` — how to display saliences, subject contributions
  and subspace stability without over-reading a polished figure.
- `vignette("dkge-contrasts-inference")` — what it takes to attach uncertainty
  to a component or a contrast, rather than describing one.
- `vignette("dkge-concepts")` — which claims a fitted component supports and
  which it does not.
