---
title: "DKGE Workflow: From Subject Betas to a Shared Map"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: false
    df_print: paged

vignette: >
  %\VignetteIndexEntry{DKGE Workflow: From Subject Betas to a Shared Map}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(dkge)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
set.seed(2024)
```

Suppose each subject has a GLM beta matrix in a common effect space but a
different spatial parcellation. Your practical goal is to learn a shared
effect-space representation, evaluate a planned contrast on held-out subjects,
and place those subject-specific contrast fields on one reference
parcellation.

This page follows that one path. It uses synthetic data so the code is
self-contained; replace the beta matrices, design matrices, and centroids with
your own. If `q by P_s` and “design kernel” are not yet familiar, begin with
`vignette("dkge")`.

## What are the inputs?

We use a 2 by 3 factorial design with a planted condition and load signal. Six
subjects have different numbers of clusters, which is the reason transport is
needed later.

```{r make-workflow-data}
toy <- dkge_sim_toy(
  factors = list(condition = list(L = 2), load = list(L = 3)),
  active_terms = c("condition", "load"),
  S = 6, P = c(16, 18, 15, 17, 16, 19), snr = 4, seed = 2024
)
vapply(toy$B_list, dim, integer(2))
```

The output confirms a common number of effect rows and subject-specific
cluster counts. In real data, beta-row names must match design-column names.
`dkge_data()` validates that contract and preserves subject IDs.

```{r bundle-workflow-data}
data_bundle <- dkge_data(
  toy$B_list, designs = toy$X_list, subject_ids = toy$subject_ids
)
c(subjects = length(data_bundle$subject_ids), effects = data_bundle$q)
```

```{r bundle-workflow-checks, include=FALSE}
stopifnot(length(data_bundle$subject_ids) == 6L)
stopifnot(data_bundle$q == nrow(toy$K))
stopifnot(length(unique(vapply(data_bundle$betas, nrow, integer(1)))) == 1L)
```

**Next operation:** fit a low-rank group basis. If your subjects lack effects
or have very unequal cell precision, stop here and read
`vignette("dkge-partial-effect-spaces")` first.

## What is the baseline fit?

Start with an identity kernel. It retains the GLM/design scaling but adds no
kernel-imposed similarity between effects.

```{r fit-identity}
K_identity <- diag(data_bundle$q)
fit_identity <- dkge(data_bundle, K = K_identity, rank = 2, w_method = "none")
dkge_plot_scree(fit_identity)
```

This fit returns a two-column group basis `fit_identity$U`. The scree plot
describes variation represented by those components; it does not establish
their reliability or scientific importance.

```{r fit-identity-checks, include=FALSE}
identity_var <- dkge_variance_explained(fit_identity)
stopifnot(dim(fit_identity$U)[1] == data_bundle$q)
stopifnot(dim(fit_identity$U)[2] == 2L)
stopifnot(all(is.finite(identity_var$prop_var)))
stopifnot(sum(identity_var$prop_var) > 0)
```

**Output:** an unconstrained effect-space baseline.

**Next operation:** fit the scientifically motivated kernel and compare it
with this baseline.

## What changes when the design kernel is used?

The simulator supplies the structured kernel used to plant the signal. A real
analysis would construct this object explicitly with `design_kernel()` and
prespecify the included terms. Both fits disable subject block weighting so
the comparison isolates the kernel change.

```{r fit-structured}
fit_structured <- dkge(data_bundle, K = toy$K, rank = 2, w_method = "none")

comparison <- data.frame(
  component = 1:2,
  identity = dkge_variance_explained(fit_identity)$prop_var,
  structured = dkge_variance_explained(fit_structured)$prop_var
)
round(comparison, 3)
```

Compare the retained subspaces as well as the scree values. Because the bases
use a kernel metric, a raw `crossprod()` is not a valid rotation comparison.

```{r compare-subspaces}
angles <- dkge_principal_angles_K(
  fit_identity$U, fit_structured$U, fit_structured$K
)
round(angles * 180 / pi, 1)
```

Here component 1 accounts for `r round(100 * comparison$identity[1], 1)`% of
the retained variation under the identity kernel and
`r round(100 * comparison$structured[1], 1)`% under the structured kernel. Both
principal angles round to `r paste(round(angles * 180 / pi, 1), collapse = " and ")`
degrees: the kernel changes how variation is allocated within this seeded
two-dimensional subspace, but not the subspace itself at displayed precision.
This is a sensitivity comparison, not a test of which kernel is true.

```{r compare-subspaces-checks, include=FALSE}
stopifnot(length(angles) == 2L, all(is.finite(angles)))
stopifnot(all(angles >= 0), all(angles <= pi / 2 + 1e-8))
stopifnot(all(is.finite(unlist(comparison[-1]))))
```

**Output:** a structured fit plus an explicit record of how it differs from
the identity baseline.

**Next operation:** inspect effect saliences, then specify the contrast you
actually want to estimate.

## Which effect directions define the components?

```{r show-effect-loadings}
dkge_plot_effect_loadings(fit_structured, comps = 1:2)
```

The heatmap displays \(K U\): effect-space saliences, not voxel coefficients.
Look for coherent relative patterns within a component. Do not compare a
component's arbitrary sign across independent fits without alignment.

## How do you obtain held-out contrast fields?

Build the contrast in fitted effect order. Here the simulator identifies the
coordinate for the planted condition term.

```{r fit-loso-contrast}
condition_contrast <- numeric(data_bundle$q)
condition_contrast[toy$active_cols$condition] <- 1

condition_loso <- dkge_contrast(
  fit_structured,
  list(condition = condition_contrast),
  method = "loso"
)
condition_loso
```

The output is a named list of six subject-specific cluster vectors. LOSO
cross-fitting keeps each subject out of the basis used to score that subject.
It does not make a randomized causal contrast out of an observational design,
nor does it supply group uncertainty by itself.

```{r fit-loso-checks, include=FALSE}
loso_values <- condition_loso$values$condition
stopifnot(length(loso_values) == length(data_bundle$subject_ids))
stopifnot(identical(unname(vapply(loso_values, length, integer(1))),
                    unname(vapply(toy$B_list, ncol, integer(1)))))
stopifnot(all(vapply(loso_values, function(x) all(is.finite(x)), logical(1))))
```

**Output:** one cross-fitted contrast field per subject, still indexed by that
subject's clusters.

**Next operation:** transport those fields before stacking or performing
feature-level group inference.

## How do subject fields reach one parcellation?

Transport needs real spatial features, normally cluster centroids in a common
coordinate system. The synthetic coordinates below exist only to demonstrate
the interface; their resulting map has no anatomical interpretation.

```{r make-centroids}
centroids <- lapply(seq_along(toy$B_list), function(s) {
  p <- ncol(toy$B_list[[s]])
  cbind(x = seq_len(p), y = rep(s / 100, p), z = rep(0, p))
})
```

Use the ridge mapper for a dependency-light example and select subject 1 as the
reference parcellation:

```{r transport-contrast}
transported <- dkge_transport_contrasts_to_medoid(
  fit_structured,
  condition_loso,
  medoid = 1,
  centroids = centroids,
  method = "ridge"
)
dim(transported$condition$subj_values)
```

The returned matrix has subjects in rows and reference clusters in columns;
`transported$condition$value` is its across-subject median on the medoid
parcellation. The attached `attr(transported, "cache")` can be reused for
other contrasts or resampling.

```{r transport-checks, include=FALSE}
transported_matrix <- transported$condition$subj_values
stopifnot(nrow(transported_matrix) == length(toy$B_list))
stopifnot(ncol(transported_matrix) == ncol(toy$B_list[[1]]))
stopifnot(all(is.finite(transported_matrix)))
stopifnot(stats::sd(as.vector(transported_matrix)) > 0)
```

Transport quality depends on the features, masses, mapper, and reference
choice. Inspect mapper diagnostics and repeat defensible sensitivity analyses;
a successful matrix multiplication is not evidence that parcels are
biologically homologous.

## What should you save and report?

For a reproducible analysis, retain:

- effect names and the design matrices used to produce each beta matrix;
- the identity and structured kernels, their construction parameters, and the
  chosen rank;
- the identity-versus-structured comparison;
- cross-fitting folds or LOSO specification;
- cluster coordinates, masses, mapper settings, medoid choice, and transport
  diagnostics; and
- the inference unit and multiplicity procedure for every reported claim.

Continue with `vignette("dkge-concepts")` for the fitted moment and inference
levels, `vignette("dkge-contrasts-inference")` for uncertainty, and
`vignette("dkge-weighting")` for the distinct roles of spatial, effect,
subject, and transport weights.
