---
title: 'DKGE Concepts: What Is Estimated and What Can Be Claimed?'
description: >
  Separate what DKGE estimates from what you may claim about it, and decide
  the estimand before fitting.
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{DKGE Concepts: What Is Estimated and What Can Be Claimed?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(dkge)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
options(digits = 3)
set.seed(2024)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

After fitting DKGE, you may see a stable-looking component, a strong planned
contrast, or a localized cluster map. Those are different results. Before
interpreting any of them, you need to know which quantity DKGE decomposed, how
the design kernel changed its geometry, and which level of evidence your next
procedure can support.

This page supplies that mental model. For runnable setup and spatial transport,
start with `vignette("dkge")` and `vignette("dkge-workflow")`.

The decomposition is easiest to understand as a sequence, not as one formula:

```text
subject beta blocks -> small effect-by-effect moments -> pooled raw moment -> design/kernel geometry -> components
```

The first arrow summarizes spatial covariation within each subject. The middle
step combines subjects. Only then does the design kernel say which directions
in effect space should count as nearby or important. This ordering is why
missingness and precision must be handled before the kernel transform.

## What enters the decomposition?

For subject \(s\), let \(B_s\) be the `q` by `P_s` beta matrix. In the simplest
unweighted case, its raw effect-space moment is

\[
M_s = B_s B_s^\top.
\]

Spatial weights replace this with \(B_s \Omega_s B_s^\top\). DKGE first builds
these small `q` by `q` moments, then pools them across subjects. This is why
the core fit scales with the number of design effects rather than with a
feature-by-feature covariance matrix.

The pooled raw moment \(M\) is transformed as

\[
\widehat C = K^{1/2} R^\top M R K^{1/2},
\]

where \(R\) is the pooled design ruler and \(K\) is the design kernel. DKGE
eigendecomposes \(\widehat C\), then maps its retained eigenvectors back through
\(K^{-1/2}\) to obtain the group basis \(U\). The columns of \(U\) are
K-orthonormal: \(U^\top K U = I\).

In plain language, (M) records which effects tend to have large spatial
patterns together. (R) puts subjects' design estimates on a common ruler.
(K) defines the scientific geometry among effect directions. The eigensolve
then finds a small set of directions that summarize the transformed moment.

You can inspect each stage on a fitted object:

```{r inspect-moment-stages}
toy <- dkge_sim_toy(
  factors = list(condition = list(L = 2), load = list(L = 3)),
  active_terms = c("condition", "load"),
  S = 5, P = 20, snr = 5, seed = 2024
)
fit <- dkge(toy$B_list, toy$X_list, K = toy$K, rank = 2)

c(
  effects = nrow(fit$effect_moment),
  transformed_rows = nrow(fit$Chat),
  retained_components = ncol(fit$U)
)
```

The three matrices below make the transformation concrete. Each panel has its
own color scale because the scientific question is the pattern within a
matrix, not equality of raw numerical ranges across stages.

```{r moment-panels, echo=FALSE, fig.width=8, fig.height=3.1, fig.alt="Three heatmaps showing the pooled raw effect moment, the design kernel, and the transformed moment used for the eigendecomposition."}
plot_matrix <- function(x, title) {
  z <- x
  z <- z / max(abs(z))
  graphics::image(
    seq_len(nrow(z)), seq_len(ncol(z)), t(z[nrow(z):1, , drop = FALSE]),
    col = grDevices::hcl.colors(41, "Blue-Red 3", rev = TRUE),
    zlim = c(-1, 1), axes = FALSE, xlab = "effect", ylab = "effect",
    main = title
  )
  graphics::axis(1, at = seq_len(ncol(z)), labels = seq_len(ncol(z)), cex.axis = 0.8)
  graphics::axis(2, at = seq_len(nrow(z)), labels = rev(seq_len(nrow(z))), cex.axis = 0.8)
  graphics::box()
}
old_par <- graphics::par(mfrow = c(1, 3), mar = c(3.2, 3.2, 2.5, 0.6))
plot_matrix(fit$effect_moment, "Raw pooled moment M")
plot_matrix(fit$K, "Design kernel K")
plot_matrix(fit$Chat, "Transformed moment C-hat")
graphics::par(old_par)
```

The left panel comes from the data; the middle panel encodes the chosen design
geometry; the right panel is the matrix actually decomposed. A strong pattern
in the right panel is therefore a joint consequence of empirical covariance
and modeling choices, not a feature discovered independently of the kernel.

```{r inspect-moment-checks, include=FALSE}
stopifnot(all(dim(fit$effect_moment) == dim(toy$K)))
stopifnot(all(dim(fit$Chat) == dim(toy$K)))
stopifnot(all(is.finite(fit$effect_moment)), all(is.finite(fit$Chat)))
stopifnot(isTRUE(all.equal(
  crossprod(fit$U, fit$K %*% fit$U), diag(fit$rank), tolerance = 1e-7
)))
```

**Input:** subject effect moments.

**Output:** a low-rank basis in the kernel metric.

**Next operation:** inspect \(K U\) with `dkge_component_saliences()`, or project
a prespecified contrast with `dkge_contrast()`.

## What does the kernel change?

The kernel is a metric, not a fixed set of component directions. It changes
which effect-space directions count as large or smooth during the eigensolve;
the empirical moment still determines the fitted components.

An identity kernel is the essential baseline. It applies no kernel-imposed
coupling among effects, although the full subject-level fit still uses GLM
effect rows and, by default, the pooled design ruler \(R\). A structured kernel
regularizes the fit toward directions favored by its factor, ordinal, or block
structure.

Fit both versions on the same data:

```{r compare-kernel-fits}
fit_identity <- dkge(
  toy$B_list, toy$X_list, K = diag(nrow(toy$K)), rank = 2,
  w_method = "none"
)
fit_structured <- dkge(
  toy$B_list, toy$X_list, K = toy$K, rank = 2,
  w_method = "none"
)

data.frame(
  component = 1:2,
  identity = dkge_variance_explained(fit_identity)$prop_var,
  structured = dkge_variance_explained(fit_structured)$prop_var
)
```

Compare retained subspaces in the structured metric:

```{r compare-kernel-angles}
angles <- dkge_principal_angles_K(
  fit_identity$U, fit_structured$U, fit_structured$K
)
round(angles * 180 / pi, 1)
```

```{r compare-kernel-checks, include=FALSE}
stopifnot(length(angles) == 2L, all(is.finite(angles)))
stopifnot(all(angles >= 0), all(angles <= pi / 2 + 1e-8))
```

Small angles mean the two fits retained similar subspaces under that metric;
large angles mean the structured kernel materially changed them. A raw
`crossprod(fit_identity$U, fit_structured$U)` is not a valid substitute because
the bases are not Euclidean-orthonormal in the same geometry.

Both fits disable subject block weighting, isolating the kernel change in this
comparison. Neither outcome proves the structured kernel is correct. Report
the kernel as a modeling choice and the identity comparison as a sensitivity
analysis.

## What happens before the kernel when effects are incomplete?

Coverage, effect precision, and finite-trial debiasing alter the raw pooled
moment \(M\) before the \(R\) and \(K\) transforms. They are not cosmetic weights
on already fitted components.

This ordering matters:

1. mark which effect rows each subject actually observed;
2. construct subject moments, optionally subtracting analytic noise or using a
   split-half cross-moment;
3. pool observed pairs with the selected effect-precision and missingness
   policy; and
4. apply \(R\), \(K\), and the eigensolve.

An absent cell is therefore not an observed zero. `dkge_data()` zero-fills
internal aligned matrices only after recording observation masks, and the fit
uses those masks when pooling. Policies such as `missingness = "rescale"` or
`"shrink"` change the estimand; they must be chosen and reported, not used as
silent repairs.

See `vignette("dkge-partial-effect-spaces")` for the coverage contract,
`vignette("dkge-unbalanced-trialwise")` for the runnable workflow, and
`vignette("dkge-weighting")` for the distinction among effect, subject,
spatial, and transport weights.

## Which claim does each analysis support?

DKGE results live at several levels. Match the tool and inference unit to the
claim rather than treating “significant DKGE” as a single outcome.

| Level | Question | Typical tool | What the result does not establish |
|---|---|---|---|
| Component diagnostics | How much fitted variation is retained, and how sensitive is the subspace to refitting choices? | eigenspectrum, identity comparison, `dkge_plot_subspace_stability()` | inferential reliability or that a particular contrast is non-zero |
| Aggregate component inference | Is a prespecified aggregate component statistic unusual under its resampling null? | `dkge_aggregate_permute()`, `dkge_aggregate_bootstrap()` | validity for the subject-wise q-space fit or an untested contrast |
| Contrast | Does a prespecified effect-space contrast produce a reproducible field? | `dkge_contrast(method = "loso")`, bootstrap or analytic contrast inference | spatial localization without transport and multiplicity control |
| Feature | Where is a component or contrast expressed after alignment? | `dkge_component_stats()`, `dkge_transport_contrasts_to_medoid()` | that the latent direction was selected without bias |
| Between-subject term | Is a subject-level covariate associated with a named multivariate target? | `dkge_between_rrr()`, `dkge_between_permute()` | a test of the DKGE component itself, or a causal effect without identification assumptions |

For population claims, the subject is the resampling unit. Clusters, anchors,
voxels, and factorial cells within a subject are not independent subjects.
Cross-fitting protects held-out scoring from basis reuse, but it is not a
replacement for uncertainty estimation or multiple-testing correction.

A prespecified contrast does not need to pass through a component-significance
gate first. Component diagnostics, q-space contrast inference, transported
feature inference, aggregate decomposition, and between-subject term tests
answer different questions; use the branch that matches the estimand.

Likewise, a contrast can describe a controlled model comparison in
observational data without identifying a causal effect. Causal language
requires design and identification assumptions outside DKGE itself.

## How the pieces fit together

A DKGE analysis moves through four stages, and each one has a public entry
point:

1. **Harmonize** subject data with `dkge_subject()` and `dkge_data()`, so every
   subject's beta rows carry the same effect labels.
2. **Fit** a group embedding with `dkge()`, which compresses to effect space and
   solves for the shared basis under the design kernel.
3. **Cross-fit** a prespecified contrast with `dkge_contrast()`, so no subject's
   own data shapes the basis used to score it.
4. **Align** the resulting fields onto common locations with
   `dkge_prepare_transport()` and `dkge_transport_contrasts_to_medoid()`.

`dkge_pipeline()` runs all four in one call. `CONTRIBUTING.md` in the package
repository carries the implementation map for contributors: which source file
owns each stage, and the rules for changing them.

## What should you decide before fitting?

Use this sequence:

1. **Name the estimand.** Decide whether you want subject-level group structure
   from `dkge()` / `dkge_fit()` or an aggregate cell-mean decomposition from
   `dkge_aggregate_fit()`.
2. **Audit the input contract.** Verify effect names, observed-cell masks,
   cluster weights, and the subject-level unit of analysis.
3. **Fit the identity baseline.** Save its eigenspectrum and leading saliences.
4. **Add only justified kernel structure.** List every requested term in
   `design_kernel()` and document its scaling.
5. **Compare fits in the K metric.** Use
   `dkge_principal_angles_K()` and compare interpreted saliences, not raw
   Euclidean cross-products.
6. **Prespecify the claim level.** Component, contrast, feature, or
   between-subject; then use inference designed for that level.
7. **Transport only with defensible features.** Record coordinates, masses,
   mapper diagnostics, and reference choice.

The next practical pages are `vignette("dkge-design-kernels")` for kernel
construction, `vignette("dkge-contrasts-inference")` for contrast uncertainty,
and `vignette("dkge-between-subjects")` for subject-level associations.
