---
title: "Between-Subject Multivariate Modeling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Between-Subject Multivariate Modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(dkge)
set.seed(2026)
```

Suppose you have one brain-derived feature vector per subject and want to ask
whether group, trait, or covariate effects are distributed across that vector.
The between-subject DKGE layer fits one supervised multivariate model and
returns a coefficient map and a resampling test for each model term. The
features may be component scores, transported parcels, or any common-space
subject-by-feature matrix.

The workflow has three steps:

1. build a subject-by-feature target with `dkge_make_target()`
2. build a second-level design with `dkge_subject_model()`
3. fit and test a reduced-rank model with `dkge_between_rrr()` and `dkge_between_permute()`

## What does the second-level model estimate?

The between-subject model is a reduced-rank regression:

\[
Y = X B + E, \qquad \mathrm{rank}(B) \le r
\]

Here `Y` is a subject-by-feature matrix, `X` is the subject-level design matrix, and `B` contains distributed coefficient maps for your terms of interest.

## How do you fit a direct subject-by-feature matrix?

In the simplest case you already have a common-space matrix with one row per subject.

```{r quick-target}
subject_data <- data.frame(
  subject_id = paste0("s", 1:18),
  group = factor(rep(c("control", "patient"), each = 9)),
  trait = seq(-1, 1, length.out = 18)
)

design <- dkge_subject_model(~ group * trait, subject_data)
X <- design$X

beta <- matrix(0, nrow = ncol(X), ncol = 6,
               dimnames = list(colnames(X), paste0("feature", 1:6)))
beta["grouppatient:trait", ] <- c(2, 1, 0, -1, -2, -1)

Y <- X %*% beta + matrix(rnorm(nrow(X) * ncol(beta), sd = 0.08), nrow(X), ncol(beta))
rownames(Y) <- subject_data$subject_id
colnames(Y) <- colnames(beta)

target <- dkge_make_target(Y = Y)
fit_between <- dkge_between_rrr(target, design, rank = 1)
```

You can inspect the fitted term map directly.

```{r quick-term-map}
round(dkge_term_map(fit_between, "grouppatient:trait"), 2)
```

For an unweighted, single-block model with a scientifically defensible
Gaussian row-sphericity assumption, request a residual-space rotation test.
The `terms` argument controls which formula terms are tested; it does not
refit a different full model for each term.

```{r quick-permutation}
perm_between <- dkge_between_permute(
  fit_between,
  terms = c("group", "trait", "group:trait"),
  method = "rotation",
  B = 99,
  seed = 42,
  scope = "both",
  feature_adjust = "maxT"
)

perm_between$summary[, c("term", "statistic", "p", "p_adjusted")]
```

The global statistic is the reduction in residual sum of squares when the
tested term is restored to the reduced model. Both methods refit at the same
fitted rank (`object$rank`). Freedman–Lane has an extra caveat: if dropping
the tested term lowers the reduced model's estimable rank below that value,
the reduced fit is clipped while the full fit is not, so the statistic also
absorbs that rank difference. Rotation does not clip — the reduced design
stays in `Q0` and the Haar draw lives in its complement. P-values use the
finite-resample correction `(1 + exceedances) / (B + 1)`. With
`scope = "both"`, the result also contains featurewise tests;
`feature_adjust = "maxT"` controls multiplicity against the largest feature
statistic in each resample.

```{r quick-permutation-checks, include = FALSE}
stopifnot(all(is.finite(fitted(fit_between))))
stopifnot(all(is.finite(residuals(fit_between))))
stopifnot(all(perm_between$summary$p >= 0 & perm_between$summary$p <= 1))
stopifnot(all(perm_between$feature_tests[["group:trait"]]$p_adjusted >= 0))
stopifnot(all(perm_between$feature_tests[["group:trait"]]$p_adjusted <= 1))
```

## What is residual-space rotation doing?

For a tested term, let \(X_0\) be the design with that term removed. Write
\(Q_0\) for an orthonormal basis of the nuisance space spanned by \(X_0\), and
\(Q_e\) for its orthogonal complement. A null replicate is

\[
Y^* = Q_0 Q_0^\top Y + Q_e G Q_e^\top Y,
\]

where \(G\) is a random orthogonal matrix. The first term fixes the fitted
nuisance signal. The second rotates only the residual coordinates. Because
the transformation is orthogonal, it preserves `crossprod(Y)` exactly and
retains arbitrary covariance among features.

This produces an exact finite-sample randomization test under a matrix-normal
null with independent, homoscedastic subject rows. It is only approximate for
heteroscedastic, clustered, or non-Gaussian subject errors. The result records
that contract rather than asking you to remember it:

```{r rotation-contract}
unlist(perm_between$resampling_assumptions)
```

The implementation evaluates each replicate from compressed design-by-feature
cross-products. It forms a square rotation only in the residual subject
dimension and never constructs a feature-by-feature covariance matrix.

## Which resampling method should you choose?

Neither method is a universal confirmatory default.

| Situation | Method | Interpretation |
|---|---|---|
| Unweighted fit, one block, Gaussian row-sphericity is plausible | `method = "rotation"` | Sharpest exactness contract; may lose power in small samples. |
| Subject/feature weights or nontrivial exchangeability blocks are required | `method = "freedman_lane"` | Supported legacy approximation; small-sample calibration is imperfect. |
| Heteroscedasticity or dependence is scientifically central | Neither automatically | Treat the result as exploratory or justify a design-specific resampling model. |

`"freedman_lane"` remains the compatibility default, so use `method =` in
analysis scripts to make the inferential choice visible. Missing block labels
(`NA` in `blocks`) are rejected for both methods. Rotation additionally fails
before resampling when weights, multiple blocks, or fewer than two residual
dimensions put the fit outside its supported scope.

### What did the frozen calibration show?

The tables below are generated from the machine-readable artifacts shipped
with the package. The null rows report rejection frequency at nominal 0.05 for
20 subjects and 399 resamples.

```{r calibration-data, include = FALSE}
rotation_calibration <- read.csv(
  system.file("extdata", "dkge-between-rotation-summary.csv", package = "dkge"),
  stringsAsFactors = FALSE
)
legacy_calibration <- read.csv(
  system.file("extdata", "dkge-null-calibration-summary.csv", package = "dkge"),
  stringsAsFactors = FALSE
)
term_order <- c("group", "trait", "group:trait")
rotation_global <- subset(rotation_calibration, arm == "gaussian_global_null")
rotation_partial <- subset(rotation_calibration, arm == "gaussian_partial_null")
legacy_primary <- subset(legacy_calibration,
                         family == "between_freedman_lane" &
                         setting == "primary_n20")
rotation_global <- rotation_global[match(term_order, rotation_global$term), ]
rotation_partial <- rotation_partial[match(term_order, rotation_partial$term), ]
legacy_primary <- legacy_primary[match(term_order, legacy_primary$estimand), ]
calibration_table <- data.frame(
  term = term_order,
  rotation_global = rotation_global$rejection_005,
  rotation_with_nuisance = rotation_partial$rejection_005,
  freedman_lane_global = legacy_primary$empirical_size_0.05,
  rotation_gate = ifelse(rotation_global$gate_pass & rotation_partial$gate_pass,
                         "pass", "fail")
)
power_table <- subset(rotation_calibration,
                      arm == "strong_interaction_power",
                      c("method", "rejection_005", "wilson_lower", "wilson_upper"))
rownames(power_table) <- NULL
stopifnot(nrow(calibration_table) == 3L)
stopifnot(all(rotation_global$gate_pass))
stopifnot(identical(rotation_partial$gate_pass, c(TRUE, TRUE, FALSE)))
stopifnot(power_table$rejection_005[power_table$method == "rotation"] == 0.43)
```

```{r calibration-table, echo = FALSE}
knitr::kable(
  calibration_table,
  digits = 3,
  col.names = c("Term", "Rotation: global null",
                "Rotation: nuisance null", "Freedman-Lane: global null",
                "Rotation gate")
)
```

Rotation removed the material global-null inflation seen for Freedman-Lane,
but the nuisance-interaction arm rejected 33 of 500 nulls (0.066), one result
above the frozen 0.065 promotion ceiling. Its Wilson interval still contained
0.05 and its other uniformity diagnostics passed. The predeclared rule did not
permit replacing that seed family or relaxing the threshold.

```{r power-table, echo = FALSE}
knitr::kable(
  power_table,
  digits = 3,
  col.names = c("Method", "Detection rate", "Wilson lower", "Wilson upper")
)
```

In the frozen strong-interaction simulation (shared DGP, power arm `B = 199`),
rotation detected 43% and Freedman-Lane 58%. That raw gap is mostly
Freedman-Lane size inflation: at matched conditions rotation's size is 0.050
and Freedman-Lane's is 0.086, and size-matched Freedman-Lane power is about
0.49. The predeclared absolute and relative gates still fail, so rotation
remains opt-in. Increasing `B` improves Monte Carlo resolution; it does not
repair either method's assumptions or recover the unadjusted power gap.

## How do you build the target from a DKGE fit?

The more DKGE-native path is to let DKGE construct the subject representation first, then model subjects in that common target space.

This example starts from heterogeneous subject maps, fits DKGE, transports one contrast to a medoid subject, and then applies the between-subject model.

```{r dkge-fit}
toy <- dkge_sim_toy(
  factors = list(condition = list(L = 2), phase = list(L = 2)),
  active_terms = c("condition", "phase"),
  S = 6,
  P = c(4, 5, 3, 6, 4, 5),
  snr = 6,
  seed = 9
)

effects <- toy$info$term_names
toy$B_list <- lapply(toy$B_list, function(B) {
  rownames(B) <- effects
  B
})
toy$X_list <- lapply(toy$X_list, function(X) {
  colnames(X) <- effects
  X
})
centroids <- lapply(toy$B_list, function(B) matrix(rnorm(ncol(B) * 3), ncol(B), 3))

fit_dkge <- dkge_fit(
  dkge_data(toy$B_list, designs = toy$X_list, subject_ids = toy$subject_ids),
  K = toy$K,
  rank = 2
)
```

```{r transported-target}
transport <- dkge_transport_spec(
  centroids = centroids,
  medoid = 1L,
  method = "sinkhorn",
  epsilon = 0.1
)

contrast <- matrix(c(1, 0, 0), ncol = 1, dimnames = list(NULL, "condition"))
rownames(contrast) <- effects

transported_target <- dkge_make_target(
  fit_dkge,
  type = "transported_maps",
  contrast = contrast,
  transport = transport,
  crossfit = "analytic"
)

dim(transported_target$Y)
```

Now the second-level model looks the same as before.

```{r dkge-between}
subject_meta <- data.frame(
  subject_id = toy$subject_ids,
  group = factor(c("control", "control", "control", "patient", "patient", "patient")),
  trait = c(-0.8, -0.3, 0.2, -0.1, 0.5, 1.0)
)

between_design <- dkge_subject_model(~ group * trait, subject_meta)
fit_transport <- dkge_between_rrr(transported_target, between_design, rank = 1)
perm_transport <- dkge_between_permute(
  fit_transport,
  terms = c("group", "trait", "group:trait"),
  method = "rotation",
  B = 49,
  seed = 7
)

perm_transport$summary[, c("term", "p")]
```

```{r dkge-between-checks, include = FALSE}
stopifnot(nrow(transported_target$Y) == nrow(subject_meta))
stopifnot(all(is.finite(predict(fit_transport))))
stopifnot(all(perm_transport$summary$p >= 0 & perm_transport$summary$p <= 1))
```

## What should you inspect after fitting?

There are three outputs worth checking first:

- `coef(fit_transport)` gives the full coefficient matrix in target space
- `dkge_term_map(fit_transport, "group:trait")` extracts the term-specific map
- `perm_transport$summary` gives a global multivariate resampling test for each term

If you asked for featurewise inference, `perm_between$feature_tests[["group:trait"]]` adds per-feature statistics and adjusted p-values in the same target space.

Report the target construction, fitted rank, tested term, resampling method,
number of resamples, exchangeability assumptions, and multiplicity adjustment.
Do not describe a term-level p-value as evidence that every feature in its map
is individually significant.

## Where to go next

Use `type = "component_scores"` in `dkge_make_target()` when you want a compact
latent target, and use `type = "transported_maps"` when spatial interpretation
in a common map space matters. See `vignette("dkge-contrasts-inference")` for within-subject contrast inference.
