---
title: Getting Started with DKGE
description: >
  Fit a first DKGE model from per-subject GLM betas, read its components,
  and find the page that answers your next question.
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Getting Started with DKGE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
set.seed(42)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

```{r load-dkge}
library(dkge)
```

You have one GLM beta matrix per subject. Its rows are the same named design
effects, but its columns may be different subject-specific clusters. You want
to find effect-space directions that recur across subjects, inspect what those
directions express, and evaluate a planned contrast without constructing a
voxel-by-voxel group covariance matrix.

DKGE takes those beta matrices, their design matrices, and a design kernel. It
returns a low-dimensional group basis in effect space plus subject-level
cluster values that can be cross-fitted and, when coordinates are available,
transported to a common parcellation.

The reason for the extra machinery is the mismatch between what is common and
what is not. Experimental effects have the same meaning across subjects, but
their clusters often do not line up. DKGE learns the shared representation in
the small, common effect space first; spatial alignment is postponed until you
have a scientifically named component or contrast to map.

```text
subject beta matrices       shared effect space       subject fields       common map
  q x P1, q x P2, ...   ->     q x rank basis      ->  P1, P2, ...    ->  reference P
  different clusters          same effects              cross-fitted          transported
```

On this page you will make one fit, read its leading components, and obtain a
held-out contrast field. That is the shortest useful DKGE path. The spatial
transport step appears on the next page because it requires coordinates and a
separate alignment assumption.

## What does the first fit accomplish?

The package includes a simulator with known factorial structure. Here the
planted signal involves two main effects in a 2 by 3 design. Each subject has a
5 by 20 beta matrix: five effect rows and twenty clusters.

```{r simulate-example}
toy <- dkge_sim_toy(
  factors = list(condition = list(L = 2), load = list(L = 3)),
  active_terms = c("condition", "load"),
  S = 5, P = 20, snr = 5, seed = 42
)
dim(toy$B_list[[1]])
```

The three objects needed for a first fit are `toy$B_list` (subject beta
matrices), `toy$X_list` (matching design rulers), and `toy$K` (the effect-space
kernel). Fit two components:

```{r fit-example, fig.alt="Scree plot showing the proportion and cumulative share of fitted variation for two DKGE components."}
fit <- dkge(toy$B_list, toy$X_list, K = toy$K, rank = 2)
dkge_plot_scree(fit)
```

The bars show each retained component's share of the fitted q-space variation;
the line is cumulative. This is a descriptive summary of the fitted latent
space, not a significance test.

```{r fit-example-checks, include=FALSE}
variance_table <- dkge_variance_explained(fit)
stopifnot(nrow(fit$U) == nrow(toy$K), ncol(fit$U) == 2L)
stopifnot(all(is.finite(variance_table$prop_var)))
stopifnot(all(variance_table$prop_var >= 0))
stopifnot(sum(variance_table$prop_var) > 0)
```

## What does a component express?

Inspect the design-weighted saliences, \(K U\), before mapping anything back to
space. Rows are effects and columns are components: a salience says how much a
named design effect contributes to a component, in the kernel's metric. See
`?dkge-glossary` for this and the other terms the documentation leans on.

```{r inspect-saliences}
round(dkge_component_saliences(fit, comps = 1:2, long = FALSE), 2)
```

Large positive and negative entries identify the effect directions associated
with each component. Component signs are arbitrary, so interpret relative
patterns rather than treating the sign itself as scientifically meaningful.

The same values are easier to scan as a heatmap:

```{r salience-heatmap, fig.alt="Heatmap of design-effect saliences for the first two DKGE components."}
dkge_plot_effect_loadings(fit, comps = 1:2)
```

Read down a column. Effects with similar color and magnitude move together in
that component; effects with opposite colors define a contrast-like pattern.
The heatmap tells you *what the component combines*. It does not yet tell you
where that pattern lies in the brain or whether it is reliable in a population.

**Next operation:** use `dkge_plot_effect_loadings(fit)` for a heatmap, or move
to a prespecified effect contrast.

## How do you evaluate a contrast without reusing each subject's data?

Construct a contrast in the same effect order as the fit. The simulator records
which effect coordinate carries the planted `condition` term.

```{r loso-contrast}
condition_contrast <- numeric(nrow(toy$K))
condition_contrast[toy$active_cols$condition] <- 1

condition_loso <- dkge_contrast(
  fit, list(condition = condition_contrast), method = "loso"
)
condition_loso
```

`method = "loso"` refits the latent basis without each held-out subject before
computing that subject's cluster values. The result contains one vector per
subject under `condition_loso$values$condition`; those vectors still live on
the subjects' own cluster systems.

```{r loso-contrast-checks, include=FALSE}
condition_values <- condition_loso$values$condition
stopifnot(length(condition_values) == length(toy$B_list))
stopifnot(all(vapply(condition_values, function(x) all(is.finite(x)), logical(1))))
stopifnot(any(vapply(condition_values, function(x) stats::sd(x) > 0, logical(1))))
```

Cross-fitting limits basis-reuse optimism; it does not by itself provide a
population p-value, spatial multiplicity correction, or causal interpretation.
Those require an inference procedure matched to the claim and an experimental
design that supports the interpretation.

At this point the object journey has reached the third box in the diagram:
`condition_loso` holds one field per subject, but the field lengths and cluster
identities can still differ. The next step is therefore alignment, not averaging.

**Next operation:** if clusters differ across subjects, transport the contrast
with `dkge_transport_contrasts_to_medoid()`. If they already share a common
index, `as.matrix(condition_loso)` stacks them directly.

## What must your real data look like?

For subject \(s\), supply:

- a numeric `q` by `P_s` beta matrix, where rows are named effects and columns
  are clusters, parcels, or voxels;
- a numeric `T_s` by `q` design matrix whose column names match the beta-row
  names; and
- optionally, cluster weights, coordinates, effect counts, or trial-level
  sufficient statistics needed by later stages.

Wrap subjects with `dkge_subject()` when you want validation and provenance,
then combine them with `dkge_data()`. Subjects may have different `P_s`. If
some effects are unobserved, declare that explicitly; DKGE records coverage and
does not treat an absent effect as an observed zero.

```{r bundle-example}
subjects <- Map(
  function(beta, design, id) dkge_subject(beta, design = design, id = id),
  toy$B_list, toy$X_list, toy$subject_ids
)
bundle <- dkge_data(subjects)
c(subjects = length(bundle$subject_ids), effects = bundle$q)
```

For unequal trial counts or missing design cells, continue with
`vignette("dkge-partial-effect-spaces")` before fitting.

## Which page should you read next?

The default path is:

1. `vignette("dkge-workflow")` — take the same objects through fit, comparison,
   cross-fitted contrast, and spatial transport.
2. `vignette("dkge-concepts")` — understand the kernel metric, the fitted
   moment, and the distinct levels of inference.
3. `vignette("dkge-design-kernels")` — construct a kernel for your own design.
4. `vignette("dkge-contrasts-inference")` — choose uncertainty and testing
   machinery for a prespecified contrast.

Common branches are `vignette("dkge-partial-effect-spaces")` for incomplete or
unequally estimated effects, `vignette("dkge-weighting")` for weighting
estimands, and `vignette("dkge-plotting")` for presentation.

For function-level help, start with `?dkge`, `?dkge_subject`,
`?design_kernel`, and `?dkge_contrast`.
