---
title: Unbalanced Trialwise Designs
description: >
  Fit an unbalanced 3 x 5 x 4 trialwise design with count weighting, analytic
  debiasing, and a split-half alternative.
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Unbalanced Trialwise Designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(dkge)
set.seed(2026)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

DKGE can represent one global effect space when subjects contribute different
cells or estimate the same cells with very different trial counts. The output
is still a common q-dimensional group basis, but three distinct problems must
be handled separately:

This page is for the case where the usual rectangular data picture is false.
Perhaps controls and patients occupy different rows of a global grid, or every
subject attempted the same factorial design but some cells have many more
trials than others. If every subject has every effect with comparable
precision, you do not need these controls; use `vignette("dkge-workflow")`.

The guiding rule is simple: **absence, imprecision, and estimation noise are
not synonyms**. The rest of the vignette shows where each enters the fit.

| Problem | What changes | DKGE mechanism |
|---|---|---|
| A cell is absent for a subject | Whether an effect pair is observed | `observed_rows` and `missingness` |
| A cell has 2 trials for one subject and 20 for another | Precision of an observed effect | `dkge_effect_weights()` |
| Cell estimates contain finite-trial estimation noise | Expected diagonal noise in the second moment | `debias` |

Coverage, precision weighting, and debiasing are complementary. A count weight
does not remove estimation-noise bias; analytic subtraction does not decide
which subject should contribute more; and zero-filling alone does not record
that a cell was unobserved.

The key contract is:

- `dkge_data()` aligns subject-local effect rows to the union of all effect
  labels.
- Missing rows are rendered as zero rows in the aligned matrices.
- The original observation pattern is retained on the returned bundle as
  `observed_rows`, plus `obs_mask` and `pair_counts` under `provenance`.
- `dkge_fit(missingness = ...)` controls how the partial coverage is handled
  when the q-space covariance is accumulated.

```text
local observed rows -> global labelled grid -> coverage-aware raw moment -> precision/debiasing -> kernel transform
```

You have a trialwise design where cells are observed unequally: some
subject-by-cell combinations carry two trials and others carry twenty. This
page fits one such design end to end, a 3 x 5 x 4 grid over 60 cells, and shows
what count weighting and debiasing each change.

It assumes the contract from `vignette("dkge-partial-effect-spaces")`: how a
global grid is declared, how subject-local rows are aligned to it, and how
coverage enters the fit. Read that first if you have not.

## How do you fit an unbalanced 3 x 5 x 4 trialwise design?

Now consider a fully within-subject `condition x delay x response` design with
3, 5, and 4 levels: 60 possible cells per subject. The response is ordinal, and
trial counts vary by subject and cell. `dkge_effect_grid()` pins the global row
order while `design_kernel()` says that adjacent response levels are more
similar than distant ones.

```{r trial-grid}
grid60 <- dkge_effect_grid(
  factors = list(
    condition = c("c1", "c2", "c3"),
    delay = paste0("d", 1:5),
    response = list(L = 4, type = "ordinal",
                    levels = as.character(1:4), l = 1)
  )
)

kernel60 <- design_kernel(
  grid60,
  terms = list(
    "condition", "delay", "response",
    c("condition", "delay"), c("condition", "response"),
    c("delay", "response"), c("condition", "delay", "response")
  ),
  basis = "cell",
  normalize = "unit_trace"
)

c(q = length(grid60$cell_labels), kernel_rows = nrow(kernel60$K))
```

```{r trial-grid-checks, include = FALSE}
stopifnot(length(grid60$cell_labels) == 60L)
stopifnot(identical(dim(kernel60$K), c(60L, 60L)))
```

The trialwise constructor fits \(Y_s = X_s B_s + E_s\) for each subject. It
retains \(B_s\), \((X_s^\top X_s)^{-1}\), residual variances, and one-hot cell
counts, but not the full trial-by-feature response. Here every cell has at least
three trials so that the within-cell split leaves both halves estimable.

```{r trialwise-data, include = FALSE}
q <- length(grid60$cell_labels)
P <- 10L
condition_score <- c(-1, 0, 1)[match(grid60$cells$condition,
                                     c("c1", "c2", "c3"))]
delay_score <- seq(-1, 1, length.out = 5)[match(grid60$cells$delay,
                                                paste0("d", 1:5))]
response_score <- as.numeric(grid60$cells$response) - 2.5
latent_effects <- cbind(condition_score, delay_score, response_score,
                        condition_score * response_score)
set.seed(810)
feature_patterns <- matrix(rnorm(ncol(latent_effects) * P),
                           ncol(latent_effects), P)
true_beta <- latent_effects %*% feature_patterns / 3
colnames(true_beta) <- paste0("feature", seq_len(P))
rownames(true_beta) <- grid60$cell_labels

make_trial_subject <- function(subject_index) {
  set.seed(810 + subject_index)
  counts <- sample(3:10, q, replace = TRUE)
  cell_index <- rep(seq_len(q), counts)
  X <- matrix(0, length(cell_index), q)
  X[cbind(seq_along(cell_index), cell_index)] <- 1
  colnames(X) <- grid60$cell_labels
  Y <- X %*% true_beta + matrix(rnorm(length(cell_index) * P, sd = 0.8),
                                length(cell_index), P)
  colnames(Y) <- colnames(true_beta)
  dkge_trial_subject(Y, X, id = paste0("s", subject_index),
                     split = "within_cell")
}

trial_subjects <- lapply(seq_len(5), make_trial_subject)
bundle_trialwise <- dkge_data(trial_subjects, effects = grid60$cell_labels)
count_table <- do.call(rbind, lapply(trial_subjects, function(subject) {
  c(total_trials = sum(subject$effect_n),
    min_cell = min(subject$effect_n),
    median_cell = stats::median(subject$effect_n),
    max_cell = max(subject$effect_n))
}))
rownames(count_table) <- paste0("s", seq_len(nrow(count_table)))
```

```{r trial-counts, echo = FALSE}
knitr::kable(count_table, digits = 1)
```

```{r trial-counts-plot, echo=FALSE, fig.width=7, fig.height=3.8, fig.alt="Boxplots and points showing unequal trial counts across sixty cells for five subjects."}
trial_count_long <- do.call(rbind, lapply(seq_along(trial_subjects), function(i) {
  data.frame(subject = paste0("s", i), trials = trial_subjects[[i]]$effect_n)
}))
ggplot2::ggplot(trial_count_long, ggplot2::aes(subject, trials, colour = subject)) +
  ggplot2::geom_boxplot(outlier.shape = NA, width = 0.55, alpha = 0.15) +
  ggplot2::geom_jitter(width = 0.16, height = 0, alpha = 0.35, size = 1.2) +
  ggplot2::scale_y_continuous(breaks = 3:10) +
  ggplot2::labs(
    title = "Observed does not mean equally precise",
    subtitle = "Each point is one condition × delay × response cell",
    x = "Subject", y = "Trials in cell"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "none")
```

Each subject contributes a different count profile, but all are aligned to the
same 60 labels. The group fit below makes three choices explicit:

1. `effect_scaling = "none"` keeps cell means in their common beta units rather
   than applying the pooled design ruler.
2. `effect_weights = dkge_effect_weights("count")` gives more influence to
   better-estimated subject-by-cell rows.
3. `debias = "analytic"` subtracts the expected finite-trial noise moment before
   pooling.

`w_method = "none"` is deliberate in this diagnostic example: it isolates
cell-level precision weighting and debiasing from the package's default
subject-level MFA scaling. It is not a general recommendation to disable
subject weighting.

```{r trial-fit}
fit_analytic <- dkge_fit(
  bundle_trialwise,
  K = kernel60,
  rank = 3,
  w_method = "none",
  effect_scaling = "none",
  effect_weights = dkge_effect_weights("count"),
  debias = "analytic",
  missingness = "none"
)

fit_analytic$rank
```

For cells \(c\) and \(c'\), count weighting uses pair reliability
\(\sqrt{n_{sc} n_{sc'}}\). DKGE first computes the precision-weighted mean for
each pair, then restores the cohort scale. `fit_analytic$pair_ess` is Kish's
effective number of contributing subjects; it falls when one or two subjects
dominate a cell pair even if every subject observed it.

```{r trial-diagnostics}
pair_diagnostics <- data.frame(
  diagnostic = c("minimum pair ESS", "maximum pair ESS",
                 "negative raw-effect mass", "negative transformed mass"),
  value = c(min(fit_analytic$pair_ess), max(fit_analytic$pair_ess),
            fit_analytic$moment_diagnostics$effect$negative_mass,
            fit_analytic$moment_diagnostics$transformed$negative_mass)
)
knitr::kable(pair_diagnostics, digits = 3)
```

```{r trial-fit-checks, include = FALSE}
stopifnot(identical(dim(fit_analytic$effect_moment), c(60L, 60L)))
stopifnot(all(is.finite(fit_analytic$effect_moment)))
stopifnot(all(is.finite(fit_analytic$pair_ess)))
stopifnot(min(fit_analytic$pair_ess) > 1)
stopifnot(max(fit_analytic$pair_ess) <= length(trial_subjects) + 1e-10)
stopifnot(all(is.finite(unlist(fit_analytic$moment_diagnostics))))
stopifnot(pair_diagnostics$value[1] < pair_diagnostics$value[2])
```

### What exactly does debiasing change?

Finite-trial noise can be addressed in either of two ways:

- `debias = "analytic"` subtracts
  `noise_trace * (X'X)^{-1}` per subject. The noise trace includes residual
  variance and any diagonal spatial weights.
- `debias = "split_half"` replaces the raw second moment with the symmetrized
  cross-product of two stored half estimates. Independent half-errors then have
  zero expected cross-product.

The chunked constructor uses the same weighted analytic correction as the
dense constructor. In particular, a non-unit spatial `omega` is included when
the noise trace is reconstructed from per-feature residual variances; an
unweighted cached trace is not reused as if it were already weighted.

```{r chunked-trace}
set.seed(19411)
X_chunk <- model.matrix(~ 0 + factor(rep(1:2, each = 6)))
colnames(X_chunk) <- c("e1", "e2")
omega_chunk <- c(0.2, 1, 3, 0.5)
make_chunk_y <- function() {
  truth <- matrix(c(1, -0.5, 0.25, 2, 0.4, -1, 0.7, 0.1), 2, 4)
  Y <- X_chunk %*% truth +
    matrix(rnorm(nrow(X_chunk) * 4, sd = 0.35), nrow(X_chunk), 4)
  colnames(Y) <- paste0("v", 1:4)
  Y
}
Y_chunk <- list(make_chunk_y(), make_chunk_y())
dense_subjects <- lapply(seq_along(Y_chunk), function(s) {
  dkge_trial_subject(Y_chunk[[s]], X_chunk, id = paste0("s", s),
                     omega = omega_chunk)
})
chunked_subjects <- lapply(seq_along(Y_chunk), function(s) {
  Y <- Y_chunk[[s]]
  dkge_trial_subject_chunks(
    list(Y[, 1:2, drop = FALSE], Y[, 3:4, drop = FALSE]),
    X_chunk,
    id = paste0("s", s),
    omega = omega_chunk
  )
})
K_chunk <- diag(2)
dimnames(K_chunk) <- list(colnames(X_chunk), colnames(X_chunk))
fit_dense <- dkge_fit(
  dkge_data(dense_subjects), K = K_chunk, rank = 1,
  w_method = "none", effect_scaling = "none", debias = "analytic"
)
fit_chunked <- dkge_fit(
  dkge_data(chunked_subjects), K = K_chunk, rank = 1,
  w_method = "none", effect_scaling = "none", debias = "analytic"
)
chunked_check <- data.frame(
  weighted_noise_trace = sum(
    omega_chunk * dense_subjects[[1]]$residual_variance
  ),
  max_abs_Chat_difference = max(abs(fit_dense$Chat - fit_chunked$Chat))
)
knitr::kable(chunked_check, digits = 12)
```

```{r chunked-trace-checks, include = FALSE}
expected_trace <- sum(omega_chunk * dense_subjects[[1]]$residual_variance)
stopifnot(isTRUE(all.equal(
  chunked_check$weighted_noise_trace,
  expected_trace,
  tolerance = 1e-12
)))
stopifnot(max(abs(fit_dense$Chat - fit_chunked$Chat)) < 1e-12)
```

Because the subjects above stored within-cell halves, the alternative fit is
runnable with the same data:

```{r split-fit}
fit_split <- dkge_fit(
  bundle_trialwise,
  K = kernel60,
  rank = 3,
  w_method = "none",
  effect_scaling = "none",
  effect_weights = dkge_effect_weights("count"),
  debias = "split_half"
)

c(analytic = fit_analytic$moment_diagnostics$effect$negative_mass,
  split_half = fit_split$moment_diagnostics$effect$negative_mass)
```

Here the split-half estimate has more negative spectral mass (about 193 versus
146), but that ordering is not a performance score. Both numbers diagnose the
finite-sample indefiniteness of their respective moment estimators; choosing
between them depends on whether the split errors are credibly independent and
whether the analytic covariance model is credible.

```{r split-fit-checks, include = FALSE}
negative_mass <- c(
  analytic = fit_analytic$moment_diagnostics$effect$negative_mass,
  split_half = fit_split$moment_diagnostics$effect$negative_mass
)
stopifnot(identical(unname(round(negative_mass)), c(146, 193)))
stopifnot(negative_mass[["split_half"]] > negative_mass[["analytic"]])
```

Analytic subtraction and pair normalization can produce an indefinite
q-by-q estimate. DKGE therefore uses a symmetric eigendecomposition, retains
the leading positive eigenpairs, and exposes negative spectral mass through
each fit's `moment_diagnostics`; an SVD would incorrectly turn negative
directions into positive components.

The constructor's `split = "within_cell"` alternates trials within each cell;
it does not prove that the two half-errors are independent. If runs, sessions,
or temporal dependence define independence in your experiment, use an
appropriately constructed split outside this convenience path or prefer the
analytic estimator with a justified covariance model.

## Where to go next

- `vignette("dkge-partial-effect-spaces")` defines the coverage and
  estimability contract this page assumes, including which contrasts are
  estimable within subject.
- `vignette("dkge-weighting")` separates the effect, subject, spatial, and
  transport weighting layers.
