---
title: CPCA Filtering
description: >
  Split the compressed covariance into design-aligned and residual parts
  before eigendecomposition.
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{CPCA Filtering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(dkge)
library(ggplot2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
options(
  tibble.print_min = 4,
  tibble.print_max = 4,
  digits = 3
)
set.seed(2025)
if ("theme_dkge" %in% ls("package:dkge")) {
  theme_set(dkge::theme_dkge())
}
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

Use CPCA filtering when you have already decided that one prespecified span of
effects should be studied separately from its complement. A common example is
to ask for components aligned with main effects while keeping interaction and
control directions in a residual subspace. Without that split, the leading
DKGE component may mix all of them simply because they covary strongly.

CPCA does **not** discover which scientific partition is correct. You supply
the partition; DKGE projects the compressed moment into design-aligned and
residual parts before fitting their bases. The two parts are
\(K\)-orthogonal, which is a geometric statement, not evidence that the
underlying cognitive processes are independent.

```text
mixed compressed moment -> declared effect span + K-orthogonal complement -> separate latent bases
```

This page first shows the default mixed fit, then applies one transparent split
and visualizes what moved. Custom bases, kernel sensitivity, and ridge
stabilization follow as advanced branches.

## When is the split scientifically useful?

Use CPCA filtering when a design carries several kinds of task effect you want
analysed separately. It focuses the latent space on one effect type and
separates the rest, instead of fitting all effects together.

Common reasons to declare a split include:

**Factorial designs**: separate main effects from interactions, so a component
that combines factors cannot be mistaken for one that does not.

**Experimental versus control conditions**: focus on the manipulation while
factoring out baseline structure.

**Multi-domain studies**: separate cognitive processes measured in the same
experiment.

**Planned versus exploratory contrasts**: isolate prespecified hypotheses from
secondary analyses.

**Comparative studies**: Apply the same declared effect spans across datasets,
which makes the fitted targets easier to compare. Reliability still has to be
measured rather than inferred from the decomposition.

## Where does filtering enter the fit?

The compressed covariance \(\hat C\) mixes variance from every experimental
effect. CPCA splits that total into subcomponents before the
eigendecomposition, preserving the K-metric algebra at each step.

`dkge_projector_K()` builds a projector onto the named effect subspace in the
\(K\) metric, so the split respects the effect similarities the kernel
encodes. `dkge_cpca_split_chat()` applies that projector, separating the
compressed covariance into design and residual parts. DKGE then fits a basis
for each requested part, \(K\)-orthogonal to the other so the two can be read
together.

Name the design-aligned subspace with `cpca_blocks`, or supply an explicit
basis matrix with `cpca_T`. `cpca_part` selects what comes back (`"design"`,
`"resid"`, or `"both"`), and `cpca_ridge` adds diagonal regularisation when
the filtered covariance is near-singular.

## Simulated Experiment: Attention-Working Memory Study

We simulate a factorial attention-working memory experiment: two main effects
(attention cue validity and working memory load), their interaction, and
additional control conditions. That structure suits CPCA, where the goal is to
isolate the main experimental effects from their interaction and the control
conditions.

The simulated data carries strong signals in the primary manipulations
(attention and working memory) and weaker structure in the secondary effects,
so the split has something to separate.

```{r simulate-data}
S <- 8
q <- 6
P <- 16
Tlen <- 80
effects <- c("attn_valid", "attn_invalid", "wmem_high", "wmem_low", "interact", "control")

betas <- replicate(S, {
  # Strong signals for main experimental effects
  main_effects <- matrix(rnorm(2 * P, sd = 1.5), 2, P)
  # Weaker signals for interaction and control conditions
  secondary_effects <- matrix(rnorm((q - 2) * P, sd = 0.4), q - 2, P)
  mat <- rbind(main_effects, secondary_effects)
  rownames(mat) <- effects
  mat
}, simplify = FALSE)

designs <- replicate(S, {
  X <- matrix(rnorm(Tlen * q), Tlen, q)
  X <- qr.Q(qr(X))
  colnames(X) <- effects
  X
}, simplify = FALSE)

subjects <- Map(function(b, X, id) dkge_subject(b, X, id = id),
                betas, designs, paste0("sub", seq_len(S)))
bundle <- dkge_data(subjects)
```

Without CPCA filtering, all task-related variance is compressed into one
latent space and the leading components mix design and residual structure.

```{r fit-standard}
fit_plain <- dkge(bundle, K = diag(q), rank = 3)
round(fit_plain$evals[1:4], 3)
```

These eigenvalues represent the mixed signal from all experimental conditions.
The standard fit captures the dominant patterns but does not separate the
primary experimental manipulations from their interactions and the control
conditions.

## Isolating Primary Experimental Effects

CPCA separates the primary experimental effects from the remaining structure,
so each basis answers one question rather than a blend of two.

`cpca_part = "both"` returns the design-aligned basis and the residual basis
together, \(K\)-orthogonal to each other, so the two can be compared directly.

```{r fit-cpca}
fit_cpca <- dkge(bundle,
                 K = diag(q),
                 cpca_blocks = 1:2,
                 cpca_part = "both",
                 rank = 3)

fit_cpca$cpca$part
round(fit_cpca$cpca$evals_design[1:3], 3)
round(fit_cpca$cpca$evals_resid[1:3], 3)
```

```{r cpca-spectrum, echo=FALSE, fig.height=3.8, fig.alt="Grouped bar chart comparing the first three eigenvalues in the design-aligned and residual CPCA subspaces."}
cpca_spectrum <- rbind(
  data.frame(component = 1:3, subspace = "Design aligned",
             eigenvalue = fit_cpca$cpca$evals_design[1:3]),
  data.frame(component = 1:3, subspace = "Residual",
             eigenvalue = fit_cpca$cpca$evals_resid[1:3])
)
ggplot2::ggplot(
  cpca_spectrum,
  ggplot2::aes(factor(component), eigenvalue, fill = subspace)
) +
  ggplot2::geom_col(position = "dodge", width = 0.7) +
  ggplot2::labs(
    title = "The declared split reorganizes the fitted covariance",
    x = "Component within subspace", y = "Eigenvalue", fill = NULL
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "top")
```

The two eigenvalue sequences summarize the covariance retained by the chosen
design and residual projectors. They provide a focused decomposition under this
model, not evidence that the underlying biological processes are independent.

The following check verifies \(K\)-orthogonality between the design and residual
bases. It is a numerical statement about the fitted coordinates, not a test of
independence or absence of shared signal:

```{r orthogonality}
Ud <- fit_cpca$cpca$U_design
Ur <- fit_cpca$cpca$U_resid
round(max(abs(t(Ud) %*% fit_cpca$K %*% Ur)), 6)
```

### Influence of the Design Kernel

The split is metric-aware: changing `kernel` alters which directions count as
"design-aligned." A smooth kernel diffuses the projector across neighboring
rows, so design energy leaks into adjacent effects.

```{r kernel-effect}
K_smooth <- outer(seq_len(q), seq_len(q), function(i, j) 0.7^abs(i - j))
fit_kernel <- dkge(bundle,
                   K = K_smooth,
                   cpca_blocks = 1:2,
                   cpca_part = "both",
                   rank = 3)
round(fit_kernel$cpca$evals_design[1:3], 3)
round(fit_kernel$cpca$evals_resid[1:3], 3)
```

The printed eigenvalues show how much fitted covariance each projector retains
under `K_smooth`. They do not display the loadings themselves. The projector
honours the correlation structure encoded by `K_smooth`, so the kernel choice
shapes which latent directions count as design-aligned. When effects do not
align with coordinate axes, pass a custom `cpca_T` that expresses the intended
K-weighted span explicitly.

The separation runs through projectors in the \(K\) metric:
`dkge_projector_K()` builds them and `dkge_cpca_split_chat()` applies them to
the compressed covariance.

```{r inspect-split}
T_design <- diag(1, q)[, 1:2]
split_plain <- dkge_cpca_split_chat(fit_plain$Chat, T_design, fit_plain$K)
round(diag(split_plain$Chat_design), 3)
round(diag(split_plain$Chat_resid), 3)
```

The diagonal shows how variance divides between design and residual
components. A large residual share means the effects you named do not account
for most of the compressed variance.

```{r chat-match}
max(abs(fit_cpca$Chat - fit_cpca$cpca$Chat_design))
```

## Using Custom Effect Combinations

When the effects of interest are not a simple subset of conditions, supply an
explicit basis instead of naming blocks.

The columns of a custom basis span the effect combinations you want analysed
together, so a contrast-like direction can define the design subspace.

```{r explicit-basis}
T_custom <- qr.Q(qr(cbind(c(1, 1, 0, 0, 0, 0),
                          c(0, 0, 2, 1, 0, 0))))
fit_custom <- dkge(bundle,
                   K = diag(q),
                   cpca_T = T_custom,
                   cpca_part = "design",
                   rank = 2)
round(fit_custom$cpca$evals_design[1:2], 3)
```

This basis creates two design components: the first combines the attention
conditions equally, the second contrasts them.

## Numerical Stabilization with Ridge Regularization

With real data the filtered covariance can be near-singular, which makes the
eigendecomposition unstable. `cpca_ridge` is the remedy.

```{r ridge}
fit_ridge <- dkge(bundle,
                  K = diag(q),
                  cpca_blocks = 1:2,
                  cpca_part = "design",
                  cpca_ridge = 1e-3,
                  rank = 3)
diag_shift <- diag(fit_ridge$cpca$Chat_design - fit_ridge$cpca$Chat_design_raw)
round(head(diag_shift), 6)
```

Ridge regularisation adds the given value to each diagonal element before the
eigendecomposition, which stabilises the solve without changing which effects
define the subspace.

## Simplified Interface for CPCA Analysis

`dkge_cpca_fit()` is a streamlined entry point for workflows that only need
the CPCA split.

```{r wrapper}
fit_wrapper <- dkge_cpca_fit(bundle,
                             K = diag(q),
                             cpca_blocks = 1:2,
                             cpca_part = "design",
                             rank = 3)
identical(round(fit_wrapper$U, 6), round(fit_cpca$U, 6))
```

## What this page showed

The split is a modeling choice, not a discovery. You declare which effects are
design-aligned — by name through `cpca_blocks` or as an explicit basis through
`cpca_T` — and `dkge_projector_K()` builds the projector in the \(K\) metric.
`dkge_cpca_split_chat()` applies it, and DKGE fits a basis for each requested
part, \(K\)-orthogonal to the other.

The returned bases are ordinary DKGE components: contrast testing, bootstrap
inference and the plotting helpers all work on them unchanged.

Whether the split improves power is design- and data-dependent, and should be
checked against an unfiltered fit rather than assumed.

## Where to go next

- `vignette("dkge-design-kernels")` — the kernel that defines which directions
  count as design-aligned, and therefore what CPCA is splitting.
- `vignette("dkge-components")` — how to read the components that come back
  from either side of the split.
- `vignette("dkge-concepts")` — why a \(K\)-orthogonal split is a statement
  about coordinates and not about independent processes.
