---
title: "DKGE Workflow"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: false
    df_print: paged

vignette: >
  %\VignetteIndexEntry{DKGE Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(dkge)
library(ggplot2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
set.seed(2024)
theme_set(theme_dkge())
```

DKGE compresses per-subject GLM beta matrices into a shared latent space that
respects the experimental design. This vignette walks through the core pipeline:
data prep → fit → component inspection → inference → bootstrap.

## Data preparation

Three subjects, three design effects, four spatial clusters each:

```{r example-data, message=FALSE}
S <- 3; q <- 3; P <- 4; T <- 30

betas   <- replicate(S, matrix(rnorm(q * P), q, P), simplify = FALSE)
designs <- replicate(S, {
  X <- matrix(rnorm(T * q), T, q)
  qr.Q(qr(X))   # orthonormalise for a clean compressed covariance
}, simplify = FALSE)
centroids <- replicate(S, matrix(runif(P * 3, -15, 15), P, 3), simplify = FALSE)

subjects    <- lapply(seq_len(S), function(s)
  dkge_subject(betas[[s]], design = designs[[s]], id = paste0("sub", s)))
data_bundle <- dkge_data(subjects)
```

`dkge_subject()` pairs each beta matrix with its design matrix. `dkge_data()`
harmonises effect ordering and subject IDs across the list.

## Fitting

```{r fit-model}
fit <- dkge(data_bundle, K = diag(data_bundle$q), rank = 2)
fit$centroids <- centroids  # convenience: attach for transport helpers; pass explicitly in production

round(fit$sdev, 3)   # component singular values
fit$weights          # per-subject MFA block weights
```

The identity kernel `diag(q)` applies no design smoothing — use
`design_kernel()` for structured factorial designs.

## Component structure

Cluster-wise loadings on the first component for each subject:

```{r inspect-components}
loadings <- lapply(fit$Btil, function(Bts) t(Bts) %*% fit$K %*% fit$U)
lapply(loadings, function(A) round(A[, 1, drop = FALSE], 3))
```

Project new subjects onto the fixed basis with `dkge_project_block()` or
`dkge_project_clusters()`.

## Component-level inference

`dkge_component_stats()` transports loadings to a shared medoid parcellation
and returns test statistics with FDR adjustment:

```{r component-stats}
comp <- dkge_component_stats(
  fit,
  mapper     = list(strategy = "sinkhorn", epsilon = 0.05, lambda_spa = 0.5),
  centroids  = centroids,
  inference  = list(type = "parametric"),
  medoid     = 1L
)
head(comp$summary)
```

`comp$transport` holds subject × medoid matrices after alignment;
`comp$statistics` holds raw component vectors.

## Mapping to voxel space

```{r voxel-transport, eval=FALSE}
voxels          <- replicate(S, matrix(runif(10 * 3, -20, 20), 10, 3), simplify = FALSE)
component_values <- lapply(loadings, function(A) A[, 1])

voxel_maps <- dkge_transport_to_voxels(fit,
  values = component_values, voxels = voxels, mapper = "ridge")
round(voxel_maps$value, 3)
```

`voxel_maps$value` is the group-average consensus map; `voxel_maps$subj_values`
preserves per-subject interpolations.

## Bootstrap inference

Cache the transport operators once; reuse them across all bootstrap draws:

```{r bootstrap-demo, message=FALSE}
cache <- dkge_prepare_transport(fit, centroids = centroids, medoid = 1)
values_medoid <- lapply(seq_len(nrow(comp$transport[[1]])),
                        function(i) comp$transport[[1]][i, ])

# Resample already-transported subject vectors
boot_proj <- dkge_bootstrap_projected(values_medoid, B = 200, seed = 99)

# Include subspace re-estimation uncertainty
boot_q <- dkge_bootstrap_qspace(fit, contrasts = c(1, -1, 0), B = 200,
                                transport_cache = cache, medoid = 1, seed = 99)

boot_proj$medoid$sd[1]
boot_q$summary[[1]]$medoid$sd[1]
```

`dkge_bootstrap_analytic()` offers a cheaper alternative via first-order
perturbation theory when thousands of draws are needed.

## Where next?

| Task | Function |
|------|----------|
| Hypothesis contrasts | `dkge_contrast()`, `dkge_pipeline()` |
| Out-of-sample scoring | `dkge_project_blocks()` |
| Structured design kernels | `design_kernel()` |
| Classification / decoding | `dkge_classify()` — see `vignette("dkge-classification")` |
| Inference details | `vignette("dkge-contrasts-inference")` |
