---
title: "Mapper Customisation and Performance"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: false
    df_print: paged

vignette: >
  %\VignetteIndexEntry{Mapper Customisation and Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(dkge)
library(ggplot2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
options(
  tibble.print_min = 4,
  tibble.print_max = 4,
  digits = 3
)
set.seed(2024)
theme_set(theme_dkge())
```


This vignette provides guidance on selecting between kNN and Sinkhorn mappers based on your analysis needs, demonstrates how to implement custom transport methods, and shows how to leverage warm starts to improve computational performance.

## Mapper Factory Recap

```{r mapper-factory}
library(dkge)
dkge_mapper("knn", k = 8, sigx = 3)
dkge_mapper("sinkhorn", epsilon = 0.05, lambda_xyz = 1, lambda_feat = 0)
```

The `dkge_mapper()` function creates a mapper configuration that stores the backend parameters for your chosen transport method. When you call `fit_mapper()` on this configuration, it produces subject-specific transport plans that include cached computational state, enabling efficient reuse across multiple mapping operations.

## Speed Comparison

```{r performance}
S <- 3; q <- 3; P <- 200
centroids <- replicate(S, matrix(rnorm(P * 3), P, 3), simplify = FALSE)
anchors <- matrix(rnorm(1500 * 3), 1500, 3)
values <- lapply(seq_len(S), function(s) rnorm(P))

knn_mapper <- dkge_mapper("knn", k = 12, sigx = 5)
knn_fit <- lapply(seq_len(S), function(s) fit_mapper(knn_mapper, centroids[[s]], anchors))

sink_mapper <- dkge_mapper("sinkhorn", epsilon = 0.02, lambda_xyz = 1)
sink_fit <- lapply(seq_len(S), function(s) fit_mapper(sink_mapper, centroids[[s]], anchors))

microbenchmark::microbenchmark(
  knn = lapply(seq_len(S), function(s) apply_mapper(knn_fit[[s]], values[[s]])),
  sinkhorn = lapply(seq_len(S), function(s) apply_mapper(sink_fit[[s]], values[[s]])),
  times = 10L
)
```

The performance characteristics of these two mapping approaches differ significantly. kNN mapping operates through purely local neighborhoods and scales linearly with the number of neighbors `k`, making it computationally efficient for most applications. In contrast, Sinkhorn mapping offers greater expressiveness by supporting feature-based costs and soft matching between points, but this flexibility comes with increased computational overhead. However, the optimized C++ solver implementation with warm start capabilities makes repeated Sinkhorn calls much more tractable for iterative analyses.

## Warm Starts and Dual Caching

```{r warm-start}
C <- as.matrix(dist(anchors[1:200, ]))
mu <- rep(1/nrow(C), nrow(C))
nu <- rep(1/ncol(C), ncol(C))
res1 <- dkge:::sinkhorn_plan_cpp(C, mu, nu, epsilon = 0.05, max_iter = 400, tol = 1e-7)
res2 <- dkge:::sinkhorn_plan_cpp(C, mu, nu, epsilon = 0.05, max_iter = 400, tol = 1e-7,
                                 log_u_init = res1$log_u, log_v_init = res1$log_v)
res1$iterations
res2$iterations
```

The second optimization run converges significantly faster because it leverages the saved dual variables from the previous computation. This warm start mechanism is automatically handled by the DKGE rendering pipeline, which stores these dual variables in the renderer object for subsequent use. When working directly with the lower-level `sinkhorn_plan_cpp()` function for experimentation, you can manually provide these initialization values to achieve similar performance gains.

## Adding Custom Mappers

To integrate a custom mapping approach into the DKGE framework, you need to implement two key S3 methods that follow the established naming convention recognized by `dkge_mapper()`.

```r
fit_mapper.dkge_mapper_mytransport <- function(mapper, subj_points, anchor_points, ...) {
  # return object with class 'dkge_mapper_fit_mytransport'
}

apply_mapper.dkge_mapper_fit_mytransport <- function(fitted_mapper, values, ...) {
  # return length(anchor_points) vector
}
```

As long as you maintain consistent output dimensions that match the expected anchor point structure, DKGE will seamlessly integrate your custom mapper into the existing analysis pipeline without requiring additional configuration.

## Practical Guidance

The choice between mapping methods should be guided by your specific analysis requirements and data characteristics. For quick exploratory analyses or situations where subject clusters already show good anatomical alignment, kNN mapping provides an efficient and straightforward solution.

However, when dealing with substantial anatomical shifts between subjects or when latent features should guide the matching process, Sinkhorn mapping becomes the preferred choice. In these cases, you can tune the `epsilon` parameter upward to achieve faster convergence and smoother transport plans, though this comes with a trade-off in matching precision.

To maximize computational efficiency across multiple analyses, make sure to reuse renderer objects when performing bootstraps or computing contrasts. This practice allows you to exploit the cached dual variables and avoid the computational overhead of recomputing transport plans from scratch.
