---
title: "Mapper Customisation and Performance"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: false
    df_print: paged

vignette: >
  %\VignetteIndexEntry{Mapper Customisation and Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(dkge)
library(ggplot2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
options(
  tibble.print_min = 4,
  tibble.print_max = 4,
  digits = 3
)
set.seed(2024)
theme_set(theme_dkge())
```


This vignette provides guidance on selecting between kNN and Sinkhorn mappers based on your analysis needs, demonstrates how to implement custom transport methods, and shows how to leverage warm starts to improve computational performance.

## Mapper Factory Recap

```{r mapper-factory}
library(dkge)
dkge_mapper("knn", k = 8, sigx = 3)
dkge_mapper("sinkhorn", epsilon = 0.05, lambda_xyz = 1, lambda_feat = 0)
```

The `dkge_mapper()` function configures the dense spatial pipeline. A fitted
Sinkhorn mapper keeps the joint plan, the target-conditional application
operator, and convergence diagnostics. Reusing that fitted object makes later
`apply_mapper()` calls matrix multiplications rather than new OT solves.

## Speed Comparison

```{r performance-compute, include=FALSE}
P <- 80
centroids <- matrix(rnorm(P * 3), P, 3)
anchors <- matrix(rnorm(200 * 3), 200, 3)
values <- rnorm(P)

knn_mapper <- dkge_mapper("knn", k = 12, sigx = 5)
sink_mapper <- dkge_mapper("sinkhorn", epsilon = 0.05, lambda_xyz = 1,
                           warm_start = FALSE)

time_fit <- function(mapper, reps = 3L) {
  median(replicate(reps, system.time(
    fit_mapper(mapper, centroids, anchors)
  )[["elapsed"]]))
}
knn_fit_seconds <- time_fit(knn_mapper)
sink_fit_seconds <- time_fit(sink_mapper)
knn_fit <- fit_mapper(knn_mapper, centroids, anchors)
sink_fit <- fit_mapper(sink_mapper, centroids, anchors)

time_apply <- function(fitted, reps = 100L) {
  system.time(replicate(reps, apply_mapper(fitted, values)))[["elapsed"]] / reps
}
performance <- data.frame(
  stage = rep(c("fit", "apply"), each = 2),
  mapper = rep(c("kNN", "Sinkhorn"), 2),
  seconds_per_call = c(knn_fit_seconds, sink_fit_seconds,
                       time_apply(knn_fit), time_apply(sink_fit))
)
stopifnot(all(is.finite(performance$seconds_per_call)),
          all(performance$seconds_per_call >= 0))
```

```{r performance}
knitr::kable(performance, digits = 5)
```

The table separates the solve from application. kNN fitting constructs local
neighbourhoods; Sinkhorn fitting constructs a dense cost matrix and iterates to
match both marginals, so its fit is usually the expensive stage. Once fitted,
both backends reuse their mappings. Measure both stages at your own parcel and
anchor counts rather than extrapolating from apply-only timings.

## Warm Starts and Dual Caching

DKGE caches duals only after a solve satisfies its requested marginal
tolerance. The key includes every cost and mass entry, so warm starts cannot
cross between merely similar problems. The cache is process-local; fitted
renderer objects store plans and operators, not dual vectors.

```{r warm-start}
dkge_clear_sinkhorn_cache()
cached_mapper <- dkge_mapper("sinkhorn", epsilon = 0.05, lambda_xyz = 1)
cold_fit <- fit_mapper(cached_mapper, centroids, anchors)
warm_fit <- fit_mapper(cached_mapper, centroids, anchors)

rbind(
  cold = unlist(cold_fit$stats$diagnostics[c("iterations", "cache_hit")]),
  repeated = unlist(warm_fit$stats$diagnostics[c("iterations", "cache_hit")])
)
```

```{r warm-start-checks, include=FALSE}
stopifnot(cold_fit$stats$diagnostics$converged,
          warm_fit$stats$diagnostics$converged,
          isTRUE(warm_fit$stats$diagnostics$cache_hit),
          identical(as.matrix(cold_fit$plan), as.matrix(warm_fit$plan)))
```

For this exact repeat, the cached converged plan is returned with zero new
iterations. If you request a tighter tolerance, DKGE instead uses the cached
duals as an initialization and continues solving.

## Adding Custom Mappers

To integrate a custom mapping approach into the DKGE framework, you need to implement two key S3 methods that follow the established naming convention recognized by `dkge_mapper()`.

```{r custom-mapper}
fit_mapper.dkge_mapper_mytransport <- function(mapper, subj_points,
                                                anchor_points, ...) {
  structure(list(Q = nrow(anchor_points)),
            class = "dkge_mapper_fit_mytransport")
}

apply_mapper.dkge_mapper_fit_mytransport <- function(fitted_mapper, values, ...) {
  rep(mean(values), fitted_mapper$Q)
}

custom <- dkge_mapper("mytransport")
custom_fit <- fit_mapper(custom, centroids, anchors)
head(apply_mapper(custom_fit, values))
```

The constructor accepts the custom identifier and S3 dispatch finds the two
methods. A renderer-compatible extension must also validate the standard point
inputs and return one value per anchor, as this example does.

## Practical Guidance

The choice between mapping methods should be guided by your specific analysis requirements and data characteristics. For quick exploratory analyses or situations where subject clusters already show good anatomical alignment, kNN mapping provides an efficient and straightforward solution.

However, when dealing with substantial anatomical shifts between subjects or when latent features should guide the matching process, Sinkhorn mapping becomes the preferred choice. In these cases, you can tune the `epsilon` parameter upward to achieve faster convergence and smoother transport plans, though this comes with a trade-off in matching precision.

For repeated maps, reuse renderer or fitted-mapper objects so the already-solved
plans and application operators are retained. Warm-start caching is most useful
when an identical fit is requested again; check `stats$diagnostics` instead of
assuming that a cache was used.
