---
title: "DKGE Concepts: Kernels, Inference, and What the Method Actually Does"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: false
    df_print: paged

vignette: >
  %\VignetteIndexEntry{DKGE Concepts: Kernels, Inference, and What the Method Actually Does}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(dkge)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
options(digits = 3)
set.seed(2024)
```

Read this before your first fit. It explains what DKGE is actually decomposing,
what choosing a kernel does to that decomposition, and at which level —
component, contrast, or feature — you can claim statistical evidence. Workflow
mechanics live in `vignette("dkge-workflow")`; kernel construction details in
`vignette("dkge-design-kernels")`; the PLS-by-PLS comparison in
`vignette("dkge-vs-pls")`. This vignette is the orientation that ties them
together.

## What DKGE actually decomposes

A naive group decomposition would eigendecompose the raw subject covariance
\(\sum_s B_s B_s^\top\). Current DKGE separates that calculation into two
auditable stages. First it constructs a pooled raw-effect moment \(M\), which
may include coverage masks, effect precision, and finite-trial debiasing. It
then eigendecomposes

\[
\widehat C \;=\; K^{1/2} R^\top M R K^{1/2},
\]

where \(R\) is the pooled design ruler. With `effect_scaling = "none"`, \(R\)
is the identity. DKGE maps the resulting eigenvectors back through
\(K^{-1/2}\) to obtain the group basis \(U\). Three consequences follow that
are worth internalising before choosing a kernel:

1. **\(K\) is a metric, not a basis.** DKGE does not project effects onto
   pre-chosen kernel directions; it changes the inner product under which the
   covariance is diagonalised. Components are free to be any direction in
   q-space — the kernel only re-weights how "large" each direction looks during
   the eigensolve.
2. **The data \(M\) still does the work.** If the empirical moment has a
   dominant direction, that direction will surface regardless of \(K\) (subject
   to scaling). The kernel amplifies directions it considers smooth /
   structurally plausible and attenuates rough or "unrelated" ones. It does not
   manufacture structure that the data does not already support.
3. **Coverage and precision are resolved before the metric mixes rows.** A
   missing cell, a low-count cell, and a noisy cell change \(M\), not \(K\).
   The kernel can subsequently smooth their pooled evidence, but it does not
   decide whether a subject observed a row or how precise that row was.

In aggregate-target mode (`dkge_aggregate_fit()`) the same machinery applies to
\(Y Y^\top\) where \(Y\) is a cells × features matrix — useful when you want a
PLSC-style decomposition that still admits a design kernel. The subject-level
fit `dkge_fit()` instead pools subject effect moments in \(M\), then applies
the design ruler \(R\) before the kernel congruence.

### Identity-K is the unconstrained baseline

`K = diag(q)` collapses the kernel transform to the identity, leaving an
eigendecomposition of \(R^\top M R\) under pooled design scaling, or of \(M\)
when `effect_scaling = "none"`.
There is no kernel-imposed similarity, smoothness, or hierarchy among effects.
For an aggregate fit this is essentially the left-singular-vector structure of
\(Y Y^\top\) — i.e. ordinary cell-mean PCA, the "let the data speak" mode.

For a full subject-level fit, identity-\(K\) is *not* equivalent to ignoring
design entirely: rows are still GLM effects, and DKGE still row-standardises
through the pooled design Cholesky factor \(R\). What identity-\(K\) means is
"do not couple effects beyond their GLM/design scaling". If you explicitly use
`effect_scaling = "none"`, identity-\(K\) instead diagonalises the pooled raw-
effect moment directly.

### Structured K is prior-regularised

Anything else — `design_kernel(factors, terms, ...)`, ordinal kernels, block
factors, term-weighted compositions — biases the decomposition toward
directions that lie in high-prior-variance subspaces and away from directions
the kernel calls rough. This is a feature, not a bug; it is how design
information gets into the latent geometry. But it does mean components from a
structured fit cannot be reported as if they had "emerged" from a neutral
decomposition.

**Discipline:** when reporting structured-\(K\) components, present an
identity-\(K\) refit alongside as the unconstrained baseline. If the same
components dominate both, the structure is in the data. If the kernel
substantially changes the loadings or eigenspectrum, the structure is partly
prior.

## Identity vs structured K — a small worked example

A 2 × 2 aggregate cell-mean matrix with four cells and forty features. We pass
the matrix straight to `dkge_aggregate_fit()` — the dedicated
`dkge_aggregate_target()` constructor is only needed when you want subject-
or weight-aware bootstrap/permutation later (see `?dkge_aggregate_target`).

```{r aggregate-data}
P <- 40
Y <- matrix(rnorm(4 * P), 4, P)
rownames(Y) <- c("A1:B1", "A1:B2", "A2:B1", "A2:B2")
```

Fit twice — identity-\(K\) baseline and a factor-structured kernel:

```{r aggregate-refit, message=FALSE}
fit_I <- dkge_aggregate_fit(Y, K = diag(4), rank = 3)

K_struct <- design_kernel(
  factors = list(A = list(L = 2, type = "nominal"),
                 B = list(L = 2, type = "nominal")),
  terms   = list("A", "B", c("A", "B")),
  rho     = c(A = 1, B = 1, "A:B" = 0.3),
  basis   = "cell",
  normalize = "unit_trace"
)
fit_K <- dkge_aggregate_fit(Y, K = K_struct, rank = 3)

data.frame(
  component   = 1:3,
  identity_K  = round(fit_I$singular_values, 3),
  structured_K = round(fit_K$singular_values, 3)
)
```

Two questions to ask of this table:

- **Did the eigenspectrum change shape?** A flat-to-steep shift means the
  kernel is concentrating energy in directions it favours.
- **Did the leading loadings change?** Compute `crossprod(fit_I$U, fit_K$U)`;
  off-diagonal mass tells you how much the components rotated. Near-identity
  means the data already aligned with the kernel — structured-\(K\) just made
  it cleaner. Far-from-identity means the kernel is doing the talking.

Either result is acceptable; the obligation is to *know which* and to say so.

## How do missing cells, unequal counts, and trial noise differ?

These three issues enter the same q-space moment but represent different
estimands.

| Issue | Stored evidence | Fit control | Diagnostic |
|---|---|---|---|
| Subject did not observe an effect | observation mask | `missingness` | `pair_counts` |
| Subject observed it with low precision | count or explicit precision | `effect_weights` | `pair_weight`, `pair_ess` |
| Estimated effect contains finite-trial noise | residual variance and \((X^\top X)^{-1}\), or split betas | `debias` | `moment_diagnostics` |

`dkge_trial_subject()` reduces trial-by-feature maps to these sufficient
statistics without forming a feature-by-feature covariance. Count weighting
then uses pair reliability \(\sqrt{p_{sc}p_{sc'}}\), while analytic debiasing
subtracts the expected effect-space noise moment. The operations precede the
\(R\) and \(K^{1/2}\) congruence, so every solver sees the same explicitly
defined pooled moment. See `vignette("dkge-partial-effect-spaces")` for the
runnable 3 x 5 x 4 example and the policy choices.

## How DKGE conducts inference

There are four distinct levels at which you can make a claim. They use
different machinery and answer different questions. Conflating them is the
single most common reporting error.

| Level | Question | Tool | Unit |
|-------|----------|------|------|
| **Component** | "Is this latent direction reliable?" | singular values; subject-label permutation; identity vs structured \(K\) refit; bootstrap of \(U\) | latent dimension |
| **Contrast** | "Is this q-space contrast non-zero?" | `dkge_contrast()` (LOSO / K-fold), `dkge_bootstrap_qspace()`, analytic SEs | linear combination of effects |
| **Feature** | "Is this cluster / voxel salient under this contrast or component?" | `dkge_component_stats()`, `dkge_transport_contrasts_to_medoid()`, FDR adjustment | spatial unit (cluster, anchor, voxel) |
| **Between-subject term** | "Does group or a trait explain a subject-level target?" | `dkge_between_rrr()` plus `dkge_between_permute()` | subjects |

The structure is hierarchical: component-level inference asks whether a
direction exists; contrast-level inference asks whether a chosen direction has
non-zero magnitude; feature-level inference asks where, spatially, that
magnitude is concentrated. Every published finding should name its level.

A few practical notes:

- **Subject is the unit of inference for between-subject claims.** The
  between-subject RRR layer tests group, trait, and interaction terms on a
  subject-by-feature target. Residual rotation fixes nuisance fits and rotates
  only the residual subject space; Freedman-Lane permutes reduced-model
  residual rows. Neither treats cells, anchors, or voxels as independent
  subjects.
- **Cross-fitting (LOSO / K-fold) is the bias-aware default for contrasts.**
  In-fold contrasts inherit the optimism of the same data used to fit \(U\);
  `method = "loso"` removes that.
- **Identity vs structured \(K\) refits double as a sanity check for
  component-level claims.** This is the natural DKGE analogue of the
  multifer-style "is this LV real?" diagnostic: a component that survives
  ablating the kernel structure is one the data is asserting, not one the
  prior is asserting.

## Where DKGE sits among related methods

DKGE is not a new family of decomposition. It is a metric-aware version of
methods you already know, plus a transport and inference layer on top.

| Method | Decomposes | Design role | Output |
|--------|-----------|-------------|--------|
| **PCA on cell means** | \(Y Y^\top\) | none | data-driven axes of cell variation |
| **PLSC** (McIntosh & Lobaugh) | cross-block \(C^\top M\) via SVD | orthonormal contrast block | covariance-maximising LVs between design and brain |
| **`dkge_aggregate_fit(K = I)`** | \(Y Y^\top\) | none beyond cell labels | identical in spirit to cell-mean PCA |
| **`dkge_aggregate_fit(K = K_struct)`** | \(K^{1/2} Y Y^\top K^{1/2}\) | metric / preconditioner | LVs in the kernel-induced inner product |
| **`dkge_fit(K = I)`** | \(R^\top M R\), or \(M\) with no effect scaling | GLM row-standardisation only | unconstrained subject-level group basis |
| **`dkge_fit(K = K_struct)`** | \(K^{1/2}R^\top M R K^{1/2}\) | metric / preconditioner | design-aware subject-level group basis |

Two broader relatives, for orientation:

- **RDA / CCA** project the response onto a subspace spanned by a design
  block. DKGE never restricts the response space; it changes the metric used
  on it. An RDA fit with rank \(r\) and identity-metric is closer to a
  *strict-K* analogue where \(K\) is rank-deficient and supported on the
  design subspace.
- **Ridge-regularised linear discriminants** (e.g. ridge LDA) likewise modify
  the metric of the underlying scatter problem. DKGE's structured-\(K\) refit
  is a closer cousin to ridge LDA than to either PCA or PLSC: both pre- and
  post-multiply the empirical second-moment matrix by a fixed PSD weight
  before solving.

If you find yourself wanting to say "this LV is design-driven", the right
comparison is identity-\(K\) DKGE, not PCA on the raw betas — those two only
coincide for aggregate fits.

## Before you fit

A short checklist that maps the above to actions:

1. **Decide which estimand you want.** Aggregate cell means
   (`dkge_aggregate_fit()`) or subject-level group structure (`dkge()` /
   `dkge_fit()`). The inference unit is subjects in both cases.
2. **Always run identity-\(K\) at least once.** It is the unconstrained
   baseline. Save the eigenspectrum, the leading loadings, and one or two
   contrast LOSO results.
3. **Add structure deliberately.** Build a `design_kernel()` only for terms
   you have prior reason to couple. Use `dkge_cv_rank_loso()` to choose rank
   and kernel parameters jointly (see `vignette("dkge-design-kernels")`).
4. **Compare identity vs structured \(K\) refits before reporting.** Diff
   singular values; compute `crossprod(U_I, U_K)`; note whether structure is
   data-asserted or prior-asserted.
5. **Resolve coverage, precision, and trial noise separately.** Use
   `observed_rows`/`missingness`, `dkge_effect_weights()`, and `debias` for their
   respective jobs; do not use one as a proxy for another.
6. **Name the inference level.** Component, contrast, feature, or
   between-subject term. Use the
   matching tool from the table above; do not transfer a permutation
   p-value from one level to another.

A finding that survives all six steps is one you can defend.

## See also

- `vignette("dkge-workflow")` — end-to-end pipeline
- `vignette("dkge-design-kernels")` — kernel construction and CV-based
  selection
- `vignette("dkge-contrasts-inference")` — LOSO, analytic, and bootstrap
  inference details
- `vignette("dkge-components")` — component interpretation and rotation
- `vignette("dkge-vs-pls")` — detailed comparison with classical PLSC
- `vignette("dkge-partial-effect-spaces")` — partial effect coverage and
  trialwise reliability/debiasing
- `vignette("dkge-between-subjects")` — reduced-rank group/trait models and
  residual-rotation versus Freedman-Lane inference
