---
title: Components and Interpretability
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Components and Interpretability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(dkge)
library(ggplot2)
library(tidyr)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
options(
  tibble.print_min = 4,
  tibble.print_max = 4,
  digits = 3
)
set.seed(2024)
theme_set(theme_dkge())
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```


This vignette asks how a fitted component is expressed in effect space and on
a shared toy cluster grid. It then shows rotation and projection without
treating those descriptive operations as inferential results.

## Example Fit

```{r fit}
library(dkge)
S <- 4; q <- 5; P <- 18; T <- 70
effect_names <- paste0("effect", seq_len(q))
cluster_names <- paste0("cluster", seq_len(P))
cluster_axis <- seq(-1, 1, length.out = P)
shared_signal <- rbind(
  exp(-4 * (cluster_axis + 0.45)^2),
  -exp(-4 * (cluster_axis - 0.45)^2),
  cluster_axis,
  rep(0, P),
  rep(0, P)
)
betas <- lapply(seq_len(S), function(s) {
  B <- shared_signal + matrix(rnorm(q * P, sd = 0.35), q, P)
  dimnames(B) <- list(effect_names, cluster_names)
  B
})
designs <- replicate(S, {
  X <- matrix(rnorm(T * q), T, q)
  X <- qr.Q(qr(X))
  colnames(X) <- effect_names
  X
}, simplify = FALSE)
subjects <- lapply(seq_len(S), function(s) dkge_subject(betas[[s]], designs[[s]], id = paste0("sub", s)))
bundle <- dkge_data(subjects)
fit <- dkge(bundle, K = diag(q), rank = 3)
```

By default `dkge()` accumulates subject contributions in the shared effect
space and diagonalizes the pooled moment. To request joint diagonalization of
the subject matrices instead, set `solver = "jd"` (and optionally pass
`jd_control = dkge_jd_control(...)` to configure the optimizer). Treat agreement
or disagreement with the pooled solution as a model comparison; the solver
choice does not guarantee greater stability.

```{r fit-jd, eval=FALSE}
fit_jd <- dkge(bundle,
               K = diag(q),
               rank = 3,
               solver = "jd",
               jd_control = dkge_jd_control(maxit = 200, tol = 1e-8))
```

Both solvers return the same object structure (`fit$U`, `fit$Chat`, projections, etc.), so the
subsequent workflow in this vignette applies unchanged regardless of the solver choice.

## Inspect Loadings and Scores

```{r loadings}
round(fit$U, 3)         # effect-space loadings
lapply(fit$Btil, function(Bts) round(Bts[, 1:3, drop = FALSE], 2))[1]
```

The columns of `fit$U` represent the loadings for each component in the effect space. These loadings define how the original design effects combine to form the extracted components. The `Btil` element stores the subject-specific beta matrices after row-standardization, which ensures comparable scaling across subjects. To obtain cluster scores for each component, these standardized betas are multiplied with the corresponding columns of `fit$U`.

## Projecting Subjects into Component Space

```{r project}
subject_scores <- dkge_project_btil(fit, fit$Btil)
str(subject_scores, max.level = 1)
```

Each subject score matrix has dimensions clusters × components. Cluster rows
are only comparable across subjects when they already refer to the same spatial
support. This simulation deliberately uses a shared 18-cluster indexing scheme,
so a row-wise mean is meaningful. With subject-specific parcellations, first
transport the scores to a common medoid, anchor, or voxel space; equal matrix
dimensions alone do not establish anatomical correspondence.

```{r score-plot, fig.height=4, fig.width=5, fig.alt="Line plot showing average DKGE component scores across clusters."}
avg_scores <- Reduce("+", subject_scores) / length(subject_scores)
avg_df <- as.data.frame(avg_scores)
component_cols <- paste0("Component ", seq_len(ncol(avg_df)))
names(avg_df) <- component_cols
avg_df$Cluster <- seq_len(nrow(avg_df))
avg_long <- tidyr::pivot_longer(avg_df,
                                cols = seq_along(component_cols),
                                names_to = "Component",
                                values_to = "Score")

ggplot(avg_long, aes(x = Cluster, y = Score, colour = Component)) +
  geom_line(linewidth = 1.1) +
  labs(title = "Average component scores", y = "Score") +
  scale_x_continuous(breaks = seq_len(nrow(avg_scores))) +
  theme(legend.position = "top")
```

The plot summarizes expression on this shared toy grid. It is descriptive: a
large value identifies a cluster-component pairing to inspect, not a
significant spatial effect.

## Rotating Components

The `dkge_procrustes_K()` function provides a principled way to rotate components toward a target basis, such as canonical contrasts or previously established loadings from prior studies. This rotation preserves the mathematical properties of the solution by staying within the design-kernel metric space.

```{r rotate}
# target basis: identity for first two effects
B_target <- dkge_k_orthonormalize(diag(1, q)[, 1:2], fit$K)
rot <- dkge_procrustes_K(B_target, fit$U[, 1:2], fit$K)
round(rot$U_aligned, 3)
```

The returned `rot$U_aligned` is the fitted basis rotated toward `B_target`.
`rot$d` reports the objective actually achieved by `rot$R`; when reflections
are forbidden, it can be smaller than `rot$unconstrained_d`.

```{r rotate-checks, include=FALSE}
stopifnot(
  max(abs(crossprod(rot$U_aligned, fit$K %*% rot$U_aligned) - diag(2))) < 1e-7,
  abs(rot$d - sum(diag(crossprod(B_target, fit$K %*% fit$U[, 1:2]) %*% rot$R))) < 1e-8
)
```

## Projecting New Data

```{r new-data}
new_beta <- matrix(rnorm(q * P), q, P)
projected <- dkge_project_clusters(fit, new_beta)
head(projected)
```

The `dkge_project_clusters()` function returns component scores for each cluster in the new data, providing a standardized representation that is ready for subsequent analyses such as contrast testing or optimal transport mapping to reference parcellations.

## Interpreting Components

Several outputs answer different interpretive questions. `fit$weights` contains
one fitting weight per subject, not a component-specific contribution. For
component-specific participation, inspect `fit$contribs`,
`dkge_plot_subject_contrib()`, or `dkge_subject_component_projections()`.
`fit$v` concatenates subject-block cluster loadings; interpret its rows spatially
only when the blocks share a declared support or after transport to a common
space. Finally, `dkge_component_stats()` supplies component summaries under its
chosen mapper and inference assumptions. Those assumptions and the target of
inference must be reported; the function does not make every extracted
component generically “significant.”
