---
title: "Design Kernels and Model Tuning"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: false
    df_print: paged

vignette: >
  %\VignetteIndexEntry{Design Kernels and Model Tuning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(dkge)
library(ggplot2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
options(
  tibble.print_min = 4,
  tibble.print_max = 4,
  digits = 3
)
set.seed(2024)
theme_set(theme_dkge())
```


This vignette demonstrates how the design kernel controls and guides DKGE model fits, and provides practical guidance on tuning rank and penalty parameters before conducting large-scale analyses. Understanding how to construct appropriate kernels and select optimal parameters is crucial for achieving robust and interpretable results in cluster-level fMRI analysis.

## Simulated Working Example

```{r data}
library(dkge)
S <- 4; q <- 4; P <- 30; T <- 80

betas <- replicate(S, matrix(rnorm(q * P, sd = 0.5), q, P), simplify = FALSE)
true_kernel <- 0.9 * (diag(q) + 0.3)
cholK <- chol(true_kernel)
betas <- lapply(betas, function(B) cholK %*% B + 0.1 * matrix(rnorm(q * P), q, P))

designs <- replicate(S, {
  X <- matrix(rnorm(T * q), T, q)
  qr.Q(qr(X))
}, simplify = FALSE)

subjects <- lapply(seq_len(S), function(s) dkge_subject(betas[[s]], designs[[s]], id = paste0("sub", s)))
bundle <- dkge_data(subjects)
```

The `dkge_data()` function serves an important preprocessing role by aligning effect ordering across subjects and caching common metadata for efficient computation. The design kernel that we specify in subsequent steps determines how strongly different effects are smoothed or grouped together during the embedding process.

## Building Kernels with `design_kernel()`

```{r kernels}
factors <- list(
  A = list(L = 2, type = "nominal"),
  B = list(L = 2, type = "nominal")
)
terms <- list("A", "B", c("A", "B"))
K_struct <- design_kernel(
  factors,
  terms = terms,
  rho = c("A" = 1, "B" = 1, "A:B" = 0.4),
  basis = "cell",
  normalize = "unit_trace"
)

round(K_struct$K, 2)
```

The kernel construction process involves several key concepts. Each term in the design introduces a block-similarity component that captures relationships between effects. The `rho` parameter weights the contribution of each term, with a default value of 1, allowing you to control the relative importance of different components. By combining multiple terms, you can encode complex relationships such as interactions or ordered trends within your experimental design. The `design_kernel()` function returns not only the kernel matrix itself but also associated metadata that will be used by the cross-validation helper functions described in the following sections.

## Which rank and kernel predict held-out structure?

`dkge_cv_kernel_rank()` first screens named kernels by alignment, then evaluates
the retained kernel-rank pairs with leave-one-subject-out explained energy. It
returns the selected pair and the tables needed to audit that choice.

```{r kernel-cv, message=FALSE}
rho_vals <- seq(0.4, 1.0, by = 0.3)
K_grid <- list()
for (ra in rho_vals) {
  for (rb in rho_vals) {
    nm <- sprintf("A%.1f_B%.1f", ra, rb)
    K_grid[[nm]] <- design_kernel(
      factors,
      terms = terms,
      rho = c("A" = ra, "B" = rb, "A:B" = 0.4),
      basis = "cell",
      normalize = "unit_trace"
    )$K
  }
}

selection <- dkge_cv_kernel_rank(
  bundle$betas, bundle$designs,
  K_grid = K_grid,
  ranks = 1:3,
  top_k = length(K_grid)
)
selection$picks_per_kernel
```

The score is the fraction of a held-out subject's K-metric energy captured by
the training-fold basis. It is descriptive predictive evidence, not a
significance test. Higher values mean that the fitted subspace captures more of
the held-out energy. The selector applies its documented one-standard-error
rule, preferring a lower rank among eligible pairs rather than blindly taking
the largest observed mean.

```{r refit}
best <- selection$pick
best

K_best <- K_grid[[best$kernel]]
fit <- dkge(bundle, K = K_best, rank = best$rank)
round(fit$sdev, 3)
```

## What does ridge regularization do here?

The `ridge` argument adds an isotropic diagonal term to the fitted q-space
moment before eigendecomposition. For a fixed moment this shifts all
eigenvalues equally and leaves its eigenvectors unchanged; it does not identify
or selectively down-weight noisy effects. Encode effect-specific structure in
the kernel or input weights instead. Ridge is mainly a numerical guard for
near-singular moment calculations.

```{r ridge}
fit_ridge <- dkge(bundle, K = K_best, rank = best$rank, ridge = 0.2)
round(fit_ridge$sdev, 3)
```

## Summary

Use `design_kernel()` to encode effect relationships you can defend. If the
structure is uncertain, include `diag(q)` as a baseline rather than assuming a
structured kernel must win. Compare kernels and ranks with
`dkge_cv_kernel_rank()`, interpreting its scores as held-out explained energy.
Refit the selected configuration before running inference,
and treat ridge as isotropic numerical regularization rather than an
effect-selection device.
