---
title: DKGE versus Partial Least Squares
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{DKGE versus Partial Least Squares}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(dkge)
library(ggplot2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
withr::local_options(list(
  tibble.print_min = 4,
  tibble.print_max = 4,
  digits = 3
))
set.seed(2024)
theme_set(theme_dkge())
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```


This page is a conceptual orientation, not an accuracy or power benchmark.
“PLS” names several related methods, and implementation choices matter. The
summary follows the neuroimaging accounts of
[McIntosh and Lobaugh (2004)](https://doi.org/10.1016/j.neuroimage.2004.07.020)
and [Krishnan et al. (2011)](https://doi.org/10.1016/j.neuroimage.2010.07.034).
It compares their classical task/behavior PLS formulations with DKGE's current
subject-beta workflow so you can decide which estimand and input contract match
your study.



---
## What classical PLS does

Classical PLS neuroimaging analysis operates by constructing latent variables that capture the strongest relationships between brain activity and experimental or behavioral variables. The **primary objective** is to build latent variables (LVs) that maximise the covariance between an exogenous block (such as task design or behavior) and a brain-data block (elements × time).

The framework encompasses several **methodological variants**, each tailored to different research questions: task PLS analyzes condition differences, behavior PLS examines brain–behavior coupling, seed PLS investigates functional connectivity patterns, and spatiotemporal PLS treats space × time jointly for fMRI/ERP/MEG data.

The **computational workflow** follows a systematic sequence of steps:
  1. Data are arranged as a single matrix \(M \in \mathbb{R}^{(n k) \times (m t)}\) with observations nested inside conditions.
  2. Cross-block covariance is formed with an orthonormal design matrix, followed by singular value decomposition \(C^\top M = U S V^\top\).
  3. Element/time saliences (singular images), design saliences, and singular values are extracted from the decomposition.
  4. Brain scores \(B = M U\) and design scores \(D = C V\) are computed to characterize patterns in each domain.
  5. For behavior PLS specifically, the design block is replaced with behavior matrices and correlated with \(M\) before applying SVD.

**Statistical inference** in PLS relies on resampling approaches: permutation tests assess LV significance, bootstrap procedures evaluate voxel salience reliability, and Procrustes alignment stabilizes resampled LVs across iterations.

This methodology emphasizes *whole-pattern* effects and time-resolved couplings, leveraging linear algebra and resampling techniques to identify distributed, reliable neural patterns that relate to experimental or behavioral variables.

## How DKGE relates

Both methods construct compact latent representations, but they begin from
different objects. Classical task or behavior PLS constructs a cross-block
matrix linking a design/behavior block to brain measurements. DKGE pools
subject-level GLM effect moments and applies an explicit effect-space kernel.
The table records implemented differences; it does not establish that one
method is generally more accurate, stable, or powerful.

<style>
table.dkge-vs-pls {
  table-layout: fixed;
  width: 100%;
}
table.dkge-vs-pls th,
table.dkge-vs-pls td {
  vertical-align: top;
}
table.dkge-vs-pls td {
  overflow-wrap: anywhere;
}
</style>

<table class="dkge-vs-pls">
  <colgroup>
    <col style="width: 20%;">
    <col style="width: 40%;">
    <col style="width: 40%;">
  </colgroup>
  <thead>
    <tr>
      <th scope="col">Aspect</th>
      <th scope="col">Partial Least Squares</th>
      <th scope="col">DKGE</th>
    </tr>
  </thead>
  <tbody>
    <tr>
      <th scope="row">Latent-space construction</th>
      <td>SVD on cross-block covariance between design/behaviour and brain data; columns of \(U\) and \(V\) are saliences.</td>
      <td>Eigen-decomposition of a design-kernel-weighted covariance, producing orthonormal components \(U\).</td>
    </tr>
    <tr>
      <th scope="row">Design information</th>
      <td>Specified through the exogenous/design block; the exact coding and available structure depend on the PLS variant.</td>
      <td>Explicit design kernel \(K\) encodes factorial structure, smoothness, or prior correlations among effects.</td>
    </tr>
    <tr>
      <th scope="row">Data normalisation</th>
      <td>Conditions averaged or mean-centred before SVD; each voxel treated equally.</td>
      <td>Row standardisation of subject betas using the pooled design Cholesky factor; optional spatial/reliability weights \(\Omega_s\).</td>
    </tr>
    <tr>
      <th scope="row">Cross-validation</th>
      <td>Classical neuroimaging workflows commonly use permutation for LV significance and bootstrap for salience stability; predictive cross-validation can be added but is not the same estimand.</td>
      <td>LOSO / K-fold cross-fitting (`dkge_contrast`), analytic approximations, parametric or bootstrap inference with cached transports.</td>
    </tr>
    <tr>
      <th scope="row">Transport / alignment</th>
      <td>Outputs latent scores; spatial interpretation relies on the original voxel grid.</td>
      <td>Provides barycentric kNN and Sinkhorn transports, anchor graphs, and voxel decoders for consistent spatial maps across parcellations.</td>
    </tr>
    <tr>
      <th scope="row">Reliability weighting</th>
      <td>Baseline formulations often use uniform observation weights; weighted, sparse, and regularized PLS variants also exist.</td>
      <td>Subject- and cluster-level reliabilities enter directly (e.g. sizes, inverse variances), influencing fits and transport.</td>
    </tr>
    <tr>
      <th scope="row">Spatiotemporal support</th>
      <td>ST-PLS handles time by stacking features.</td>
      <td>DKGE works on any GLM-derived beta blocks; temporal modelling is delegated to the design matrix and optional kernels.</td>
    </tr>
    <tr>
      <th scope="row">Implementation focus</th>
      <td>Exploratory LVs; complementary to other analyses.</td>
      <td>Integrated workflow for group GLM analysis, transport, and inference tailored to fMRI/ERP pipelines.</td>
    </tr>
  </tbody>
</table>

## Similarities worth noting

Despite their methodological differences, PLS and DKGE share several fundamental characteristics that reflect their common mathematical foundations. Both approaches rely on SVD or eigendecomposition techniques to obtain orthogonal latent patterns and their associated scores, ensuring that the derived components capture independent sources of variation in the data.

Resampling methodology plays a central role in both frameworks, though implemented differently: PLS employs permutation tests and bootstrap procedures for statistical inference, while DKGE provides a broader toolkit including analytic approximations, leave-one-subject-out (LOSO) cross-fitting, bootstrap procedures, and transport-aware resampling utilities.

Both methods require careful interpretation that involves examining latent loadings or saliences in conjunction with subject scores to properly understand how experimental conditions or behavioral variables relate to the underlying neural patterns.

## Which method matches the question?

Choose a PLS formulation when the target quantity is cross-block covariance
between a brain block and an experimental or behavioral block and the chosen
PLS implementation supplies the resampling procedure you need. Choose DKGE
when the inputs are already subject-level GLM beta matrices, the scientific
model requires an explicit effect-space metric, or subject-specific
parcellations must be transported before feature-level summaries.

These are workflow distinctions, not demonstrated performance advantages.
DKGE's cross-fitting addresses basis reuse only when the selected mode actually
refits the basis; reliability weights help only when their construction is
defensible; and spatial transport is useful only when its features and mapper
support plausible correspondence. Conversely, the table does not cover the
many sparse, regularized, predictive, or multilevel PLS variants.

For an empirical comparison, prespecify one estimand, fit both methods to the
same analysis units, use identical outer resampling splits, and compare a
held-out metric with uncertainty. Do not infer superiority from the fact that
one package exposes a helper that the other workflow leaves to the analyst.
