---
title: Feature-Anchored DKGE
author: dkge team
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Feature-Anchored DKGE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

## Overview

This vignette describes how to run DKGE when each subject is observed on a different set of stimuli but every item carries a feature vector in a shared space (e.g., a 100-dimensional embedding). The feature-anchored workflow replaces discrete-cell completion with a common anchor basis derived directly from the feature space.

We will:

1. build an anchor descriptor from subject-specific features and item kernels,
2. fit DKGE through the shared `dkge_input` interface, and
3. evaluate LOSO contrasts conditional on that fixed anchor representation.

```{r libraries}
library(dkge)
library(Matrix)
set.seed(1)
```

## Simulated feature-aligned data

In this example we simulate three subjects. Each subject has their own set of item features sampled around four latent prototypes. We generate subject-specific beta maps by projecting the item responses through SVD loadings and adding gaussian noise.

```{r simulate-data}
# Number of latent anchors and feature dimension
d <- 20L
anchors_true <- matrix(rnorm(4 * d), 4, d)

make_subject <- function(n_items, n_vox, seed) {
  set.seed(seed)
  # Each subject observes item features around the latent anchors
  centers <- anchors_true[sample.int(nrow(anchors_true), n_items, replace = TRUE), , drop = FALSE]
  features <- centers + matrix(rnorm(n_items * d, sd = 0.4), n_items, d)

  # Build an item similarity kernel (e.g., RSA over betas)
  latent <- matrix(rnorm(n_items * 3), n_items, 3)
  beta_loadings <- matrix(rnorm(3 * n_vox), 3, n_vox)
  betas <- latent %*% beta_loadings + matrix(rnorm(n_items * n_vox, sd = 0.2), n_items, n_vox)
  item_kernel <- betas %*% t(betas)

  list(features = features,
       item_kernel = item_kernel)
}

subjects <- list(
  s1 = make_subject(30, 120, seed = 11),
  s2 = make_subject(45, 120, seed = 12),
  s3 = make_subject(35, 120, seed = 13)
)

features_list <- lapply(subjects, `[[`, "features")
K_item_list  <- lapply(subjects, `[[`, "item_kernel")
```

## Build an anchor descriptor

We choose 16 anchors via the default d-kpp selector. The descriptor records both the anchor configuration and the DKGE options to be used after congruence.

```{r build-descriptor}
anchor_input <- dkge_input_anchor(
  features_list = features_list,
  K_item_list = K_item_list,
  anchors = list(L = 16, method = "dkpp", seed = 99L),
  dkge_args = list(w_method = "none")
)
```

This call chooses anchors from the pooled subjects; no fold-specific anchor
construction is requested. The anchor representation is therefore fixed before
the LOSO analysis below. That is a useful conditional analysis, but it is not
end-to-end cross-validation of feature preprocessing.

## Fit DKGE through the shared interface

`dkge_fit_from_input()` converts the descriptor into aligned anchor kernels and calls the standard fitter. The resulting object is a regular `dkge` fit containing the anchor provenance.

```{r fit-anchor}
fit_anchor <- dkge_fit_from_input(anchor_input)
fit_anchor
fit_anchor$provenance$anchors$coverage
```

## LOSO basis refitting on a fixed anchor representation

Contrasts are vectors over the anchor index. For illustration we select the
first anchor coordinate and refit the DKGE basis without each held-out subject.
The anchors themselves remain the pooled anchors created above.

```{r contrasts-anchor}
contrast_vec <- rep(0, fit_anchor$provenance$anchors$L)
contrast_vec[1] <- 1

res_contrast <- dkge_contrast(fit_anchor,
                               contrasts = list(anchor1 = contrast_vec),
                               method = "loso")
res_contrast$values$anchor1
```

## Using the pipeline helper

The same workflow integrates with `dkge_pipeline()` by supplying the descriptor via the new `input` argument. All downstream services (contrasts, classification, inference, transport) operate exactly as with design-level inputs.

```{r pipeline}
pipeline_res <- dkge_pipeline(input = anchor_input,
                               contrasts = list(anchor1 = contrast_vec),
                               method = "analytic",
                               inference = NULL)
summary(pipeline_res$contrasts)
```

## Classification targets

Anchor-based fits do not store the design-factor mapping that
`dkge_targets()` relies on. When you want to classify anchor effects you must
provide explicit weight matrices (rows = classes, columns = anchors) or
pre-built `dkge_target` objects. The helpers
`dkge_anchor_targets_from_prototypes()` and
`dkge_anchor_targets_from_directions()` turn feature-space prototypes or
directions into the required matrices.

```{r anchor-targets}
anchors_mat <- fit_anchor$provenance$anchors$anchors
proto_list <- list(
  classA = anchors_mat[c(1, 2), , drop = FALSE],
  classB = anchors_mat[c(3, 4), , drop = FALSE]
)
target_matrix <- dkge_anchor_targets_from_prototypes(anchors_mat, proto_list)
target_matrix

# Ready for classification
cls <- dkge_classify(fit_anchor,
                     targets = target_matrix,
                     method = "lda",
                     folds = 2)
cls$summary
```

Here `folds = 2` cross-validates the classifier conditional on the fitted anchor
representation and DKGE fit. It does not make anchor selection or basis fitting
fold-specific, so treat this chunk as an API example rather than an estimate of
end-to-end generalization performance.

## Diagnostics and provenance

Anchor coverage, leverage, and bandwidth settings are stored under `fit$provenance$anchors`. These diagnostics are useful for checking whether the median heuristic and chosen number of anchors provide adequate coverage across subjects.

```{r diagnostics}
dkge_anchor_diagnostics(fit_anchor)
```

## Summary

The feature-anchored path aligns subjects with disjoint item sets without
imputing missing cells. It selects a shared anchor representation, orthonormalizes
that representation, and delegates the aligned kernels to the core DKGE fitter.
For confirmatory prediction, define anchors independently or wrap anchor
construction and model fitting in an outer resampling loop; the convenience
workflow shown here fixes anchors using all subjects.

### Special cases: shared and mixed item sets

- **All subjects share the same items.** The anchor pipeline re-expresses the
  common item kernel because every subject projects identical feature rows into
  the same anchor coordinates. You can keep the anchor path for consistency or
  fit the shared item kernel with `dkge_fit_from_kernels()`; these are different
  representations and should be compared empirically if the choice matters.
- **Subgroups with identical items.** Subjects with identical feature rows receive
  projections into the same pooled anchor basis. Coverage and leverage diagnostics
  in `fit$provenance$anchors` help reveal uneven representation; large leverage
  spikes indicate anchors dominated by a subset and may motivate a smaller `L` or
  tighter bandwidth. These diagnostics do not substitute for nested preprocessing
  when the target is out-of-sample performance.
