---
title: "Between-Subject Multivariate Modeling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Between-Subject Multivariate Modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(dkge)
set.seed(2026)
```

If you have one brain-derived representation per subject and want to test group, trait, or covariate effects jointly, the between-subject DKGE layer gives you a supervised multivariate model instead of a voxel-by-voxel second-level GLM.

The workflow has three steps:

1. build a subject-by-feature target with `dkge_make_target()`
2. build a second-level design with `dkge_subject_model()`
3. fit and test a reduced-rank model with `dkge_between_rrr()` and `dkge_between_permute()`

## What does the second-level model estimate?

The between-subject model is a reduced-rank regression:

\[
Y = X B + E, \qquad \mathrm{rank}(B) \le r
\]

Here `Y` is a subject-by-feature matrix, `X` is the subject-level design matrix, and `B` contains distributed coefficient maps for your terms of interest.

## Start from a direct subject-by-feature matrix

In the simplest case you already have a common-space matrix with one row per subject.

```{r quick-target}
subject_data <- data.frame(
  subject_id = paste0("s", 1:18),
  group = factor(rep(c("control", "patient"), each = 9)),
  trait = seq(-1, 1, length.out = 18)
)

design <- dkge_subject_model(~ group * trait, subject_data)
X <- design$X

beta <- matrix(0, nrow = ncol(X), ncol = 6,
               dimnames = list(colnames(X), paste0("feature", 1:6)))
beta["grouppatient:trait", ] <- c(2, 1, 0, -1, -2, -1)

Y <- X %*% beta + matrix(rnorm(nrow(X) * ncol(beta), sd = 0.08), nrow(X), ncol(beta))
rownames(Y) <- subject_data$subject_id
colnames(Y) <- colnames(beta)

target <- dkge_make_target(Y = Y)
fit_between <- dkge_between_rrr(target, design, rank = 1)
```

You can inspect the fitted term map directly.

```{r quick-term-map}
round(dkge_term_map(fit_between, "grouppatient:trait"), 2)
```

And you can run a permutation test that keeps the formula structure intact.

```{r quick-permutation}
perm_between <- dkge_between_permute(
  fit_between,
  terms = c("group", "trait", "group:trait"),
  B = 99,
  seed = 42,
  scope = "both",
  feature_adjust = "maxT"
)

perm_between$summary[, c("term", "statistic", "p", "p_adjusted")]
```

```{r quick-permutation-checks, include = FALSE}
stopifnot(all(is.finite(fitted(fit_between))))
stopifnot(all(is.finite(residuals(fit_between))))
stopifnot(all(perm_between$summary$p >= 0 & perm_between$summary$p <= 1))
stopifnot(all(perm_between$feature_tests[["group:trait"]]$p_adjusted >= 0))
```

## Build the target from a DKGE fit

The more DKGE-native path is to let DKGE construct the subject representation first, then model subjects in that common target space.

This example starts from heterogeneous subject maps, fits DKGE, transports one contrast to a medoid subject, and then applies the between-subject model.

```{r dkge-fit}
toy <- dkge_sim_toy(
  factors = list(condition = list(L = 2), phase = list(L = 2)),
  active_terms = c("condition", "phase"),
  S = 6,
  P = c(4, 5, 3, 6, 4, 5),
  snr = 6,
  seed = 9
)

effects <- toy$info$term_names
toy$B_list <- lapply(toy$B_list, function(B) {
  rownames(B) <- effects
  B
})
toy$X_list <- lapply(toy$X_list, function(X) {
  colnames(X) <- effects
  X
})
centroids <- lapply(toy$B_list, function(B) matrix(rnorm(ncol(B) * 3), ncol(B), 3))

fit_dkge <- dkge_fit(
  dkge_data(toy$B_list, designs = toy$X_list, subject_ids = toy$subject_ids),
  K = toy$K,
  rank = 2
)
```

```{r transported-target}
transport <- dkge_transport_spec(
  centroids = centroids,
  medoid = 1L,
  method = "sinkhorn",
  epsilon = 0.1
)

contrast <- matrix(c(1, 0, 0), ncol = 1, dimnames = list(NULL, "condition"))
rownames(contrast) <- effects

transported_target <- dkge_make_target(
  fit_dkge,
  type = "transported_maps",
  contrast = contrast,
  transport = transport,
  crossfit = "analytic"
)

dim(transported_target$Y)
```

Now the second-level model looks the same as before.

```{r dkge-between}
subject_meta <- data.frame(
  subject_id = toy$subject_ids,
  group = factor(c("control", "control", "control", "patient", "patient", "patient")),
  trait = c(-0.8, -0.3, 0.2, -0.1, 0.5, 1.0)
)

between_design <- dkge_subject_model(~ group * trait, subject_meta)
fit_transport <- dkge_between_rrr(transported_target, between_design, rank = 1)
perm_transport <- dkge_between_permute(
  fit_transport,
  terms = c("group", "trait", "group:trait"),
  B = 49,
  seed = 7
)

perm_transport$summary[, c("term", "p")]
```

```{r dkge-between-checks, include = FALSE}
stopifnot(nrow(transported_target$Y) == nrow(subject_meta))
stopifnot(all(is.finite(predict(fit_transport))))
stopifnot(all(perm_transport$summary$p >= 0 & perm_transport$summary$p <= 1))
```

## What should you inspect after fitting?

There are three outputs worth checking first:

- `coef(fit_transport)` gives the full coefficient matrix in target space
- `dkge_term_map(fit_transport, "group:trait")` extracts the term-specific map
- `perm_transport$summary` gives a global multivariate permutation test for each term

If you asked for featurewise inference, `perm_between$feature_tests[["group:trait"]]` adds per-feature statistics and adjusted p-values in the same target space.

## Where to go next

Use `type = "component_scores"` in `dkge_make_target()` when you want a compact latent target, and use `type = "transported_maps"` when spatial interpretation in a common map space matters.
