---
title: Design Kernels and Model Tuning
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Design Kernels and Model Tuning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(dkge)
library(ggplot2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
options(
  tibble.print_min = 4,
  tibble.print_max = 4,
  digits = 3
)
set.seed(2024)
theme_set(theme_dkge())
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```


Two analysts can fit the same beta matrices and obtain different latent
representations because they encoded different relationships among effects.
That choice is the design kernel. It should express a scientific belief you can
name—for example, that adjacent ordinal levels are similar—not serve as an
opaque tuning matrix.

This page answers three practical questions: what a kernel matrix means, how to
compare defensible candidates on held-out subjects, and what ridge
regularization can and cannot repair. The working sequence is:

```text
declare factors and terms -> inspect the resulting similarity matrix -> compare kernel x rank candidates -> refit once
```

If you do not yet have aligned `q` by `P_s` beta matrices, read
`vignette("dkge-workflow")` first. The goal here is model selection within a
fixed input contract, not a second introduction to the entire pipeline.

## What data will the candidates be judged on?

```{r data}
library(dkge)
S <- 4; q <- 4; P <- 30; T <- 80

betas <- replicate(S, matrix(rnorm(q * P, sd = 0.5), q, P), simplify = FALSE)
true_kernel <- 0.9 * (diag(q) + 0.3)
cholK <- chol(true_kernel)
betas <- lapply(betas, function(B) cholK %*% B + 0.1 * matrix(rnorm(q * P), q, P))

designs <- replicate(S, {
  X <- matrix(rnorm(T * q), T, q)
  qr.Q(qr(X))
}, simplify = FALSE)

subjects <- lapply(seq_len(S), function(s) dkge_subject(betas[[s]], designs[[s]], id = paste0("sub", s)))
bundle <- dkge_data(subjects)
```

`dkge_data()` aligns effect order across subjects and records the common
metadata. Every candidate below sees these same beta and design matrices; only
the effect-space geometry changes.

## What does a design kernel look like?

```{r kernels}
factors <- list(
  A = list(L = 2, type = "nominal"),
  B = list(L = 2, type = "nominal")
)
terms <- list("A", "B", c("A", "B"))
K_struct <- design_kernel(
  factors,
  terms = terms,
  rho = c("A" = 1, "B" = 1, "A:B" = 0.4),
  basis = "cell",
  normalize = "unit_trace"
)

round(K_struct$K, 2)
```

Rows and columns are effect-coded coordinates. Bright off-diagonal cells mean
that two coordinates are treated as similar by the declared factor terms;
near-zero cells leave them largely uncoupled.

```{r kernel-heatmap, echo=FALSE, fig.height=4.4, fig.width=5.2, fig.alt="Heatmap of the structured design kernel for the two-by-two factorial example."}
kernel_df <- as.data.frame(as.table(K_struct$K), stringsAsFactors = FALSE)
names(kernel_df) <- c("row", "column", "similarity")
ggplot(kernel_df, aes(column, row, fill = similarity)) +
  geom_tile(colour = "white", linewidth = 0.35) +
  scale_fill_gradient2(low = "#315f8c", mid = "white", high = "#b33b2e") +
  coord_equal() +
  labs(
    title = "Similarity implied by the declared design",
    x = "Effect coordinate", y = "Effect coordinate", fill = "K"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Each term contributes one similarity component. `rho` controls its relative
strength: here the two main effects receive weight 1 and their interaction
receives weight 0.4. Those numbers encode a preference, not an empirical
finding. `design_kernel()` also retains the factor and term metadata needed by
target and cross-validation helpers.

## Which rank and kernel predict held-out structure?

`dkge_cv_kernel_rank()` first screens named kernels by alignment, then evaluates
the retained kernel-rank pairs with leave-one-subject-out explained energy. It
returns the selected pair and the tables needed to audit that choice.

```{r kernel-cv, message=FALSE}
rho_vals <- seq(0.4, 1.0, by = 0.3)
K_grid <- list()
for (ra in rho_vals) {
  for (rb in rho_vals) {
    nm <- sprintf("A%.1f_B%.1f", ra, rb)
    K_grid[[nm]] <- design_kernel(
      factors,
      terms = terms,
      rho = c("A" = ra, "B" = rb, "A:B" = 0.4),
      basis = "cell",
      normalize = "unit_trace"
    )$K
  }
}

selection <- dkge_cv_kernel_rank(
  bundle$betas, bundle$designs,
  K_grid = K_grid,
  ranks = 1:3,
  top_k = length(K_grid)
)
selection$picks_per_kernel
```

The score is the fraction of a held-out subject's K-metric energy captured by
the training-fold basis. It is descriptive predictive evidence, not a
significance test. Higher values mean that the fitted subspace captures more of
the held-out energy. The selector applies its documented one-standard-error
rule, preferring a lower rank among eligible pairs rather than blindly taking
the largest observed mean.

The full rank curves reveal more than the final pick:

```{r kernel-cv-plot, echo=FALSE, fig.height=4.2, fig.alt="Cross-validation curves with uncertainty bars showing held-out explained energy by rank for each candidate design kernel."}
cv_scores <- selection$tables$cv
ggplot(cv_scores, aes(rank, mean, colour = kernel, group = kernel)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.08, alpha = 0.55) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = sort(unique(cv_scores$rank))) +
  labs(
    title = "Held-out energy, not in-sample fit, chooses the model",
    x = "Retained rank", y = "Mean held-out explained energy", colour = "Kernel"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right")
```

Look for a curve that is both high and reasonably stable across folds. A tiny
advantage at a larger rank may not justify the added complexity; that is why
the selector applies the one-standard-error rule.

```{r refit}
best <- selection$pick
best

K_best <- K_grid[[best$kernel]]
fit <- dkge(bundle, K = K_best, rank = best$rank)
round(fit$sdev, 3)
```

## What does ridge regularization do here?

The `ridge` argument adds an isotropic diagonal term to the fitted q-space
moment before eigendecomposition. For a fixed moment this shifts all
eigenvalues equally and leaves its eigenvectors unchanged; it does not identify
or selectively down-weight noisy effects. Encode effect-specific structure in
the kernel or input weights instead. Ridge is mainly a numerical guard for
near-singular moment calculations.

```{r ridge}
fit_ridge <- dkge(bundle, K = K_best, rank = best$rank, ridge = 0.2)
round(fit_ridge$sdev, 3)
```

## Summary

Use `design_kernel()` to encode effect relationships you can defend. If the
structure is uncertain, include `diag(q)` as a baseline rather than assuming a
structured kernel must win. Compare kernels and ranks with
`dkge_cv_kernel_rank()`, interpreting its scores as held-out explained energy.
Refit the selected configuration before running inference,
and treat ridge as isotropic numerical regularization rather than an
effect-selection device.
