---
title: Unbalanced and Partial Effect Spaces
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Unbalanced and Partial Effect Spaces}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(dkge)
set.seed(2026)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

DKGE can represent one global effect space when subjects contribute different
cells or estimate the same cells with very different trial counts. The output
is still a common q-dimensional group basis, but three distinct problems must
be handled separately:

| Problem | What changes | DKGE mechanism |
|---|---|---|
| A cell is absent for a subject | Whether an effect pair is observed | `observed_rows` and `missingness` |
| A cell has 2 trials for one subject and 20 for another | Precision of an observed effect | `dkge_effect_weights()` |
| Cell estimates contain finite-trial estimation noise | Expected diagonal noise in the second moment | `debias` |

Coverage, precision weighting, and debiasing are complementary. A count weight
does not remove estimation-noise bias; analytic subtraction does not decide
which subject should contribute more; and zero-filling alone does not record
that a cell was unobserved.

The key contract is:

- `dkge_data()` aligns subject-local effect rows to the union of all effect
  labels.
- Missing rows are rendered as zero rows in the aligned matrices.
- The original observation pattern is retained as `observed_rows`,
  `provenance$obs_mask`, and `provenance$pair_counts`.
- `dkge_fit(missingness = ...)` controls how the partial coverage is handled
  when the q-space covariance is accumulated.

## How do you declare a partial global grid?

Use `dkge_effect_grid()` to declare the global effect cells and which factors
are within- versus between-subject.

```{r grid}
grid <- dkge_effect_grid(
  factors = list(
    group = c("control", "patient"),
    task = c("A", "B"),
    measure = c("low", "high")
  ),
  scope = c(group = "between", task = "within", measure = "within"),
  block_factors = "group"
)

grid$cell_labels
```

`block_factors = "group"` requests independent group blocks in the design
kernel. Terms such as `task` and `measure` are replicated within each group
block rather than coupled across groups by an all-ones group factor.

```{r kernel}
kernel <- design_kernel(
  grid,
  terms = list(
    "group", "task", "measure",
    c("group", "task"),
    c("group", "measure"),
    c("task", "measure"),
    c("group", "task", "measure")
  ),
  basis = "cell",
  normalize = "none"
)

kernel$info$term_scope
kernel$info$block_factors
```

The kernel metadata classifies terms as `within`, `between`, or `mixed`.
Contrast helpers use this metadata to recommend the appropriate inference
route.

## How are subject-local rows aligned?

Each subject supplies only the four rows for their own group. `dkge_data()`
embeds those rows into the eight-row union and records which global rows were
actually observed.

```{r data}
cell4 <- c("A:low", "A:high", "B:low", "B:high")
subject_info <- data.frame(
  subject_id = paste0("s", 1:4),
  group = c("control", "control", "patient", "patient")
)

make_subject_beta <- function(group) {
  rows <- paste(group, sub(":.*$", "", cell4), sub("^.*:", "", cell4), sep = ":")
  B <- matrix(rnorm(4 * 12), nrow = 4,
              dimnames = list(rows, paste0("feature", 1:12)))
  X <- diag(4)
  colnames(X) <- rows
  list(B = B, X = X)
}

subjects <- lapply(subject_info$group, make_subject_beta)
betas <- lapply(subjects, `[[`, "B")
designs <- lapply(subjects, `[[`, "X")

dat <- dkge_data(betas, designs, subject_ids = subject_info$subject_id)
dat$effects
dat$observed_rows
dat$provenance$pair_counts
```

Controls observe rows 1-4 and patients observe rows 5-8. Cross-group row pairs
have zero pair counts.

## How should partial coverage enter the fit?

The default `missingness = "none"` preserves the historical zero-filled
accumulation. That compatibility setting is appropriate when every subject
observes every effect. For a genuinely partial global space, choose a policy
that uses the recorded coverage:

- `"mask"` zeros entries whose applicable coverage measure is below
  `min_pairs`: observed pair mass without effect weights, and Kish pair ESS
  when effect-precision weights are active.
- `"rescale"` returns a per-pair mean: it divides by observed pair mass
  without effect weights and by total pair precision when those weights are
  active.
- `"shrink"` blends that per-pair mean toward its diagonal according to pair
  coverage (pair ESS on the precision-weighted branch).

```{r fit}
fit <- dkge_fit(
  dat,
  K = kernel,
  rank = 2,
  w_method = "none",
  missingness = "mask",
  miss_args = list(min_pairs = 1)
)

fit$missingness
fit$pair_counts[1:4, 5:8]
```

Here too, `w_method = "none"` isolates the coverage policy. With the default
MFA subject scaling, zero filling still occurs first, but the later transformed
block energies also affect each subject's scalar contribution.

```{r partial-coverage-checks, include = FALSE}
stopifnot(identical(dim(fit$pair_counts), c(8L, 8L)))
stopifnot(all(fit$pair_counts[1:4, 5:8] == 0))
stopifnot(all(fit$pair_counts[1:4, 1:4] == 2))
```

### What the zero-filled rows do, and do not, do

Coverage policies act in *raw* effect space, before the pooled ruler `R` and
the design kernel mix rows. A subject's raw second moment
`B_s B_s'` therefore has exact zero rows and columns wherever that subject
observed nothing: placeholders contribute no energy of their own. Zero filling
also precedes subject-weight derivation, but a non-`"none"` `w_method` scores
the transformed block after the pooled ruler `R` and `Khalf` have mixed effect
coordinates. Subject weights therefore do not evaluate each raw zero row in
isolation.

They are not, however, insulated from the metric. The fit embeds each moment as
`K^{1/2} R' (B_s B_s') R K^{1/2}`, and if `K` couples an observed cell to an
unobserved one, `K^{1/2}` will place some of the observed energy on the
unobserved coordinate. That is the K-metric doing its job — it is the same
smoothing that makes an ordinal or circular kernel useful — not a leak of
values the subject never supplied.

If you want strict separation, say so in the kernel rather than in the
accumulator. `block_factors = "group"`, as used above, makes `K` block-diagonal
across groups, so `K^{1/2}` is block-diagonal too and no control-group energy
can reach a patient-group coordinate:

```{r block-diagonal}
Khalf <- kernel_roots(kernel$K)$Khalf
# Cross-group blocks of K^{1/2} are numerically zero.
max(abs(Khalf[1:4, 5:8]))
```

```{r block-diagonal-check, include = FALSE}
stopifnot(max(abs(Khalf[1:4, 5:8])) < 1e-12)
```

## How do you fit an unbalanced 3 x 5 x 4 trialwise design?

Now consider a fully within-subject `condition x delay x response` design with
3, 5, and 4 levels: 60 possible cells per subject. The response is ordinal, and
trial counts vary by subject and cell. `dkge_effect_grid()` pins the global row
order while `design_kernel()` says that adjacent response levels are more
similar than distant ones.

```{r trial-grid}
grid60 <- dkge_effect_grid(
  factors = list(
    condition = c("c1", "c2", "c3"),
    delay = paste0("d", 1:5),
    response = list(L = 4, type = "ordinal",
                    levels = as.character(1:4), l = 1)
  )
)

kernel60 <- design_kernel(
  grid60,
  terms = list(
    "condition", "delay", "response",
    c("condition", "delay"), c("condition", "response"),
    c("delay", "response"), c("condition", "delay", "response")
  ),
  basis = "cell",
  normalize = "unit_trace"
)

c(q = length(grid60$cell_labels), kernel_rows = nrow(kernel60$K))
```

```{r trial-grid-checks, include = FALSE}
stopifnot(length(grid60$cell_labels) == 60L)
stopifnot(identical(dim(kernel60$K), c(60L, 60L)))
```

The trialwise constructor fits \(Y_s = X_s B_s + E_s\) for each subject. It
retains \(B_s\), \((X_s^\top X_s)^{-1}\), residual variances, and one-hot cell
counts, but not the full trial-by-feature response. Here every cell has at least
three trials so that the within-cell split leaves both halves estimable.

```{r trialwise-data, include = FALSE}
q <- length(grid60$cell_labels)
P <- 10L
condition_score <- c(-1, 0, 1)[match(grid60$cells$condition,
                                     c("c1", "c2", "c3"))]
delay_score <- seq(-1, 1, length.out = 5)[match(grid60$cells$delay,
                                                paste0("d", 1:5))]
response_score <- as.numeric(grid60$cells$response) - 2.5
latent_effects <- cbind(condition_score, delay_score, response_score,
                        condition_score * response_score)
set.seed(810)
feature_patterns <- matrix(rnorm(ncol(latent_effects) * P),
                           ncol(latent_effects), P)
true_beta <- latent_effects %*% feature_patterns / 3
colnames(true_beta) <- paste0("feature", seq_len(P))
rownames(true_beta) <- grid60$cell_labels

make_trial_subject <- function(subject_index) {
  set.seed(810 + subject_index)
  counts <- sample(3:10, q, replace = TRUE)
  cell_index <- rep(seq_len(q), counts)
  X <- matrix(0, length(cell_index), q)
  X[cbind(seq_along(cell_index), cell_index)] <- 1
  colnames(X) <- grid60$cell_labels
  Y <- X %*% true_beta + matrix(rnorm(length(cell_index) * P, sd = 0.8),
                                length(cell_index), P)
  colnames(Y) <- colnames(true_beta)
  dkge_trial_subject(Y, X, id = paste0("s", subject_index),
                     split = "within_cell")
}

trial_subjects <- lapply(seq_len(5), make_trial_subject)
trial_data <- dkge_data(trial_subjects, effects = grid60$cell_labels)
count_table <- do.call(rbind, lapply(trial_subjects, function(subject) {
  c(total_trials = sum(subject$effect_n),
    min_cell = min(subject$effect_n),
    median_cell = stats::median(subject$effect_n),
    max_cell = max(subject$effect_n))
}))
rownames(count_table) <- paste0("s", seq_len(nrow(count_table)))
```

```{r trial-counts, echo = FALSE}
knitr::kable(count_table, digits = 1)
```

Each subject contributes a different count profile, but all are aligned to the
same 60 labels. The group fit below makes three choices explicit:

1. `effect_scaling = "none"` keeps cell means in their common beta units rather
   than applying the pooled design ruler.
2. `effect_weights = dkge_effect_weights("count")` gives more influence to
   better-estimated subject-by-cell rows.
3. `debias = "analytic"` subtracts the expected finite-trial noise moment before
   pooling.

`w_method = "none"` is deliberate in this diagnostic example: it isolates
cell-level precision weighting and debiasing from the package's default
subject-level MFA scaling. It is not a general recommendation to disable
subject weighting.

```{r trial-fit}
fit_analytic <- dkge_fit(
  trial_data,
  K = kernel60,
  rank = 3,
  w_method = "none",
  effect_scaling = "none",
  effect_weights = dkge_effect_weights("count"),
  debias = "analytic",
  missingness = "none"
)

fit_analytic$rank
```

For cells \(c\) and \(c'\), count weighting uses pair reliability
\(\sqrt{n_{sc} n_{sc'}}\). DKGE first computes the precision-weighted mean for
each pair, then restores the cohort scale. `fit$pair_ess` is Kish's effective
number of contributing subjects; it falls when one or two subjects dominate a
cell pair even if every subject observed it.

```{r trial-diagnostics}
pair_diagnostics <- data.frame(
  diagnostic = c("minimum pair ESS", "maximum pair ESS",
                 "negative raw-effect mass", "negative transformed mass"),
  value = c(min(fit_analytic$pair_ess), max(fit_analytic$pair_ess),
            fit_analytic$moment_diagnostics$effect$negative_mass,
            fit_analytic$moment_diagnostics$transformed$negative_mass)
)
knitr::kable(pair_diagnostics, digits = 3)
```

```{r trial-fit-checks, include = FALSE}
stopifnot(identical(dim(fit_analytic$effect_moment), c(60L, 60L)))
stopifnot(all(is.finite(fit_analytic$effect_moment)))
stopifnot(all(is.finite(fit_analytic$pair_ess)))
stopifnot(min(fit_analytic$pair_ess) > 1)
stopifnot(max(fit_analytic$pair_ess) <= length(trial_subjects) + 1e-10)
stopifnot(all(is.finite(unlist(fit_analytic$moment_diagnostics))))
stopifnot(pair_diagnostics$value[1] < pair_diagnostics$value[2])
```

### What exactly does debiasing change?

Finite-trial noise can be addressed in either of two ways:

- `debias = "analytic"` subtracts
  `noise_trace * (X'X)^{-1}` per subject. The noise trace includes residual
  variance and any diagonal spatial weights.
- `debias = "split_half"` replaces the raw second moment with the symmetrized
  cross-product of two stored half estimates. Independent half-errors then have
  zero expected cross-product.

The chunked constructor uses the same weighted analytic correction as the
dense constructor. In particular, a non-unit spatial `omega` is included when
the noise trace is reconstructed from per-feature residual variances; an
unweighted cached trace is not reused as if it were already weighted.

```{r chunked-trace}
set.seed(19411)
X_chunk <- model.matrix(~ 0 + factor(rep(1:2, each = 6)))
colnames(X_chunk) <- c("e1", "e2")
omega_chunk <- c(0.2, 1, 3, 0.5)
make_chunk_y <- function() {
  truth <- matrix(c(1, -0.5, 0.25, 2, 0.4, -1, 0.7, 0.1), 2, 4)
  Y <- X_chunk %*% truth +
    matrix(rnorm(nrow(X_chunk) * 4, sd = 0.35), nrow(X_chunk), 4)
  colnames(Y) <- paste0("v", 1:4)
  Y
}
Y_chunk <- list(make_chunk_y(), make_chunk_y())
dense_subjects <- lapply(seq_along(Y_chunk), function(s) {
  dkge_trial_subject(Y_chunk[[s]], X_chunk, id = paste0("s", s),
                     omega = omega_chunk)
})
chunked_subjects <- lapply(seq_along(Y_chunk), function(s) {
  Y <- Y_chunk[[s]]
  dkge_trial_subject_chunks(
    list(Y[, 1:2, drop = FALSE], Y[, 3:4, drop = FALSE]),
    X_chunk,
    id = paste0("s", s),
    omega = omega_chunk
  )
})
K_chunk <- diag(2)
dimnames(K_chunk) <- list(colnames(X_chunk), colnames(X_chunk))
fit_dense <- dkge_fit(
  dkge_data(dense_subjects), K = K_chunk, rank = 1,
  w_method = "none", effect_scaling = "none", debias = "analytic"
)
fit_chunked <- dkge_fit(
  dkge_data(chunked_subjects), K = K_chunk, rank = 1,
  w_method = "none", effect_scaling = "none", debias = "analytic"
)
chunked_check <- data.frame(
  weighted_noise_trace = sum(
    omega_chunk * dense_subjects[[1]]$residual_variance
  ),
  max_abs_Chat_difference = max(abs(fit_dense$Chat - fit_chunked$Chat))
)
knitr::kable(chunked_check, digits = 12)
```

```{r chunked-trace-checks, include = FALSE}
expected_trace <- sum(omega_chunk * dense_subjects[[1]]$residual_variance)
stopifnot(isTRUE(all.equal(
  chunked_check$weighted_noise_trace,
  expected_trace,
  tolerance = 1e-12
)))
stopifnot(max(abs(fit_dense$Chat - fit_chunked$Chat)) < 1e-12)
```

Because the subjects above stored within-cell halves, the alternative fit is
runnable with the same data:

```{r split-fit}
fit_split <- dkge_fit(
  trial_data,
  K = kernel60,
  rank = 3,
  w_method = "none",
  effect_scaling = "none",
  effect_weights = dkge_effect_weights("count"),
  debias = "split_half"
)

c(analytic = fit_analytic$moment_diagnostics$effect$negative_mass,
  split_half = fit_split$moment_diagnostics$effect$negative_mass)
```

Here the split-half estimate has more negative spectral mass (about 193 versus
146), but that ordering is not a performance score. Both numbers diagnose the
finite-sample indefiniteness of their respective moment estimators; choosing
between them depends on whether the split errors are credibly independent and
whether the analytic covariance model is credible.

```{r split-fit-checks, include = FALSE}
negative_mass <- c(
  analytic = fit_analytic$moment_diagnostics$effect$negative_mass,
  split_half = fit_split$moment_diagnostics$effect$negative_mass
)
stopifnot(identical(unname(round(negative_mass)), c(146, 193)))
stopifnot(negative_mass[["split_half"]] > negative_mass[["analytic"]])
```

Analytic subtraction and pair normalization can produce an indefinite
q-by-q estimate. DKGE therefore uses a symmetric eigendecomposition, retains
the leading positive eigenpairs, and exposes negative spectral mass through
`fit$moment_diagnostics`; an SVD would incorrectly turn negative directions
into positive components.

The constructor's `split = "within_cell"` alternates trials within each cell;
it does not prove that the two half-errors are independent. If runs, sessions,
or temporal dependence define independence in your experiment, use an
appropriately constructed split outside this convenience path or prefer the
analytic estimator with a justified covariance model.

## Which contrasts are estimable within subject?

Contrasts named after kernel terms are tagged with the term scope.

```{r contrasts}
task4 <- c(-0.5, -0.5, 0.5, 0.5)
contrasts <- list(
  task = rep(task4, 2),
  group = c(rep(-0.25, 4), rep(0.25, 4)),
  "group:task" = c(-task4, task4)
)

task_res <- dkge_contrast(fit, contrasts["task"], method = "loso", align = FALSE)
knitr::kable(task_res$metadata$contrast_estimability)
```

```{r task-estimability-checks, include = FALSE}
stopifnot(identical(
  unname(task_res$metadata$contrast_estimability$estimability),
  "within"
))
```

`task` is a within-subject contrast, so LOSO/k-fold contrast inference is the
right family. `group` and `group:task` are between or mixed effects. DKGE will
still return descriptive cross-fitted maps, but the confirmatory route is the
between-subject RRR layer: construct a subject model with
`dkge_subject_model()`, then call `dkge_between_rrr()` and assess it with
`dkge_between_permute()`.

```{r group-contrast, warning = TRUE}
group_res <- dkge_contrast(
  fit, contrasts[c("group", "group:task")],
  method = "loso", align = FALSE
)
knitr::kable(group_res$metadata$contrast_estimability)
```

```{r group-estimability-checks, include = FALSE}
stopifnot(identical(
  unname(group_res$metadata$contrast_estimability$estimability),
  c("between", "mixed")
))
stopifnot(all(grepl(
  "dkge_between",
  group_res$metadata$contrast_estimability$recommended_inference,
  fixed = TRUE
)))
```

## What is deliberately outside this workflow?

A supervised tensor or CP/Tucker model could be useful for a research extension,
especially when the goal is to discover a latent group-conditioned expression
pattern. That is not part of the DKGE core contract here. The core feature is
partial global effect spaces with explicit coverage, q-space missingness
policies, and term-scope-aware inference recommendations.

For group, trait, or mixed-effect tests on the resulting subject
representations, continue with `vignette("dkge-between-subjects")`. For a
conceptual map of component-, contrast-, and feature-level claims, see
`vignette("dkge-concepts")`.
