---
title: 'Model-Level Spatial Regularization in DKGE'
description: >
  Build a graph Laplacian with adjoin, let smoothing change the fitted DKGE
  solution, tune its strength on raw held-out fields, and carry it into inference.
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Model-Level Spatial Regularization in DKGE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(dkge)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4,
  out.width = "90%"
)
set.seed(20240824)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

Suppose each subject contributes a $q \times P_s$ GLM beta matrix: the rows
are the same $q$ effects, while the columns are that subject's voxels or
parcels. DKGE solves its eigensystem in the small $q$-dimensional effect
space. That does **not** mean spatial regularization must be a cosmetic step
after fitting. A spatial operator applied to the beta columns changes the
effect-space moment and therefore can change the learned basis itself.

This article follows one estimand:

> Learn the rank-one effect direction that best explains spatial beta fields
> after a prespecified graph-Laplacian regularization, then obtain held-out
> subject fields under that same regularized model.

The expected outputs are a $q \times 1$ group basis, one regularized loading
field per subject, a held-out score for each candidate penalty, and LOSO
contrast fields that inherit the selected operator.

```text
q x P_s beta fields -> sparse spatial solve -> q x q pooled moment -> group basis -> held-out fields
```

## Where does the regularizer enter?

Let $L_s$ be a combinatorial graph Laplacian on subject $s$'s spatial units.
DKGE uses the Tikhonov resolvent

\[
H_{s,\lambda} = (I + \lambda L_s)^{-1}, \qquad \lambda \ge 0,
\]

and replaces the design-scaled beta block $\widetilde B_s$ by
$\widetilde B_s H_{s,\lambda}^{\mathsf T}$. With a spatial metric
$\Omega_s$, the subject contribution becomes

\[
M_s(\lambda) =
\widetilde B_s H_{s,\lambda}^{\mathsf T}\,
\Omega_s\,
H_{s,\lambda}\widetilde B_s^{\mathsf T}.
\]

Thus $\lambda$ changes the pooled $q \times q$ moment that DKGE
diagonalizes. The same $H_{s,\lambda}$ is then used for components,
cross-fitted contrasts, bootstrap inputs, transport inputs, and prediction. It
is never necessary to form a dense inverse: DKGE stores a sparse Cholesky
factorization and solves against beta fields as needed. At $\lambda=0$, the
operation is exactly the identity.

The graph and the design kernel play different roles. $L_s$ says which
**spatial units** should vary together. $K$ says which **experimental
effects** should vary together. Neither can substitute for the other.

## Build an example with a rough distractor

The example has two effects at 40 ordered locations. The first effect varies
smoothly; the second alternates at every location and has larger amplitude.
An unsmoothed rank-one fit therefore prefers the rough field.

```{r make-spatial-data}
S <- 6L
P <- 40L
effects <- c("smooth", "rough")
locations <- sprintf("location%02d", seq_len(P))

coords <- cbind(x = seq_len(P) - 1, y = 0, z = 0)
rownames(coords) <- locations

smooth_truth <- sin(seq(0, 2 * pi, length.out = P))
rough_truth <- 1.5 * (-1)^seq_len(P)

B_list <- setNames(lapply(seq_len(S), function(s) {
  B <- rbind(
    smooth = smooth_truth + rnorm(P, sd = 0.08),
    rough = rough_truth + rnorm(P, sd = 0.08)
  )
  colnames(B) <- locations
  B
}), sprintf("sub%02d", seq_len(S)))

X0 <- diag(2)
colnames(X0) <- effects
X_list <- setNames(replicate(S, X0, simplify = FALSE), names(B_list))

K <- diag(2)
dimnames(K) <- list(effects, effects)

c(beta_rows = nrow(B_list[[1]]), spatial_units = ncol(B_list[[1]]))
```

```{r spatial-data-checks, include=FALSE}
stopifnot(length(B_list) == S)
stopifnot(all(vapply(B_list, nrow, integer(1)) == length(effects)))
stopifnot(all(vapply(B_list, ncol, integer(1)) == P))
stopifnot(identical(colnames(B_list[[1]]), rownames(coords)))
```

## Construct the graph with `adjoin`

`dkge_spatial_regularizer()` delegates coordinate-based graph construction to
`adjoin::spatial_laplacian()`. DKGE then validates the Laplacian, binds it to
the beta-column domain, and records the graph recipe and `adjoin` version.
Here a distance just above one connects adjacent locations on the line.

```{r construct-spatial-regularizer}
spatial_demo <- dkge_spatial_regularizer(
  coords = coords,
  lambda = 2,
  dthresh = 1.01,
  nnk = 3,
  weight_mode = "binary",
  normalized = FALSE,
  handle_isolates = "keep_zero"
)
spatial_demo
```

Because `coords` is one matrix, this is a shared spatial domain: every subject
must have the same number and ordering of columns. The matching location names
make that contract explicit. If subjects have different parcellations, supply
a named list of coordinate matrices or precomputed Laplacians instead; DKGE
will bind one operator to each subject and fail on missing or mismatched
domains. Names on that list are matched to the fitted subject IDs. For raw
lists, declare them explicitly with
`dkge_data(B_list, X_list, subject_ids = names(B_list))` before fitting.

The combinatorial Laplacian is intentional. It preserves constant fields, so
DKGE currently requires `normalized = FALSE`. For a surface mesh or a
scientifically defined adjacency, construct the Laplacian in `adjoin` and pass
it through `laplacian =` rather than approximating the topology with Euclidean
distance.

## Verify that smoothing changes the solution

We first compare an unsmoothed fit with a deliberately visible penalty. Both
use the same identity design kernel and disable subject weighting, so the only
changed assumption is the spatial regularizer.

```{r fit-raw-and-spatial}
fit_raw <- dkge_fit(
  B_list, X_list, K,
  rank = 1,
  w_method = "none",
  effect_scaling = "none"
)

fit_demo <- dkge_fit(
  B_list, X_list, K,
  rank = 1,
  w_method = "none",
  effect_scaling = "none",
  spatial = spatial_demo
)

effect_comparison <- data.frame(
  effect = effects,
  raw = abs(fit_raw$U[, 1]),
  regularized = abs(fit_demo$U[, 1])
)
transform(effect_comparison,
          raw = round(raw, 3), regularized = round(regularized, 3))
```

Component signs are arbitrary, hence the absolute values. The raw component
is almost entirely the high-amplitude rough effect. Under the regularized
moment, the dominant direction rotates toward the smooth effect. This is the
key distinction from post-hoc image smoothing: the fitted $U$ changed.

The reconstructed subject field uses the same operator. For a compact spatial
diagnostic, define graph-path roughness as the sum of squared adjacent
differences.

```{r compare-spatial-fields}
roughness <- function(x) sum(diff(x)^2)

map_raw <- dkge_project_btil(fit_raw, fit_raw$Btil[[1]])[, 1]
map_demo <- dkge_project_btil(
  fit_demo, fit_demo$Btil[[1]], subject = 1
)[, 1]

map_raw <- map_raw * sign(stats::cor(map_raw, rough_truth))
map_demo <- map_demo * sign(stats::cor(map_demo, smooth_truth))

round(c(raw = roughness(map_raw), regularized = roughness(map_demo)), 3)
```

```{r plot-spatial-fields, echo=FALSE, fig.cap="A deliberately strong spatial penalty rotates the rank-one solution away from the alternating field and yields a smooth loading field. Component signs were aligned only for display.", fig.alt="Two line plots: the unsmoothed rank-one loading alternates sharply at adjacent locations, while the regularized loading varies smoothly."}
map_frame <- rbind(
  data.frame(location = seq_len(P), loading = map_raw,
             fit = "No spatial regularization"),
  data.frame(location = seq_len(P), loading = map_demo,
             fit = "Laplacian regularization (lambda = 2)")
)
ggplot2::ggplot(map_frame, ggplot2::aes(location, loading)) +
  ggplot2::geom_hline(yintercept = 0, colour = "grey80") +
  ggplot2::geom_line(colour = "#b33b2e", linewidth = 0.65) +
  ggplot2::facet_wrap(~ fit, ncol = 1, scales = "free_y") +
  ggplot2::labs(
    title = "The spatial operator changes both the basis and its loading field",
    x = "Ordered spatial unit", y = "Component loading"
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

```{r solution-change-checks, include=FALSE}
stopifnot(abs(fit_raw$U[match("rough", effects), 1]) > 0.99)
stopifnot(abs(fit_demo$U[match("smooth", effects), 1]) > 0.99)
stopifnot(roughness(map_demo) < 0.01 * roughness(map_raw))
stopifnot(all(is.finite(map_demo)))
```

The fitted object retains auditable domain and graph diagnostics. Identical
fingerprints here are expected because all subjects share one ordered graph.

```{r inspect-spatial-provenance}
spatial_diagnostics <- dkge_diagnostics(fit_demo)$spatial
spatial_diagnostics$diagnostics[
  , c("subject", "n_units", "n_edges", "lambda", "domain_mode")
]
```

## Select lambda on raw held-out fields

The visible $\lambda=2$ comparison was chosen to explain the mechanism, not
to justify that penalty. Use `dkge_cv_spatial_grid()` when $\lambda$ is to be
selected from the data. Keep the graph, design kernel, rank, and validation
geometry fixed, and include zero as the unsmoothed candidate.

```{r select-spatial-lambda}
cv_spatial <- dkge_cv_spatial_grid(
  B_list, X_list, K,
  spatial = spatial_demo,
  lambdas = c(0, 0.1, 0.5, 2),
  rank = 1,
  w_method = "none",
  effect_scaling = "none"
)

transform(
  cv_spatial$table[, c("lambda", "mean", "se", "admissible")],
  mean = round(mean, 4),
  se = round(se, 4)
)
cv_spatial$pick
```

Each candidate basis is trained without the held-out subject, but all
candidates are scored against the **unsmoothed** held-out beta block in one
fixed effect-space geometry. Prespecified effect scaling, location weights,
and `Omega_list` remain in that fixed geometry; only the candidate Laplacian
solve is omitted. A candidate therefore cannot improve its own score by
shrinking the field in both the numerator and denominator. `best` is the
largest mean score; `pick` is the largest, smoothest $\lambda$ within one
standard error of that best score. In this example the selected value is
`r cv_spatial$pick`, much weaker than the explanatory $\lambda=2$.

```{r spatial-cv-checks, include=FALSE}
stopifnot(cv_spatial$pick %in% cv_spatial$table$lambda)
stopifnot(cv_spatial$best %in% cv_spatial$table$lambda)
stopifnot(all(cv_spatial$table$admissible))
stopifnot(identical(cv_spatial$heldout_geometry, "raw_beta_block"))
stopifnot(identical(cv_spatial$selection_rule,
                    "largest_lambda_within_one_se"))
```

As with kernel selection, a saturated warning means the held-out criterion
cannot discriminate the candidates well. Also remember that $\lambda$'s
scale depends on the graph weights: compare penalties only while holding the
graph construction fixed. If graph topology or edge weights are themselves
learned from the outcome, they require their own nested validation rather than
being tuned on the same folds.

Refit the complete cohort with the selected specification returned by CV:

```{r refit-selected-spatial-model}
fit_selected <- dkge_fit(
  B_list, X_list, K,
  rank = 1,
  w_method = "none",
  effect_scaling = "none",
  spatial = cv_spatial$spatial
)
dkge_diagnostics(fit_selected)$spatial[c("active", "lambda", "shared")]
```

## Carry the operator into LOSO inference

Do not smooth a final $t$- or $p$-map. Instead, construct held-out subject
fields under the fitted spatial model and perform inference on those fields.
`dkge_contrast()` does this automatically because the operator is part of the
fit.

```{r selected-loso-contrast}
rough_contrast <- c(smooth = 0, rough = 1)
loso_raw <- dkge_contrast(fit_raw, rough_contrast, method = "loso")
loso_selected <- dkge_contrast(
  fit_selected, rough_contrast, method = "loso"
)

loso_roughness <- data.frame(
  model = c("raw", "selected"),
  median_roughness = c(
    median(vapply(loso_raw$values[[1]], roughness, numeric(1))),
    median(vapply(loso_selected$values[[1]], roughness, numeric(1)))
  )
)
transform(loso_roughness,
          median_roughness = round(median_roughness, 2))
```

```{r selected-loso-checks, include=FALSE}
stopifnot(length(loso_selected$values[[1]]) == S)
stopifnot(all(vapply(loso_selected$values[[1]], length, integer(1)) == P))
stopifnot(loso_roughness$median_roughness[2] <
            loso_roughness$median_roughness[1])
```

Transport to a medoid, anchor field, or voxel grid happens **after** these
regularized subject fields are constructed. It aligns spatial domains; it does
not redefine the fitted effect-space solution.

## Keep the spatial controls distinct

Several DKGE arguments mention space, but they do not estimate the same thing.

| Control | What it changes | Changes fitted basis? |
|---|---|---:|
| `spatial = dkge_spatial_regularizer(...)` | Correlation across neighboring beta columns through $H_{s,\lambda}$ | Yes |
| `Omega_list` or adaptive location weights | Spatial mass, precision, or reliability in the subject moment | Yes |
| transport mapper and masses | Alignment of already estimated subject fields to a common domain | No |
| `dkge_anchor_aggregate(L=, lambda=)` | Optional smoothing of an already transported anchor average | No |

The operations can be combined, but order matters: DKGE applies adaptive
column scaling first, the Laplacian solve second, and `Omega_list` as the final
spatial metric in the moment. See `vignette("dkge-weighting")` for the complete
weighting pipeline.

Two further boundaries are deliberate:

* Active spatial regularization is not available with
  `debias = "analytic"`. The current analytic contract stores diagonal
  residual variances, while $H$ induces spatial noise covariance that those
  variances do not identify. DKGE fails closed; use `debias = "split_half"` or
  `"none"`.
* A shared graph is automatically reused for prediction. A fit with
  subject-specific graphs requires a new `spatial` specification aligned to
  the prediction subjects; smoothing cannot be added only at prediction time.
* `dkge_fit_from_kernels()` and anchor-input fits do not expose physical beta
  columns: their synthetic factor columns are algebraic, not spatial. They
  reject model-level `spatial` regularization rather than treating those
  factors as neighboring locations.

## Reporting checklist

For a reproducible analysis, report:

1. the spatial domain and whether the graph was shared or subject-specific;
2. how the graph was built (`dthresh`, `nnk`, edge weights, isolate policy, or
   the provenance of a precomputed Laplacian);
3. the candidate $\lambda$ grid, validation geometry, selection rule, and
   whether scores saturated;
4. the selected $\lambda$, latent rank, design kernel, `Omega_list`, and other
   weighting choices;
5. that inference used regularized held-out subject fields, with no direct
   smoothing of the final inferential statistic.

Continue with `vignette("dkge-contrasts-inference")` for the inferential
estimand, `vignette("dkge-weighting")` for the neighboring weighting controls,
or `vignette("dkge-dense-rendering")` for transport and rendering after the
regularized subject fields have been estimated.
