---
title: Adaptive Voxel Weighting in DKGE
description: >
  Define, inspect, and structurally check a training-fold voxel weighting
  rule with `dkge_weights()`.
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Adaptive Voxel Weighting in DKGE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(dkge)
library(ggplot2)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
options(
  tibble.print_min = 4,
  tibble.print_max = 4,
  digits = 3
)
set.seed(2024)
theme_set(theme_dkge())
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```



DKGE fits are driven by cross-subject second moments. Equal voxel weights can
let large noisy regions dominate those moments, but adaptive weights can also
create leakage if they use held-out data or change under a permutation. This
vignette shows how to define, inspect, and structurally check a weighting rule.
Those checks are necessary safeguards; they are not a substitute for
design-specific null calibration.

**See also:** `vignette("dkge-weighting")` covers the
three complementary weighting *layers* — spatial (`Omega_list`), subject-level
(`w_method`/`w_tau`), and transport (`sizes`). This vignette focuses on the
separate `dkge_weights()` API for training-fold *voxel-level* adaptive
weighting. The mechanisms enter different stages of the computation and can be
combined when their joint estimand is explicit.

---

The examples use small synthetic datasets so they run fast.

The examples use small synthetic datasets so they run fast.

## 1. Understanding the ingredients

The weight specification is created with `dkge_weights()`, which packages four
essential pieces of information that work together to define the weighting strategy:

- **Prior weights** describe locations you already trust from external
  information, such as a reliability map from an earlier study. They must be
  *sign invariant*: flipping the sign of a beta map must not change a voxel's
  variances all meet this important requirement and can be used safely as priors.
- **Adaptive statistics** are computed dynamically from the training subjects inside each
  cross-validation split. The built-in options examine *energy* (squared
  amplitude in the design metric), *precision* (inverse variance), or their
  product. Energy is $\|K^{1/2} B_{:\alpha}\|_2^2$, the signal a voxel carries in
  effect space; precision is $1/(\sigma^2 + \varepsilon)$, its measurement reliability
- **Combination rules** set how prior and adaptive components blend
  (product, sum, or override). `mix` controls the balance between them.
- **Shrinkage** keeps the final weights well-behaved: winsorize extremes, then
  everything back towards 1 to ensure that no single voxel dominates the fit.

With `scope = "fold"`, adaptive statistics are computed from the training
subjects in each fold. This prevents the held-out subject from entering the
weight estimate. It addresses that specific leakage path, but does not by
itself guarantee calibrated p-values or external generalisation.

## 2. Building weight specifications and fitting DKGE

The default configuration relies purely on adaptive statistics, with no prior
knowledge.

```{r make-weight}
library(dkge)

wts_adapt <- dkge_weights(
  prior   = NULL,            # no prior information
  adapt   = "kenergy_prec",  # energy × precision, a simple SNR proxy
  combine = "product",
  mix     = 0.6,             # adaptive term has more influence than the prior
  shrink  = list(alpha = 0.5, winsor = 0.99, normalize = "mean"),
  scope   = "fold"           # recomputed on the training subjects of each fold
)

print(wts_adapt)
```

The dataset below has ten subjects, four effects, and sixty voxels per subject.
The scale is arbitrary: it is small enough to run fast and large enough to show
the weighting strategies apart.

```{r toy-data}
make_subject <- function(id, q = 4L, v = 60L) {
  design <- diag(q)
  colnames(design) <- paste0("eff", seq_len(q))
  beta <- matrix(rnorm(q * v), nrow = q)
  dkge_subject(beta, design = design, id = paste0("sub", id))
}

subjects <- lapply(seq_len(10L), make_subject)
K <- diag(4)

fit_uniform  <- dkge(subjects, K = K, rank = 2, w_method = "none")
fit_adaptive <- dkge(subjects, K = K, rank = 2, weights = wts_adapt)
```

The second fit computes voxel weights for every training fold under the
adaptive rule. Inspect the global weights and the per-subject averages to see
how the fit adjusted each voxel's influence.

```{r inspect-weights}
head(fit_adaptive$voxel_weights)
dkge_diagnostics(fit_adaptive)$voxel_weights
```

```{r adaptive-weight-distribution, echo=FALSE, fig.height=3.6, fig.alt="Histogram of adaptive voxel weights centered near one after shrinkage."}
hist(
  fit_adaptive$voxel_weights, breaks = 14,
  col = "#d9e1e8", border = "white",
  main = "Shrinkage keeps adaptive weights controlled",
  xlab = "Voxel weight"
)
abline(v = 1, lty = 2, lwd = 2, col = "#b33b2e")
```

Expect modest deviations from 1.0 rather than spikes. That is what the
shrinkage parameters enforce: no single voxel dominates.

## 3. Mixing priors with adaptive statistics

Suppose you have an external reliability image from a previous study. Convert
it into a sign-invariant prior, then blend it with the adaptive component.
component is unchanged; it is now blended with that external knowledge.

```{r prior}
reliability_prior <- runif(60, min = 0.1, max = 1)
reliability_weight <- reliability_prior^2 + 1e-6

wts_mixed <- dkge_weights(
  prior   = reliability_weight,
  adapt   = "kenergy_prec",
  combine = "product",
  mix     = 0.4,   # stronger prior, weaker adaptive term
  shrink  = list(alpha = 0.6, winsor = 0.995)
)

fit_mixed <- dkge(subjects, K = K, rank = 2, weights = wts_mixed)
```

The squared transform is monotone on this non-negative reliability scale, so a
larger reliability value implies a larger prior weight. The adaptive term and
shrinkage can still modify the final ordering. Inspect that relationship rather
than assuming the prior had its intended effect:

```{r prior-diagnostics}
prior_check <- data.frame(
  reliability = reliability_prior,
  final_weight = fit_mixed$voxel_weights
)
stats::cor(prior_check$reliability, prior_check$final_weight,
           method = "spearman")
```

```{r prior-weight-plot, echo=FALSE, fig.height=3.8, fig.alt="Scatter plot showing the relationship between external reliability values and final mixed adaptive weights."}
plot(
  prior_check$reliability, prior_check$final_weight,
  pch = 19, col = grDevices::adjustcolor("#315f8c", alpha.f = 0.65),
  xlab = "External reliability", ylab = "Final mixed weight",
  main = "Did the prior influence the weights as intended?"
)
lines(stats::lowess(prior_check$reliability, prior_check$final_weight),
      col = "#b33b2e", lwd = 2)
```

The upward trend confirms that the prior still affects the final ordering,
while the scatter shows that the adaptive statistic and shrinkage also matter.
That is more informative than printing the correlation alone.

```{r prior-diagnostics-checks, include=FALSE}
stopifnot(all(reliability_weight > 0))
stopifnot(stats::cor(
  prior_check$reliability, prior_check$final_weight,
  method = "spearman"
) > 0)
```

## 4. Safety checks and post-hoc updates

### Sign-invariance check

For sign-flip inference, a fixed weighting rule should not change when every
beta is multiplied by -1. The following executable check verifies that
structural property for this example.

```{r sign-invariance}
subjects_flipped <- lapply(subjects, function(subject) {
  subject$beta <- -subject$beta
  subject
})
fit_flipped <- dkge(subjects_flipped, K = K, rank = 2, weights = wts_mixed)
max(abs(fit_flipped$voxel_weights - fit_mixed$voxel_weights))
```

```{r sign-invariance-checks, include=FALSE}
stopifnot(max(abs(
  fit_flipped$voxel_weights - fit_mixed$voxel_weights
)) < 1e-12)
```

This equality is a necessary check for the implemented sign-flip route, not a
claim of exact finite-sample calibration. A new weighting rule still needs a
simulation study matched to its folds, permutation scheme, sample size, and
null data-generating process, with Monte Carlo uncertainty reported.

The public API currently accepts only the built-in adaptive rules shown above;
it does not accept a custom weighting function. Sign-flip inference operates on
already computed contrast values and holds the fitted voxel weights fixed. There
is no `perm_recompute` argument or permutation-time weight-recomputation path.
Consequently, an odd, sign-sensitive custom rule is not supported for this
inference route. `dkge_weights()` rejects custom rules and unknown arguments, so
the public API fails closed. Supporting such a rule would require a new,
end-to-end resampling path that refits and validates the weights for every flip.

For the supported rules, keep `scope = "fold"` so the held-out subject does not
contaminate the training statistics used to construct adaptive weights.

### Updating an existing fit

You can fit once and revisit the weighting later without rerunning
preprocessing. `dkge_update_weights()` rebuilds the fold bases
in place with your new weighting scheme.

```{r update}
fit_refit <- dkge_update_weights(fit_uniform, wts_mixed)
```

The update touches only the weighting machinery; subject-level projections and
stored results are reused.
projections remain fully compatible with the updated model.

## Summary

Adaptive weighting in DKGE has four building blocks: a prior, an adaptive
statistic, a combination rule, and shrinkage parameters. Training-fold scope
and sign-invariance close two important leakage paths. They do not prove that a
particular analysis has calibrated error rates.

The examples above showed how to inspect the resulting weights, mix prior
knowledge with adaptive rules, and retrofit a new weighting choice onto an
existing fit. Use these
weighting controls when external reliability information or a prespecified
adaptive rule justifies unequal voxel influence. Compare the weighted fit with
an equal-weight baseline, report the rule and shrinkage settings, and calibrate
the full inferential procedure under a defensible null before making
sensitivity or error-control claims.


## Where to go next

- `vignette("dkge-weighting")` — the three weighting
  layers this page sits inside: spatial, subject-level, and transport.
- `vignette("dkge-contrasts-inference")` — the inference
  machinery that a weighting choice has to survive.
- `vignette("dkge-concepts")` — what changing a weight does to the
  estimand.
