---
title: Getting Started with DKGE
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Getting Started with DKGE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
set.seed(42)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

```{r load-dkge}
library(dkge)
```

You have one GLM beta matrix per subject. Its rows are the same named design
effects, but its columns may be different subject-specific clusters. You want
to find effect-space directions that recur across subjects, inspect what those
directions express, and evaluate a planned contrast without constructing a
voxel-by-voxel group covariance matrix.

DKGE takes those beta matrices, their design matrices, and a design kernel. It
returns a low-dimensional group basis in effect space plus subject-level
cluster values that can be cross-fitted and, when coordinates are available,
transported to a common parcellation.

## Can you fit a complete example?

The package includes a simulator with known factorial structure. Here the
planted signal involves two main effects in a 2 by 3 design. Each subject has a
5 by 20 beta matrix: five effect rows and twenty clusters.

```{r simulate-example}
toy <- dkge_sim_toy(
  factors = list(condition = list(L = 2), load = list(L = 3)),
  active_terms = c("condition", "load"),
  S = 5, P = 20, snr = 5, seed = 42
)
dim(toy$B_list[[1]])
```

The three objects needed for a first fit are `toy$B_list` (subject beta
matrices), `toy$X_list` (matching design rulers), and `toy$K` (the effect-space
kernel). Fit two components:

```{r fit-example}
fit <- dkge(toy$B_list, toy$X_list, K = toy$K, rank = 2)
dkge_plot_scree(fit)
```

The bars show each retained component's share of the fitted q-space variation;
the line is cumulative. This is a descriptive summary of the fitted latent
space, not a significance test.

```{r fit-example-checks, include=FALSE}
variance_table <- dkge_variance_explained(fit)
stopifnot(nrow(fit$U) == nrow(toy$K), ncol(fit$U) == 2L)
stopifnot(all(is.finite(variance_table$prop_var)))
stopifnot(all(variance_table$prop_var >= 0))
stopifnot(sum(variance_table$prop_var) > 0)
```

## What does a component express?

Inspect the design-weighted saliences, \(K U\), before mapping anything back to
space. Rows are effects and columns are components.

```{r inspect-saliences}
round(dkge_component_saliences(fit, comps = 1:2, long = FALSE), 2)
```

Large positive and negative entries identify the effect directions associated
with each component. Component signs are arbitrary, so interpret relative
patterns rather than treating the sign itself as scientifically meaningful.

**Next operation:** use `dkge_plot_effect_loadings(fit)` for a heatmap, or move
to a prespecified effect contrast.

## How do you evaluate a contrast without reusing each subject's data?

Construct a contrast in the same effect order as the fit. The simulator records
which effect coordinate carries the planted `condition` term.

```{r loso-contrast}
condition_contrast <- numeric(nrow(toy$K))
condition_contrast[toy$active_cols$condition] <- 1

condition_loso <- dkge_contrast(
  fit, list(condition = condition_contrast), method = "loso"
)
condition_loso
```

`method = "loso"` refits the latent basis without each held-out subject before
computing that subject's cluster values. The result contains one vector per
subject under `condition_loso$values$condition`; those vectors still live on
the subjects' own cluster systems.

```{r loso-contrast-checks, include=FALSE}
condition_values <- condition_loso$values$condition
stopifnot(length(condition_values) == length(toy$B_list))
stopifnot(all(vapply(condition_values, function(x) all(is.finite(x)), logical(1))))
stopifnot(any(vapply(condition_values, function(x) stats::sd(x) > 0, logical(1))))
```

Cross-fitting limits basis-reuse optimism; it does not by itself provide a
population p-value, spatial multiplicity correction, or causal interpretation.
Those require an inference procedure matched to the claim and an experimental
design that supports the interpretation.

**Next operation:** if clusters differ across subjects, transport the contrast
with `dkge_transport_contrasts_to_medoid()`. If they already share a common
index, `as.matrix(condition_loso)` stacks them directly.

## What must your real data look like?

For subject \(s\), supply:

- a numeric `q` by `P_s` beta matrix, where rows are named effects and columns
  are clusters, parcels, or voxels;
- a numeric `T_s` by `q` design matrix whose column names match the beta-row
  names; and
- optionally, cluster weights, coordinates, effect counts, or trial-level
  sufficient statistics needed by later stages.

Wrap subjects with `dkge_subject()` when you want validation and provenance,
then combine them with `dkge_data()`. Subjects may have different `P_s`. If
some effects are unobserved, declare that explicitly; DKGE records coverage and
does not treat an absent effect as an observed zero.

```{r bundle-example}
subjects <- Map(
  function(beta, design, id) dkge_subject(beta, design = design, id = id),
  toy$B_list, toy$X_list, toy$subject_ids
)
data_bundle <- dkge_data(subjects)
c(subjects = length(data_bundle$subject_ids), effects = data_bundle$q)
```

For unequal trial counts or missing design cells, continue with
`vignette("dkge-partial-effect-spaces")` before fitting.

## Which page should you read next?

The default path is:

1. `vignette("dkge-workflow")` — take the same objects through fit, comparison,
   cross-fitted contrast, and spatial transport.
2. `vignette("dkge-concepts")` — understand the kernel metric, the fitted
   moment, and the distinct levels of inference.
3. `vignette("dkge-design-kernels")` — construct a kernel for your own design.
4. `vignette("dkge-contrasts-inference")` — choose uncertainty and testing
   machinery for a prespecified contrast.

Common branches are `vignette("dkge-partial-effect-spaces")` for incomplete or
unequally estimated effects, `vignette("dkge-weighting")` for weighting
estimands, and `vignette("dkge-plotting")` for presentation.

For function-level help, start with `?dkge`, `?dkge_subject`,
`?design_kernel`, and `?dkge_contrast`.
