---
title: "Unbalanced Trialwise and Partial Effect Spaces"
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Unbalanced Trialwise and Partial Effect Spaces}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(dkge)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 3)
set.seed(73101)
```

DKGE can represent a common factorial effect space even when subjects have
unequal trial counts or locally empty cells. The key distinction is between the
*global* 60-cell coordinate system and each subject's observed subset. Missing
cells are declared structurally; they are not supplied as ordinary zero beta
rows.

## Declare the complete design and every intended term

This 3 by 5 design has an ordinal response with four levels. The ordinal kernel
uses distances between level indices, so adjacent responses are more similar
than responses 1 and 4.

```{r grid}
grid <- dkge_effect_grid(list(
  A = list(L = 3, type = "nominal", levels = paste0("A", 1:3)),
  B = list(L = 5, type = "nominal", levels = paste0("B", 1:5)),
  response = list(L = 4, type = "ordinal", levels = as.character(1:4), l = 1)
))

terms <- list(
  "A", "B", "response",
  c("A", "B"), c("A", "response"), c("B", "response"),
  c("A", "B", "response")
)
kernel <- design_kernel(grid, terms = terms, basis = "cell",
                        normalize = "max_diag")
c(cells = nrow(grid$cells), kernel_rows = nrow(kernel$K))
```

Spell out `terms`: with three factors the default is the three main effects
plus the full three-way interaction, not all two-way interactions. The grid's
`scope` field labels terms for downstream contrast routing; changing it does
not change `kernel$K`.

## Combine subjects with local empty cells

Here each subject observes 50 of the 60 global cells. Counts differ strongly
within subject, and different subjects omit different cells. A local subject
contains only its observed rows; `dkge_data()` constructs and records the
global union.

```{r partial-subjects}
q <- nrow(grid$cells)
P <- 12
response_index <- as.numeric(grid$cells$response)
effect_signal <- as.numeric(scale(response_index +
  0.3 * as.integer(factor(grid$cells$A, levels = paste0("A", 1:3))),
  scale = FALSE))
spatial_signal <- cos(seq(0, 2 * pi, length.out = P))

subjects <- lapply(1:6, function(s) {
  observed <- setdiff(seq_len(q), seq.int(s, q, by = 6))
  counts <- 2 + ((observed * (s + 2)) %% 18)
  beta <- tcrossprod(effect_signal[observed], spatial_signal) +
    matrix(rnorm(length(observed) * P), length(observed), P) /
      sqrt(counts)
  rownames(beta) <- grid$cell_labels[observed]
  design <- diag(length(observed))
  colnames(design) <- rownames(beta)
  covariance <- diag(1 / counts, length(observed))
  dimnames(covariance) <- list(rownames(beta), rownames(beta))
  suppressWarnings(dkge_subject(
    beta, design, id = paste0("s", s),
    effect_n = setNames(counts, rownames(beta)),
    effect_noise_cov = covariance,
    residual_variance = rep(1, P), residual_df = 80
  ))
})
bundle <- dkge_data(subjects)
range(diag(bundle$provenance$pair_counts))
```

Count weighting treats precision as proportional to the number of trials.
Random-effects weighting combines within-subject uncertainty with a capped
between-subject variance estimate. Both are explicit choices. Analytic
correction addresses the finite-trial diagonal bias; weighting alone does not.

```{r fits}
fit_count <- dkge_fit(
  bundle, K = kernel, rank = 3, w_method = "none",
  effect_weights = dkge_effect_weights("count"),
  debias = "analytic", missingness = "rescale"
)
fit_random <- dkge_fit(
  bundle, K = kernel, rank = 3, w_method = "none",
  effect_weights = dkge_effect_weights("random_effects", within = "noise"),
  debias = "analytic", missingness = "rescale"
)
summary(fit_random)
```

The summary makes the estimator auditable: coverage, pair effective sample
size, negative spectral mass, error-model provenance, and the q-space versus
physical-block representation are visible before inference.

## Supply trial covariance when trialwise betas are correlated

`dkge_trial_subject()` fits `Y = X B + E` in trial space. Its IID default is
appropriate only when trial-estimate errors are independent with a common
scale model. Overlapping first-level estimates or temporal dependence require
a covariance or precision matrix.

```{r covariance-input}
trial_cell <- rep(1:6, each = 4)
X <- model.matrix(~ 0 + factor(trial_cell, levels = 1:6))
colnames(X) <- paste0("cell", 1:6)
truth <- matrix(rnorm(6 * 10), 6, 10)
Sigma <- outer(seq_len(nrow(X)), seq_len(nrow(X)),
               function(i, j) 0.35^abs(i - j))
Y <- X %*% truth + t(chol(Sigma)) %*%
  matrix(rnorm(nrow(X) * 10), nrow(X), 10)
trial_subject <- dkge_trial_subject(Y, X, trial_covariance = Sigma)
trial_subject$error_model[c("estimator", "trial_dependence")]
```

The constructor stores the `T x q` design and q-space sufficient statistics,
but not `Y`. At imaging scale, stream feature blocks so the full `T x P`
matrix is never materialized:

```{r chunks}
chunked <- dkge_trial_subject_chunks(
  list(Y[, 1:4, drop = FALSE], Y[, 5:10, drop = FALSE]),
  X, trial_covariance = Sigma
)
chunked$error_model[c("storage", "n_chunks", "effect_score_retained")]
```

A saturated 60-cell trialwise design must observe every fitted local column to
be full rank. For a subject with empty cells, either fit only its observed cell
columns and let `dkge_data()` align the named subset, or use a lower-dimensional
second-stage basis whose coefficients remain estimable. Do not insert empty
cells as observed zero rows.

## Split moments require an independence argument

Within-cell alternation balances counts but does not prove that halves are
independent. Run-disjoint splitting establishes a stronger provenance contract;
otherwise DKGE labels the cross-moment exploratory. Setting
`split_independent = TRUE` is an assertion by the analyst, not a test performed
by the package.

The tracked 20-seed simulation under `inst/extdata` is deliberately mixed: the
covariance-aware analytic and split estimators reduce inverse-count artifacts,
but reliability weighting alone does not pass that bias gate, and analytic
interval coverage remains below its predeclared target. These failures are part
of the evidence, not hidden by the API.
